Aggregate HUMAnN outputs by `stool.ID`. I will average relabs and TPM values. I will exclude the control samples.

# Load libraries

```{r, warning=F, message=F}
library(tidyverse)
```

# Import metadata

```{r}
seq.meta.df <- read.table("../data/metadata/sequencing_metadata_after_QC_n3473_230620.txt", sep="\t", header=T)
lib.meta.df <- read.table("../data/metadata/library_metadata_after_QC_n3213_230620.txt", sep="\t", header=T)
stool.meta.df <- read.table("../data/metadata/stool_metadata_after_QC_n3124_230620.txt", sep="\t", header=T)
mouse.meta.df <- read.csv("../data/metadata/AnimalData_Processed_20230712.csv")
```

```{r}
seq.meta.annot.df <- seq.meta.df %>%
  merge(lib.meta.df, by="lib.ID") %>%
  merge(stool.meta.df, by="stool.ID") %>% 
  merge(mouse.meta.df, by.x="mouse.ID", by.y="MouseID", all.x=T)

dim(seq.meta.df)
dim(seq.meta.annot.df)
```

Exclude controls.

```{r}
seq.meta.annot.df.no.ctls <- seq.meta.annot.df %>% 
  dplyr::filter(sample.type == "DO")

dim(seq.meta.annot.df)
dim(seq.meta.annot.df.no.ctls)
```

```{r}
n_distinct(seq.meta.annot.df$stool.ID)
n_distinct(seq.meta.annot.df.no.ctls$stool.ID)
```

* 3124 unique stool IDs, 2997 unique DO stool IDs

# Import MetaPhlan data

```{r}
mpa.df <- read.table("../data/metaphlan.txt", sep="\t", header=T)
dim(mpa.df)
```

Subset to genus-level relative abundances.

```{r}
genus.df <- mpa.df %>% 
  
  # subset to genera (this step excludes UNCLASSIFIED)
  dplyr::filter(str_detect(clade_name, "g__")) %>% 
  dplyr::filter(!str_detect(clade_name, "s__")) %>% 
  
  # create "genus" column and set as rownames
  mutate(genus = str_extract(clade_name, "g__(.*)")) %>% 
  mutate(genus = str_sub(genus, 4)) %>% # remove g__ prefix
  dplyr::select(-clade_name) %>%
  column_to_rownames("genus")
```

```{r}
summary(rowSums(genus.df != 0))
```

* No genera are all 0s

```{r}
sum(rowSums(genus.df != 0) == 1)
```

* But 25 genera have just one non-zero value

# Import pathway data

```{r}
path.rpk.df <- read.table(
  "../data/pathabundance_unstratified.txt", 
  sep="\t", header=T, quote="", row.names=1)
dim(path.rpk.df)
```

* 529 pathways x 3473 samples

# % unclassified

Exclude controls.

```{r}
mpa.df %>% 
  dplyr::select(clade_name, all_of(
    seq.meta.annot.df.no.ctls %>% dplyr::filter(sample.type == "DO") %>% pull(seq.ID))) %>% 
  dplyr::filter(clade_name == "UNCLASSIFIED") %>% 
  dplyr::select(-clade_name) %>% 
  as.numeric() %>% summary()
```

* DO samples: median 12.3%, mean 11.9%, min 0%, max 32.0%

# % UNMAPPED & UNINTEGRATED

```{r}
(path.rpk.df["UNMAPPED", ] / colSums(path.rpk.df)) %>% 
  t() %>% data.frame() %>% 
  setNames(c("frac.unmapped")) %>% 
  rownames_to_column("seq.ID") %>% 
  merge(seq.meta.annot.df.no.ctls, by="seq.ID") %>% 
  pull(frac.unmapped) %>% 
  summary()
```

* DO samples: mean 54.6% UNMAPPED

```{r}
(path.rpk.df["UNINTEGRATED", ] / colSums(path.rpk.df)) %>% 
  t() %>% data.frame() %>% 
  setNames(c("frac.unmapped")) %>% 
  rownames_to_column("seq.ID") %>% 
  merge(seq.meta.annot.df.no.ctls, by="seq.ID") %>% 
  pull(frac.unmapped) %>% 
  summary()
```

* DO samples: mean 42.8% UNINTEGRATED

# Aggregate

## Genera

Recompute relabs so all columns add to 100.

```{r}
genus.df.cols.add.to.100 <- 100 * sweep(genus.df, MARGIN=2, STATS=colSums(genus.df), FUN="/")
```

```{r}
agg.genus.long.df <- genus.df.cols.add.to.100 %>% 
  rownames_to_column("genus") %>% 
  pivot_longer(-genus, names_to="seq.ID", values_to="relab") %>% 
  merge(seq.meta.annot.df.no.ctls, by="seq.ID") %>% 
  group_by(stool.ID, genus) %>% 
  summarise(relab = mean(relab), .groups="drop")
```

```{r}
agg.genus.df <- agg.genus.long.df %>% 
  pivot_wider(id_cols=genus, names_from=stool.ID, values_from=relab) %>% 
  column_to_rownames("genus")
dim(agg.genus.df)
```

* 383 genera x 2997 stool IDs

```{r}
sum(rowSums(agg.genus.df != 0) == 0)
```

Remove 7 genera that are all zeros (after removing controls).

```{r}
agg.genus.filt.df <- agg.genus.df[rowSums(agg.genus.df != 0) > 0, ]

dim(agg.genus.df)
dim(agg.genus.filt.df)
```

* 383 --> 376 genera

### Sanity-check: do samples with the same `stool.ID` look similar?

```{r}
# seq.meta.annot.df.no.ctls %>% 
#   dplyr::count(stool.ID) %>% 
#   arrange(desc(n))
# 
# genus.df.cols.add.to.100[, (
#   seq.meta.annot.df.no.ctls %>% 
#     dplyr::filter(stool.ID == "DO_2D_4047_069w") %>% 
#     pull(seq.ID))]
```

## Pathways

We have RPK values, but we'll need TPM to aggregate across samples. 

**Remove UNMAPPED and UNCLASSIFIED before computing TPM.**

```{r}
path.tpm.df <- path.rpk.df %>%
  rownames_to_column("pathway") %>% 
  dplyr::filter(!(pathway %in% c("UNMAPPED", "UNINTEGRATED"))) %>% 
  column_to_rownames("pathway")
```

```{r}
path.tpm.df <- 1e6 * sweep(path.tpm.df, MARGIN=2, STATS=colSums(path.tpm.df), FUN="/")
```

```{r}
summary(colSums(path.tpm.df))
```

The NAs are for the negative controls that have zero total pathway TPM abundances.

```{r}
agg.path.tpm.long.df <- path.tpm.df %>% 
  rownames_to_column("Pathway") %>% 
  pivot_longer(-Pathway, names_to="seq.ID", values_to="tpm") %>% 
  merge(seq.meta.annot.df.no.ctls, by="seq.ID") %>% # this merge removes controls
  group_by(stool.ID, Pathway) %>% 
  summarise(tpm = mean(tpm), .groups="drop")
```

```{r}
agg.path.tpm.df <- agg.path.tpm.long.df %>% 
  pivot_wider(id_cols=Pathway, names_from=stool.ID, values_from=tpm) %>% 
  column_to_rownames("Pathway")
dim(agg.path.tpm.df)
```

```{r}
sum(rowSums(agg.path.tpm.df != 0) == 0)
```

Remove 105 pathways that are all zeros (after removing controls).

```{r}
agg.path.tpm.filt.df <- agg.path.tpm.df[rowSums(agg.path.tpm.df != 0) > 0, ]

dim(agg.path.tpm.df)
dim(agg.path.tpm.filt.df)
```

* 527 --> 422 pathways

```{r}
summary(colSums(agg.path.tpm.filt.df))
```

* All samples add to 1 M

# Export

```{r}
agg.genus.filt.df %>%
  rownames_to_column("genus") %>% 
  write.table(
    sprintf("../data/metaphlan_genus_agg_by_stool_ID_n%ix%i.txt",
            nrow(.), ncol(.)-1),
    sep="\t", quote=F, row.names=F)
```

```{r}
agg.path.tpm.filt.df %>%
  rownames_to_column("Pathway") %>% 
  write.table(
    sprintf("../data/pathabundance_tpm_agg_by_stool_ID_n%ix%i.txt",
            nrow(.), ncol(.)-1),
    sep="\t", quote=F, row.names=F)
```
