Create cross-sectional slices of the data at specific points in time.

We will create 3 slices:
* Slice 1 (2018 Q1) = 573 samples collected from November 20, 2017 to March 5, 2018
* Slice 2 (2018 Q3) = 424 samples collected from April 30, 2018 to August 13, 2018
* Slice 3 (2019 Q1) = 381 samples collected from October 22, 2018 to February 11, 2019

```{r, warning=F, message=F}
library(tidyverse)
library(lubridate)

cohort.color.palette <- c(
  G22W1 = "royalblue",
  G22W2 = "royalblue4",
  G23W1 = "peachpuff",
  G23W2 = "peachpuff3",
  G24W1 = "purple",
  G24W2 = "purple4",
  G26W1 = "maroon",
  G26W2 = "maroon4",
  G27W1 = "seagreen",
  G27W2 = "seagreen3",
  G28W1 = "darkorange",
  G28W2 = "darkorange4")
```

# Import data

```{r}
genus.data.df <- read.table(
  "../results/kraken_genus_clr_filt_w_comm_n107x2997_231017.txt",
  sep="\t", header=T, row.names=1)

pathway.data.df <- read.table(
  "../results/DO_pathway_log2tpm_filt_w_comm_n273x2997.txt",
  sep="\t", header=T, row.names=1)
```

# Import metadata

```{r}
stool.meta.df <- read.table(
  "../data/metadata/stool_metadata_after_QC_no_controls_n2997_240418.txt", 
  sep="\t", header=T)
mouse.meta.df <- read.csv(
  "../data/metadata/AnimalData_Processed_20230712.csv") %>% 
  mutate(Cage = paste0("c", HID))

stool.meta.annot.df <- merge(
  stool.meta.df, mouse.meta.df, 
  by.x="mouse.ID", by.y="MouseID")

stool.meta.annot.df <- stool.meta.annot.df %>% 
  
  # pre-randomization timepoints should be considered AL
  mutate(Diet.5mo.as.AL = case_when(
    age.approx.months == 5 ~ "AL",
    TRUE ~ as.character(Diet))) %>% 
  mutate(Diet.5mo.as.AL = factor(Diet.5mo.as.AL, levels=c(
      "AL", "1D", "2D", "20", "40"))) %>% 
  
  # convert to date type
  mutate(collection.date = ymd(date.stool.collection.approx)) %>%

  # create new column with days since first stool collection  
  mutate(days.from.first.stool.collection = time_length(
    collection.date - min(collection.date), unit="day")) %>% 
  
  # create new column for quarter
  mutate(quarter.stool.collection = round_date(collection.date, "quarter"))
```

# Take cross-sectional slices

```{r}
start.2018.Q1 <- ymd("2017-11-20")
end.2018.Q1 <- ymd("2018-03-05")

start.2018.Q3 <- ymd("2018-04-30")
end.2018.Q3 <- ymd("2018-08-13")

start.2019.Q1 <- ymd("2018-10-22")
end.2019.Q1 <- ymd("2019-02-11")
```

Rename the slices based on the number of days from the start of the experiment, rather than date.

```{r}
start.of.experiment <- min(stool.meta.annot.df$collection.date)

start.2018.Q1.in.days <- time_length(
  start.2018.Q1 - start.of.experiment, unit="day")
end.2018.Q1.in.days <- time_length(
  end.2018.Q1 - start.of.experiment, unit="day")

start.2018.Q3.in.days <- time_length(
  start.2018.Q3 - start.of.experiment, unit="day")
end.2018.Q3.in.days <- time_length(
  end.2018.Q3 - start.of.experiment, unit="day")

start.2019.Q1.in.days <- time_length(
  start.2019.Q1 - start.of.experiment, unit="day")
end.2019.Q1.in.days <- time_length(
  end.2019.Q1 - start.of.experiment, unit="day")
```

```{r}
stool.meta.annot.df.2018.Q1 <- stool.meta.annot.df %>% 
  dplyr::filter(
    (date.stool.collection.approx >= start.2018.Q1) & (
      date.stool.collection.approx <= end.2018.Q1))

stool.meta.annot.df.2018.Q3 <- stool.meta.annot.df %>% 
  dplyr::filter(
    (date.stool.collection.approx >= start.2018.Q3) & (
      date.stool.collection.approx <= end.2018.Q3))

stool.meta.annot.df.2019.Q1 <- stool.meta.annot.df %>% 
  dplyr::filter(
    (date.stool.collection.approx >= start.2019.Q1) & (
      date.stool.collection.approx <= end.2019.Q1))
```

```{r}
stool.meta.annot.df.2018.Q1 %>% nrow()
stool.meta.annot.df.2018.Q3 %>% nrow()
stool.meta.annot.df.2019.Q1 %>% nrow()
```

* 573, 424, and 381 samples per slice

```{r}
genus.data.df.2018.Q1 <- genus.data.df[, stool.meta.annot.df.2018.Q1$stool.ID]
genus.data.df.2018.Q3 <- genus.data.df[, stool.meta.annot.df.2018.Q3$stool.ID]
genus.data.df.2019.Q1 <- genus.data.df[, stool.meta.annot.df.2019.Q1$stool.ID]

pathway.data.df.2018.Q1 <- pathway.data.df[, stool.meta.annot.df.2018.Q1$stool.ID]
pathway.data.df.2018.Q3 <- pathway.data.df[, stool.meta.annot.df.2018.Q3$stool.ID]
pathway.data.df.2019.Q1 <- pathway.data.df[, stool.meta.annot.df.2019.Q1$stool.ID]
```

# Plot cross-sectional slices

Plot the number of samples versus age and time, colored by cohort.

```{r}
(day.of.experiment.v.age.dots.p <- stool.meta.annot.df %>% 
   dplyr::count(age.approx.months, Cohort, days.from.first.stool.collection) %>% 
   ggplot(aes(x=days.from.first.stool.collection,
              y=fct_rev(factor(age.approx.months)),
              size=n, color=Cohort)) +
   annotate("rect",
     xmin=c(start.2018.Q1.in.days, start.2018.Q3.in.days, start.2019.Q1.in.days),
     xmax=c(end.2018.Q1.in.days, end.2018.Q3.in.days, end.2019.Q1.in.days),
     ymin=-Inf, ymax=Inf,
     alpha=0.1) +
   geom_point() +
   theme_classic(base_size=8) +
   scale_color_manual(values=cohort.color.palette) +
   scale_size(range=c(0.5, 3)) +
   labs(x="Day of Experiment", y="Age (months)",
        size="# stool samples"))
```

```{r}
# pdf("plots/day_of_experiment_v_age_dots_w_slices.pdf", height=1.4, width=2.2)
# day.of.experiment.v.age.dots.p + theme(legend.position="none")
# dev.off()
```

# Export data slices

```{r}
# genus.data.df.2018.Q1 %>%
#   rownames_to_column("feature") %>%
#   write.table(sprintf(
#     "../results/cross_sectional_slices/kraken_genus_clr_filt_w_comm_slice1_n107x%i.txt",
#     ncol(genus.data.df.2018.Q1)), sep="\t", quote=F, row.names=F)
# 
# genus.data.df.2018.Q3 %>% 
#   rownames_to_column("feature") %>% 
#   write.table(sprintf(
#     "../results/cross_sectional_slices/kraken_genus_clr_filt_w_comm_slice2_n107x%i.txt",
#     ncol(genus.data.df.2018.Q3)), sep="\t", quote=F, row.names=F)
# 
# genus.data.df.2019.Q1 %>% 
#   rownames_to_column("feature") %>% 
#   write.table(sprintf(
#     "../results/cross_sectional_slices/kraken_genus_clr_filt_w_comm_slice3_n107x%i.txt",
#     ncol(genus.data.df.2019.Q1)), sep="\t", quote=F, row.names=F)
```

```{r}
# pathway.data.df.2018.Q1 %>% 
#   rownames_to_column("feature") %>% 
#   write.table(sprintf(
#     "../results/cross_sectional_slices/pathway_log2tpm_filt_w_comm_slice1_n273x%i.txt",
#     ncol(pathway.data.df.2018.Q1)), sep="\t", quote=F, row.names=F)
# 
# pathway.data.df.2018.Q3 %>% 
#   rownames_to_column("feature") %>% 
#   write.table(sprintf(
#     "../results/cross_sectional_slices/pathway_log2tpm_filt_w_comm_slice2_n273x%i.txt",
#     ncol(pathway.data.df.2018.Q3)), sep="\t", quote=F, row.names=F)
# 
# pathway.data.df.2019.Q1 %>% 
#   rownames_to_column("feature") %>% 
#   write.table(sprintf(
#     "../results/cross_sectional_slices/pathway_log2tpm_filt_w_comm_slice3_n273x%i.txt",
#     ncol(pathway.data.df.2019.Q1)), sep="\t", quote=F, row.names=F)
```
