Create supplementary tables.

Supplementary Table 3 = genera, Kraken, Model 1
Supplementary Table 4 = species, Kraken, Model 1
Supplementary Table 5 = genera, MetaPhlAn, Model 1
Supplementary Table 6 = pathways, Model 1
Supplementary Table 7 = DO AL, genera, MetaPhlAn, MaAsLin2, Model 5
Supplementary Table 8 = B6, genera, MetaPhlAn, MaAsLin2, Model 6
Supplementary Table 9 = human, genera, MetaPhlAn, MaAsLin2, Model 7
Supplementary Table 10 = DO AL, pathways, MaAsLin2, Model 5
Supplementary Table 11 = B6, pathways, MaAsLin2, Model 6
Supplementary Table 12 = human, pathways, MaAsLin2, Model 7
Supplementary Table 13 = genera, Kraken, Model 10 (all random effects)
Supplementary Table 14 = pathways, Model 10 (all random effects)
Supplementary Table 15 = associations between genera and phenotypes, Model 11 (longitudinal)
Supplementary Table 16 = associations between pathways and phenotypes, Model 11 (longitudinal)
Supplementary Table 17 = mediation results between genera and phenotypes, Models 13 and 14 (longitudinal)
Supplementary Table 18 = mediation results between pathways and phenotypes, Models 13 and 14 (longitudinal)
Supplementary Table 19 = associations between genera and phenotypes, Model 12 (cross-sectional)
Supplementary Table 20 = associations between pathways and phenotypes, Model 12 (cross-sectional)

**Tables 15-20 are exported in `pheno_assoc_and_mediation.Rmd`.**

For fixed effects, we report coefficients, p-values (Wald test), and adjusted p-values (BH). For random effects, we report proportion of variance explained, p-values (LRT), and adjusted p-values (BH).

```{r, warning=F, message=F}
library(tidyverse)
```

```{r}
community.features <- c(
  "PCoA1", "PCoA2", "PCoA3", "Shannon", "Simpson", "Chao1",
  
  # used in MaAsLin
  "Shannon", "Simpson", "InvSimpson", "Observed",
  
  # inconsistently named
  "uniq", "Uniqueness")
```

# Import metadata

## Short and long pathway names

```{r}
pathway.names.df <- read.table(
  "../data/metadata/pathway_long_and_short_names_n703.txt",
  sep="\t", header=T, quote="")
```

```{r}
n_distinct(pathway.names.df$original)
n_distinct(pathway.names.df$short.clean)
```

`original` is unique, but `short.clean` is not. Let's make the mapping one-to-one. Not sure which original pathway name is best to keep, so will choose randomly.

```{r}
pathway.names.uniq.df <- pathway.names.df %>% 
  distinct(short.clean, .keep_all=T)
```

# Define functions

## Fixed effects

1. Age: rename columns, calculate adjusted p-values.
2. DR: pivot wider to end up with one feature per row, rename columns, calculate adjusted p-values.
3. Merge the age and DR dataframes
4. Indicate which features are community features

```{r}
process_fixef_df <- function(fixef.df) {
  
  # age
  age.df <- fixef.df %>% 
    dplyr::filter(fixef == "age.wks.scaled") %>% 
    dplyr::select(feature, solution, std.error, Pr.chisq..con.) %>% 
    dplyr::rename(solution.age = solution,
                  std.error.age = std.error,
                  pval.age = Pr.chisq..con.) %>% 
    mutate(padj.age = p.adjust(pval.age, method="BH"))
  
  # DR
  DR.df <- fixef.df %>% 
    dplyr::filter(fixef == "Diet.5mo.as.AL") %>% 
    dplyr::filter(fixef.value != "AL") %>% 
    pivot_wider(id_cols = c(feature, Pr.chisq..con.), 
                values_from = c(solution, std.error),
                names_from="fixef.value", names_sep=".") %>% 
    dplyr::rename(pval.DR = Pr.chisq..con.) %>%
    relocate(pval.DR, .after = last_col()) %>% 
    mutate(padj.DR = p.adjust(pval.DR, method="BH"))
  
  # combine
  out.df <- merge(
    age.df,
    DR.df,
    by="feature"
  )
  
  return(out.df)
  
}
```

## Random effects

1. Rename variables to have cleaner names
2. Export PVE for each random effect
3. For genetics, we have standard error and a p-value, so we'll add those

```{r}
process_ranef_df <- function(ranef.df, herit.df) {
  
  PVE.df <- ranef.df %>% 
    
    # ignore residual
    dplyr::filter(term != "units!R") %>% 
    
    # create cleaner names for certain experimental variables
    mutate(term.clean = case_match(
      term,
      "vm(Mouse, kinship.mat.x2)" ~ "Genetics",
      "ide(Mouse)" ~ "Mouse",
      .default = term)) %>% 
    
    # pivot
    pivot_wider(id_cols = feature,
                names_from = term.clean,
                values_from = PVE)
  
  # prepend column names with "PVE." (except for the first column "feature")
  colnames(PVE.df)[2:ncol(PVE.df)] <- paste0("PVE.", colnames(PVE.df)[2:ncol(PVE.df)])
  
  # create adjusted p-values for heritability
  herit.df <- herit.df %>% 
    dplyr::rename(herit.pval = LRT.pval) %>% 
    mutate(herit.padj = p.adjust(herit.pval, method="BH"))
  
  # merge
  out.df <- merge(herit.df, PVE.df, by="feature") %>% 
    
    # "PVE.Genetics" and "herit" columns are redundant
    dplyr::select(-PVE.Genetics)
  
  return(out.df)
  
}
```

# Table 3: Kraken, genera, Model 1

```{r}
kraken.genus.model1.fixef.df <- read.table(
  "../results/asreml_kraken_genus_w_time_ranef/fixef.txt",
  sep="\t", header=T)

kraken.genus.model1.ranef.df <- read.table(
  "../results/asreml_kraken_genus_w_time_ranef/ranef.txt",
  sep="\t", header=T)

kraken.genus.model1.herit.df <- read.table(
  "../results/asreml_kraken_genus_w_time_ranef/herit.txt",
  sep="\t", header=T)
```

```{r}
kraken.genus.model1.fixef.out.df <- process_fixef_df(kraken.genus.model1.fixef.df)
kraken.genus.model1.ranef.out.df <- process_ranef_df(kraken.genus.model1.ranef.df, kraken.genus.model1.herit.df)

tmp.kraken.genus.model1.out.df <- merge(
  kraken.genus.model1.fixef.out.df,
  kraken.genus.model1.ranef.out.df,
  by="feature")

# more adjustments
kraken.genus.model1.out.df <- tmp.kraken.genus.model1.out.df %>% 
  
  # indicate if community feature
  mutate(is.community.feat = case_when(
    feature %in% community.features ~ "Y",
    TRUE ~ "N")) %>% 
  
  # move is.community.feat to second column
  relocate(is.community.feat, .after=feature) %>% 
  
  # round all numbers to 8 decimal places
  mutate_if(is.numeric, ~ round(., digits=8) %>% format()) %>% 
  
  # sort alphabetically
  arrange(feature)

```  

```{r}
kraken.genus.model1.out.df %>%
  write.table("../supp_tables/table3.txt", sep="\t", quote=F, row.names=F)
```

# Table 4: Kraken, species, Model 1

```{r}
kraken.species.model1.fixef.df <- read.table(
  "../results/asreml_kraken_species_w_time_ranef/fixef.txt",
  sep="\t", header=T)

kraken.species.model1.ranef.df <- read.table(
  "../results/asreml_kraken_species_w_time_ranef/ranef.txt",
  sep="\t", header=T)

kraken.species.model1.herit.df <- read.table(
  "../results/asreml_kraken_species_w_time_ranef/herit.txt",
  sep="\t", header=T)
```

```{r}
kraken.species.model1.fixef.out.df <- process_fixef_df(kraken.species.model1.fixef.df)
kraken.species.model1.ranef.out.df <- process_ranef_df(kraken.species.model1.ranef.df, kraken.species.model1.herit.df)

tmp.kraken.species.model1.out.df <- merge(
  kraken.species.model1.fixef.out.df,
  kraken.species.model1.ranef.out.df,
  by="feature")

# more adjustments
kraken.species.model1.out.df <- tmp.kraken.species.model1.out.df %>% 
  
  # indicate if community feature
  mutate(is.community.feat = case_when(
    feature %in% community.features ~ "Y",
    TRUE ~ "N")) %>% 
  
  # move is.community.feat to second column
  relocate(is.community.feat, .after=feature) %>% 
  
  # round all numbers to 8 decimal places
  mutate_if(is.numeric, ~ round(., digits=8) %>% format()) %>% 
  
  # sort alphabetically
  arrange(feature)

```

```{r}
kraken.species.model1.out.df %>%
  write.table("../supp_tables/table4.txt", sep="\t", quote=F, row.names=F)
```

# Table 5: Kraken, genera, Model 1

```{r}
metaphlan.genus.model1.fixef.df <- read.table(
  "../results/asreml_metaphlan_genus_w_time_ranef/fixef.txt",
  sep="\t", header=T)

metaphlan.genus.model1.ranef.df <- read.table(
  "../results/asreml_metaphlan_genus_w_time_ranef/ranef.txt",
  sep="\t", header=T)

metaphlan.genus.model1.herit.df <- read.table(
  "../results/asreml_metaphlan_genus_w_time_ranef/herit.txt",
  sep="\t", header=T)
```

```{r}
metaphlan.genus.model1.fixef.out.df <- process_fixef_df(metaphlan.genus.model1.fixef.df)
metaphlan.genus.model1.ranef.out.df <- process_ranef_df(metaphlan.genus.model1.ranef.df, metaphlan.genus.model1.herit.df)

tmp.metaphlan.genus.model1.out.df <- merge(
  metaphlan.genus.model1.fixef.out.df,
  metaphlan.genus.model1.ranef.out.df,
  by="feature")

# more adjustments
metaphlan.genus.model1.out.df <- tmp.metaphlan.genus.model1.out.df %>% 
  
  # indicate if community feature
  mutate(is.community.feat = case_when(
    feature %in% community.features ~ "Y",
    TRUE ~ "N")) %>% 
  
  # move is.community.feat to second column
  relocate(is.community.feat, .after=feature) %>% 
  
  # round all numbers to 8 decimal places
  mutate_if(is.numeric, ~ round(., digits=8) %>% format()) %>% 
  
  # sort alphabetically
  arrange(feature)

```  

```{r}
metaphlan.genus.model1.out.df %>%
  write.table("../supp_tables/table5.txt", sep="\t", quote=F, row.names=F)
```

# Table 6: pathways, Model 1

Pathways.

```{r}
pathways.model1.fixef.df <- read.table(
  "../results/asreml_humann_pathways_w_time_ranef/fixef.txt",
  sep="\t", header=T)

pathways.model1.ranef.df <- read.table(
  "../results/asreml_humann_pathways_w_time_ranef/ranef.txt",
  sep="\t", header=T)

pathways.model1.herit.df <- read.table(
  "../results/asreml_humann_pathways_w_time_ranef/herit.txt",
  sep="\t", header=T)
```

```{r}
pathways.model1.fixef.out.df <- process_fixef_df(pathways.model1.fixef.df)
pathways.model1.ranef.out.df <- process_ranef_df(pathways.model1.ranef.df, pathways.model1.herit.df)

tmp.pathways.model1.out.df <- merge(
  pathways.model1.fixef.out.df,
  pathways.model1.ranef.out.df,
  by="feature")

# more adjustments
pathways.model1.out.df <- tmp.pathways.model1.out.df %>% 
  
  # indicate if community feature
  mutate(is.community.feat = case_when(
    feature %in% community.features ~ "Y",
    TRUE ~ "N")) %>% 
  
  # move is.community.feat to second column
  relocate(is.community.feat, .after=feature) %>% 
  
  # add long pathway name
  merge(pathway.names.uniq.df %>% dplyr::select(original, short.clean),
        by.x="feature", by.y="short.clean", all.x=T) %>% 
  
  # rename 'original' to 'full.pathway.name' and move to second column
  dplyr::rename(full.pathway.name = original) %>% 
  relocate(full.pathway.name, .after=feature) %>% 
  
  # round all numbers to 8 decimal places
  mutate_if(is.numeric, ~ round(., digits=8) %>% format()) %>% 
  
  # sort alphabetically
  arrange(feature)

```

```{r}
pathways.model1.out.df %>%
  write.table("../supp_tables/table6.txt", sep="\t", quote=F, row.names=F)
```

# MaAsLin2

## Table 7: DO AL, genera, MetaPhlAn, Model 5

```{r}
DO.AL.genus.maaslin.df <- read.table(
  "../results/maaslin2_DO_AL_genus/all_results.tsv",
  sep="\t", header=T)

DO.AL.genus.maaslin.out.df <- DO.AL.genus.maaslin.df %>% 
  
  # indicate if community feature
  mutate(is.community.feat = case_when(
    feature %in% community.features ~ "Y",
    TRUE ~ "N")) %>% 
  
  # rename columns for consistency
  dplyr::rename(solution.age = coef,
                std.error.age = stderr,
                pval.age = pval,
                padj.age = qval) %>% 
  
  # subset to columns of interest
  dplyr::select(feature, is.community.feat, solution.age,	std.error.age,	pval.age,	padj.age) %>% 
  
  # round all numbers to 8 decimal places
  mutate_if(is.numeric, ~ round(., digits=8) %>% format()) %>% 
  
  # sort alphabetically
  arrange(feature)
```

```{r}
DO.AL.genus.maaslin.out.df %>%
  write.table("../supp_tables/table7.txt", sep="\t", quote=F, row.names=F)
```

## Table 8: B6, genera, MetaPhlAn, Model 6

```{r}
B6.genus.maaslin.df <- read.table(
  "../results/maaslin2_B6_genus/all_results.tsv",
  sep="\t", header=T)

B6.genus.maaslin.out.df <- B6.genus.maaslin.df %>% 
  
  # indicate if community feature
  mutate(is.community.feat = case_when(
    feature %in% community.features ~ "Y",
    TRUE ~ "N")) %>% 
  
  # rename columns for consistency
  dplyr::rename(solution.age = coef,
                std.error.age = stderr,
                pval.age = pval,
                padj.age = qval) %>% 
  
  # subset to columns of interest
  dplyr::select(feature, is.community.feat, solution.age,	std.error.age,	pval.age,	padj.age) %>% 
  
  # round all numbers to 8 decimal places
  mutate_if(is.numeric, ~ round(., digits=8) %>% format()) %>% 
  
  # sort alphabetically
  arrange(feature)
```

```{r}
B6.genus.maaslin.out.df %>%
  write.table("../supp_tables/table8.txt", sep="\t", quote=F, row.names=F)
```

## Table 9: Human, MetaPhlAn, Model 7

```{r}
human.genus.maaslin.df <- read.table(
  "../results/maaslin2_human_genus/all_results.tsv",
  sep="\t", header=T)

human.genus.maaslin.out.df <- human.genus.maaslin.df %>% 
  
  # indicate if community feature
  mutate(is.community.feat = case_when(
    feature %in% community.features ~ "Y",
    TRUE ~ "N")) %>% 
  
  # rename columns for consistency
  dplyr::rename(solution.age = coef,
                std.error.age = stderr,
                pval.age = pval,
                padj.age = qval) %>% 
  
  # subset to columns of interest
  dplyr::select(feature, is.community.feat, solution.age,	std.error.age,	pval.age,	padj.age) %>% 
  
  # round all numbers to 8 decimal places
  mutate_if(is.numeric, ~ round(., digits=8) %>% format()) %>% 
  
  # sort alphabetically
  arrange(feature)
```

```{r}
human.genus.maaslin.out.df %>%
  write.table("../supp_tables/table9.txt", sep="\t", quote=F, row.names=F)
```

## Table 10: DO AL, pathways, Model 5

```{r}
DO.AL.pathway.maaslin.df <- read.table(
  "../results/maaslin2_DO_AL_pathway/all_results.tsv",
  sep="\t", header=T)

DO.AL.pathway.maaslin.out.df <- DO.AL.pathway.maaslin.df %>% 
  
  # indicate if community feature
  mutate(is.community.feat = case_when(
    feature %in% community.features ~ "Y",
    TRUE ~ "N")) %>% 
  
  # add long pathway name
  merge(pathway.names.uniq.df %>% dplyr::select(original, short.clean),
        by.x="feature", by.y="short.clean", all.x=T) %>% 
  
  # rename columns for consistency
  dplyr::rename(solution.age = coef,
                std.error.age = stderr,
                pval.age = pval,
                padj.age = qval,
                full.pathway.name = original) %>% 
  
  # subset to columns of interest
  dplyr::select(feature, full.pathway.name, is.community.feat, solution.age,	std.error.age,	pval.age,	padj.age) %>% 
  
  # round all numbers to 8 decimal places
  mutate_if(is.numeric, ~ round(., digits=8) %>% format()) %>% 
  
  # sort alphabetically
  arrange(feature)
```

```{r}
DO.AL.pathway.maaslin.out.df %>%
  write.table("../supp_tables/table10.txt", sep="\t", quote=F, row.names=F)
```

## Table 11: DO AL, pathways, Model 5

```{r}
B6.pathway.maaslin.df <- read.table(
  "../results/maaslin2_B6_pathway/all_results.tsv",
  sep="\t", header=T)

B6.pathway.maaslin.out.df <- B6.pathway.maaslin.df %>% 
  
  # indicate if community feature
  mutate(is.community.feat = case_when(
    feature %in% community.features ~ "Y",
    TRUE ~ "N")) %>% 
  
  # add long pathway name
  merge(pathway.names.uniq.df %>% dplyr::select(original, short.clean),
        by.x="feature", by.y="short.clean", all.x=T) %>% 
  
  # rename columns for consistency
  dplyr::rename(solution.age = coef,
                std.error.age = stderr,
                pval.age = pval,
                padj.age = qval,
                full.pathway.name = original) %>% 
  
  # subset to columns of interest
  dplyr::select(feature, full.pathway.name, is.community.feat, solution.age,	std.error.age,	pval.age,	padj.age) %>% 
  
  # round all numbers to 8 decimal places
  mutate_if(is.numeric, ~ round(., digits=8) %>% format()) %>% 
  
  # sort alphabetically
  arrange(feature)
```

```{r}
B6.pathway.maaslin.out.df %>%
  write.table("../supp_tables/table11.txt", sep="\t", quote=F, row.names=F)
```

## Table 12: Human, pathways, Model 5

```{r}
human.pathway.maaslin.df <- read.table(
  "../results/maaslin2_human_pathway/all_results.tsv",
  sep="\t", header=T)

human.pathway.maaslin.out.df <- human.pathway.maaslin.df %>% 
  
  # indicate if community feature
  mutate(is.community.feat = case_when(
    feature %in% community.features ~ "Y",
    TRUE ~ "N")) %>% 
  
  # add long pathway name
  merge(pathway.names.uniq.df %>% dplyr::select(original, short.clean),
        by.x="feature", by.y="short.clean", all.x=T) %>% 
  
  # rename columns for consistency
  dplyr::rename(solution.age = coef,
                std.error.age = stderr,
                pval.age = pval,
                padj.age = qval,
                full.pathway.name = original) %>% 
  
  # subset to columns of interest
  dplyr::select(feature, full.pathway.name, is.community.feat, solution.age,	std.error.age,	pval.age,	padj.age) %>% 
  
  # round all numbers to 8 decimal places
  mutate_if(is.numeric, ~ round(., digits=8) %>% format()) %>% 
  
  # sort alphabetically
  arrange(feature)
```

```{r}
human.pathway.maaslin.out.df %>%
  write.table("../supp_tables/table12.txt", sep="\t", quote=F, row.names=F)
```

# Table 13: Kraken, genera, Model 10 (all random effects)

```{r}
kraken.genus.model10.ranef.df <- read.table(
  "../results/asreml_kraken_genus_all_ranef_w_time/ranef.txt",
  sep="\t", header=T)


kraken.genus.model10.ranef.out.df <- kraken.genus.model10.ranef.df %>% 

  # rename p-value column
  dplyr::rename(pval = LRT.pval) %>% 
    
  # create adjusted p-values, separately within each experimental variable
  group_by(term.clean) %>% 
  mutate(padj = p.adjust(pval, method="BH")) %>% 
  ungroup() %>% 

  # pivot wider
  pivot_wider(id_cols = feature, names_from = term.clean, 
              values_from = c(PVE, pval, padj), names_sep = ".") %>% 
  
  # indicate if community feature
  mutate(is.community.feat = case_when(
    feature %in% community.features ~ "Y",
    TRUE ~ "N")) %>%
  
  # specify order of columns
  dplyr::select(feature, is.community.feat,
                PVE.Age, pval.Age, padj.Age,
                PVE.Diet, pval.Diet, padj.Diet,
                PVE.Genetics, pval.Genetics, padj.Genetics,
                PVE.Time, pval.Time, padj.Time,
                PVE.Cohort, pval.Cohort, padj.Cohort,
                PVE.Batch, pval.Batch, padj.Batch,
                PVE.Cage, pval.Cage, padj.Cage,
                PVE.Mouse, pval.Mouse, padj.Mouse,
                PVE.Residual, pval.Residual, padj.Residual) %>% 
  
  # round all numbers to 8 decimal places
  mutate_if(is.numeric, ~ round(., digits=8) %>% format()) %>% 
  
  # sort alphabetically
  arrange(feature)
  
```

```{r}
kraken.genus.model10.ranef.out.df %>%
  write.table("../supp_tables/table13.txt", sep="\t", quote=F, row.names=F)
```

# Table 14: pathways, Model 10 (all random effects)

```{r}
pathways.model10.ranef.df <- read.table(
  "../results/asreml_humann_pathways_all_ranef_w_time/ranef.txt",
  sep="\t", header=T)

pathways.model10.ranef.out.df <- pathways.model10.ranef.df %>% 

  # rename p-value column
  dplyr::rename(pval = LRT.pval) %>% 
    
  # create adjusted p-values, separately within each experimental variable
  group_by(term.clean) %>% 
  mutate(padj = p.adjust(pval, method="BH")) %>% 
  ungroup() %>% 

  # pivot wider
  pivot_wider(id_cols = feature, names_from = term.clean, 
              values_from = c(PVE, pval, padj), names_sep = ".") %>% 
  
  # indicate if community feature
  mutate(is.community.feat = case_when(
    feature %in% community.features ~ "Y",
    TRUE ~ "N")) %>%
  
  # add long pathway name
  merge(pathway.names.uniq.df %>% dplyr::select(original, short.clean),
        by.x="feature", by.y="short.clean", all.x=T) %>% 
  
  # rename 'original' to 'full.pathway.name'n
  dplyr::rename(full.pathway.name = original) %>% 
  
  # specify order of columns
  dplyr::select(feature, full.pathway.name, is.community.feat,
                PVE.Age, pval.Age, padj.Age,
                PVE.Diet, pval.Diet, padj.Diet,
                PVE.Genetics, pval.Genetics, padj.Genetics,
                PVE.Time, pval.Time, padj.Time,
                PVE.Cohort, pval.Cohort, padj.Cohort,
                PVE.Batch, pval.Batch, padj.Batch,
                PVE.Cage, pval.Cage, padj.Cage,
                PVE.Mouse, pval.Mouse, padj.Mouse,
                PVE.Residual, pval.Residual, padj.Residual) %>% 
  
    # round all numbers to 8 decimal places
  mutate_if(is.numeric, ~ round(., digits=8) %>% format()) %>% 
  
  # sort alphabetically
  arrange(feature)
```

```{r}
pathways.model10.ranef.out.df %>%
  write.table("../supp_tables/table14.txt", sep="\t", quote=F, row.names=F)
```
