Identify the phenotype value closest in time to each microbiome sample. Exclude microbiome-phenotype pairs more than 100 days apart.

```{r, message=F, warning=F}
library(tidyverse)
library(foreach)
```

```{r}
# see below for how I arrived at this number
days.bw.mb.and.pheno <- 100
```

# Import microbiome metadata

```{r}
stool.meta.df <- read.table(
  "../data/metadata/stool_metadata_after_QC_no_controls_n2997_230620.txt", 
  sep="\t", header=T)
mouse.meta.df <- read.csv(
  "../data/metadata/AnimalData_Processed_20230712.csv")

stool.meta.annot.df <- stool.meta.df %>%
  merge(mouse.meta.df, by.x="mouse.ID", by.y="MouseID")
```

# Import phenotypes

```{r}
pheno.wide.df <- read.csv("../data/AllData_20230731.csv")
```

For each timepoint-domain combination, record the age of the mouse when the phenotype was collected.

```{r}
timepoint.domain.to.age.df <- pheno.wide.df %>% 
  dplyr::select(MouseID, contains("_AgeInDays")) %>% 
  pivot_longer(-MouseID, names_to="timepoint.domain.pheno", values_to="age.days") %>% 
  separate(timepoint.domain.pheno, into=c("timepoint", "domain", "pheno"),
           sep="_", fill="left", extra="merge", remove=T) %>% 
  dplyr::select(-pheno) %>% 
  dplyr::filter(!is.na(age.days))
```

```{r}
n_distinct(timepoint.domain.to.age.df$domain)
sort(unique(timepoint.domain.to.age.df$domain))
```

* 14 domains (Glucose counted twice)

# Merge microbiome and phenotype age information

For each mouse, create a long dataframe with all combinations of microbiome ages and phenotype ages.

```{r}
mult.pheno.ages.per.mb.age.df <- merge(
  
  # microbiome
  stool.meta.annot.df %>%
    mutate(age.days.mb = age.wks * 7) %>% 
    dplyr::select(mouse.ID, stool.ID, age.days.mb),
  
  # phenotypes
  timepoint.domain.to.age.df %>% 
    dplyr::rename(mouse.ID = MouseID, age.days.pheno = age.days),
  
  by="mouse.ID"
)
```

# Find the closest phenotype

For each stool-domain combination, find the closest phenotype age for that domain.

```{r}
age.df <- mult.pheno.ages.per.mb.age.df %>% 
  group_by(mouse.ID, domain, stool.ID, age.days.mb) %>% 
  summarise(age.days.pheno = age.days.pheno[which.min(abs(age.days.mb-age.days.pheno))], .groups="drop") %>%
  mutate(mb.minus.pheno = age.days.mb - age.days.pheno)
```

# How close are microbiome and phenotype ages?

```{r}
age.df %>% 
  ggplot(aes(mb.minus.pheno)) +
  geom_histogram(bins=100) +
  labs(x="Microbiome sample age in days minus corresponding phenotype age in days")
```

```{r}
summary(age.df$mb.minus.pheno)
```


* Most microbiome-phenotype pairs are very close to each other (± 50 days), but plenty of outliers

# Select a cutoff for how far away a stool sample and phenotype can be

```{r}
age.df %>% 
  mutate(abs.mb.minus.pheno = abs(mb.minus.pheno)) %>% 
  dplyr::filter(abs.mb.minus.pheno < 250) %>% # zoom in
  dplyr::count(abs.mb.minus.pheno) %>% 
  ggplot(aes(x=abs.mb.minus.pheno, y=n)) +
  geom_col() +
  geom_vline(xintercept=c(51, 100, 206), lty=2)
```

* Strictest: 51 would be the strictest
* Most lenient: 206 would allow me to keep the second bump
* 100 is a good compromise

# Apply cutoff

```{r}
age.filt.df <- age.df %>% 
  mutate(abs.mb.minus.pheno = abs(mb.minus.pheno)) %>% 
  dplyr::filter(abs.mb.minus.pheno <= days.bw.mb.and.pheno)

dim(age.df)
dim(age.filt.df)
```

* 41k to 24k microbiome-phenotype combos

# Make sure microbiome and phenotype are both pre-DR or both post-DR

```{r}
age.filt2.df <- age.filt.df %>%
  
  dplyr::filter(!xor(
    (age.days.mb > 180), (age.days.pheno > 180)))

dim(age.filt.df)
dim(age.filt2.df)
nrow(age.filt.df) - nrow(age.filt2.df)
```

* If within 100 days, then we remove 828 combos where microbiome is prior to diet randomization and phenotype is after (or vice versa)

# Pivot phenotypes longer

```{r}
phenos <- colnames(pheno.wide.df) %>% 
  str_subset("_AgeInDays", negate=T) %>% 
  str_subset("^Y")

pheno.long.df <- pheno.wide.df %>% 
  
  # remove AgeInDays columns
  dplyr::select(-ends_with("AgeInDays")) %>% 
  
  # pivot longer
  pivot_longer(cols = all_of(c(phenos, "SurvDays")), 
               names_to="timepoint.domain.pheno") %>% 
  
  # separate the full phenotype into chunks
  separate(timepoint.domain.pheno, into=c("timepoint", "domain", "trait"),
           sep="_", fill="left", extra="merge", remove=F)
```

# Figure out which mouse-domain-timepoint combinations to keep

```{r}
age.filt2.df.w.timepoint <- age.filt2.df %>% 
  dplyr::rename(age.days = age.days.pheno,
                MouseID = mouse.ID) %>% 
  merge(timepoint.domain.to.age.df, by=c("MouseID", "domain", "age.days"))
                
mouse.domain.timepoints.to.keep <- age.filt2.df.w.timepoint %>% 
  distinct(MouseID, domain, timepoint) %>% 
  mutate(mouse.domain.timepoint = paste(MouseID, domain, timepoint, sep="_")) %>% 
  pull(mouse.domain.timepoint)

timepoint.domain.to.age.df %>% 
  distinct(MouseID, domain, timepoint) %>% 
  nrow

length(mouse.domain.timepoints.to.keep)
```

* We subset from 37k to 22k mouse-domain-ages

```{r}
pheno.long.filt.df <- pheno.long.df %>% 
  mutate(mouse.domain.timepoint = paste(MouseID, domain, timepoint, sep="_")) %>% 
  dplyr::filter(mouse.domain.timepoint %in% mouse.domain.timepoints.to.keep)

nrow(pheno.long.df)
nrow(pheno.long.filt.df)
```

* We subset from 688k to 333k individual phenotype values

# Merge phenotype data with stool IDs

```{r}
pheno.long.filt.df.w.stool.ID <- pheno.long.filt.df %>% 
  dplyr::select(MouseID, timepoint, domain, trait, value) %>% 
  merge(age.filt2.df.w.timepoint %>% dplyr::select(MouseID, timepoint, domain, stool.ID), 
        by=c("MouseID", "timepoint", "domain"))

nrow(pheno.long.filt.df)
nrow(pheno.long.filt.df.w.stool.ID)
```

* 333k -> 355k individual phenotype values
* This number grew in size because a few phenotype values correspond to multiple stool IDs, i.e. some stool samples share the same corresponding phenotype

# Make sure each phenotype has data from each dietary group

```{r}
pheno.long.filt.df.w.stool.ID %>% 
  merge(stool.meta.annot.df, by="stool.ID") %>% 
  dplyr::filter(!is.na(value)) %>% 
  distinct(trait, Diet) %>% 
  dplyr::count(trait) %>% 
  arrange(n) %>% head()
```

* PhenoDelta doesn't include all 5 diets
* Omit this phenotype

```{r}
pheno.long.filt.df.w.stool.ID <- pheno.long.filt.df.w.stool.ID %>% 
  dplyr::filter(trait != "PhenoDelta")
```

# Pivot wider

```{r}
pheno.wide.df.stool.ID.in.rows <- pheno.long.filt.df.w.stool.ID %>% 
  mutate(pheno = paste(domain, trait, sep="_")) %>% 
  pivot_wider(id_cols=stool.ID, names_from=pheno, values_from=value)

dim(pheno.wide.df.stool.ID.in.rows)
```

* 2915 stool samples x 210 phenotypes

```{r}
pheno.wide.df.stool.ID.in.rows[1:6, 1:6]
```

# How much data is missing?

```{r}
num.na.per.pheno.df <- pheno.wide.df.stool.ID.in.rows %>% 
  dplyr::select(-stool.ID) %>% 
  lapply(FUN=function(x) {sum(is.na(x))}) %>% 
  data.frame() %>% t() %>% data.frame() %>% 
  setNames("num.na") %>% 
  rownames_to_column("pheno") %>% 
  mutate(frac.na = num.na / nrow(pheno.wide.df.stool.ID.in.rows))

num.na.per.pheno.df %>% 
  ggplot(aes(frac.na)) +
  geom_histogram()
```

* Most phenotypes are >40% missing
* Ton of missing values in this df, but we're always comparing the microbiome against one phenotype at a time, so that's okay

# Add back lifespan

```{r}
pheno.wide.df.stool.ID.in.rows.w.lifespan <- pheno.wide.df.stool.ID.in.rows %>% 
  merge(stool.meta.annot.df %>% dplyr::select(stool.ID, SurvDays), by="stool.ID")
dim(pheno.wide.df.stool.ID.in.rows.w.lifespan)
```

# Export

```{r}
# pheno.wide.df.stool.ID.in.rows.w.lifespan %>%
#   write.table(sprintf(
#     "../results/phenotype_values_closest_to_each_stool_ID_within_%i_days_n%ix%i_230811.txt",
#     days.bw.mb.and.pheno, nrow(.), ncol(.)-1),
#     sep="\t", quote=F, row.names=F)
```
