I prefer running longitudinal associations (to keep power high) but we can't do longitudinal association for lifespan. So I need to run cross-sectional associations.

Model at 5 months: y_pheno ~ y_mb
Model at other ages: y_pheno ~ y_mb + DR

# Load libraries

```{r, warning=F, message=F}
library(tidyverse)
library(foreach)
library(doParallel)
```

```{r}
# minimum number of observations required to do a microbiome-phenotype association
MIN.N <- 200
```

# Set up cluster

```{r}
cl <- makeCluster(6)
registerDoParallel(cl)
```

# Import metadata

```{r}
stool.meta.df <- read.table(
  "../data/metadata/stool_metadata_after_QC_no_controls_n2997_230620.txt", 
  sep="\t", header=T)
mouse.meta.df <- read.csv(
  "../data/metadata/AnimalData_Processed_20230712.csv")

stool.meta.annot.df <- stool.meta.df %>%
  merge(mouse.meta.df, by.x="mouse.ID", by.y="MouseID")
```

# Import microbiome data

## Genera

```{r}
mb.df.feats.in.rows <- read.table(
  "../results/kraken_genus_clr_filt_w_comm_n107x2997_231017.txt",
  sep="\t", header=T, row.names=1)

# we need stool IDs in rows
mb.df <- mb.df.feats.in.rows %>% t() %>% data.frame()

dim(mb.df)
```

* 2997 stool samples x 107 taxonomic features

## Pathways

```{r}
# mb.df.feats.in.rows <- read.table(
#   "../results/DO_pathway_log2tpm_filt_w_comm_n273x2997.txt",
#   sep="\t", header=T, row.names=1)
# 
# # we need stool IDs in rows
# mb.df <- mb.df.feats.in.rows %>% t() %>% data.frame()
# 
# dim(mb.df)
```

* 2997 stool samples x 273 pathways

# Import phenotype data

```{r}
pheno.wide.df <- read.csv(
  "../results/phenotype_values_closest_to_each_stool_ID_within_100_days_n2915x210_230811.txt", 
  sep="\t", header=T, row.names=1)
dim(pheno.wide.df)
```

* 2915 stool IDs x 210 phenotypes

# Combine microbiome and phenotype data

```{r}
mb.feats <- colnames(mb.df)
length(mb.feats)

phenos <- colnames(pheno.wide.df)
length(phenos)
```

```{r}
pheno.and.mb.df <- pheno.wide.df %>%
  merge(mb.df, by="row.names") %>% 
  dplyr::rename(stool.ID = Row.names) %>% 
  merge(stool.meta.annot.df %>% dplyr::select(-SurvDays), by="stool.ID")

dim(pheno.and.mb.df)
```

* 2915 stool IDs

# Run linear model for each phenotype

```{r}
run_lm_one_pheno <- function(this.pheno, this.age.df, reduced.formula, full.formula, mb.feats) {

  cat(this.pheno, "\n")
  
  # make sure phenotype name is correct
  stopifnot(this.pheno %in% colnames(this.age.df))
  
  this.age.filt.df <- this.age.df %>% 
    
    # remove missing phenotype values
    dplyr::filter(!is.na(.data[[this.pheno]]))
  
  this.n <- nrow(this.age.filt.df)
  
  # scale phenotype
  this.age.filt.df[["pheno.scaled"]] <- as.numeric(scale(this.age.filt.df[[this.pheno]]))
  
  # model without microbiome (this step can be run just once per pheno)
  reduced.model <- lm(reduced.formula, data=this.age.filt.df)
  
  # iterate over microbiome features
  this.pheno.res <- sapply(
    mb.feats, FUN=run_lm_one_pheno_one_mb_feat,
    this.pheno=this.pheno, this.df=this.age.filt.df, 
    full.formula=full.formula, 
    reduced.model=reduced.model)
    
  # transpose, convert to df, unnest
  this.pheno.res.df <- this.pheno.res %>% 
    t() %>% data.frame() %>% unnest(everything())
  
  # convert first 2 columns to numeric
  this.pheno.res.df[, 1:2] <- sapply(this.pheno.res.df[, 1:2], as.numeric)
  
  # include n
  this.pheno.res.df$n <- this.n
  
  return(this.pheno.res.df)
  
}

run_lm_one_pheno_one_mb_feat <- function(this.pheno, this.mb.feat, this.df,
                                         full.formula, reduced.model) {
  
  # scale microbiome feature
  this.df[["mb.scaled"]] <- as.numeric(scale(this.df[[this.mb.feat]]))
  
  # full model
  full.model <- lm(full.formula, data=this.df)
  
  # LRT
  this.anova.res <- anova(full.model, reduced.model)
  
  # export desired outputs
  out.vec <- c(
    coef = coef(full.model)["mb.scaled"],
    pval = this.anova.res$`Pr(>F)`[2],
    feature = this.mb.feat,
    pheno = this.pheno)
  
  return(out.vec)
}
```

## 5 months

We'll do 5, 10, 16, 22, and 28 months.

```{r}
this.age <- 5

# subset to this age
this.age.pheno.and.mb.df <- pheno.and.mb.df %>% 
  dplyr::filter(age.approx.months == this.age)

# only include phenotypes with enough observations at this age
this.age.n.per.pheno.df <- data.frame(
  n.non.na = sapply(this.age.pheno.and.mb.df[phenos], FUN=function(x) {length(x[!is.na(x)])}),
  n.uniq = sapply(this.age.pheno.and.mb.df[phenos], FUN=function(x) {n_distinct(x[!is.na(x)])})) %>% 
  rownames_to_column("pheno") 

this.age.phenos <- this.age.n.per.pheno.df %>% 
  dplyr::filter(n.non.na > MIN.N) %>% 
  dplyr::filter(n.uniq > 1) %>% 
  pull(pheno)

length(this.age.phenos)
```

```{r}
# survdays.5mo.res.df <- run_lm_one_pheno(
#   this.pheno="SurvDays", this.age.df=this.age.pheno.and.mb.df,
#   reduced.formula=reduced.formula.5mo, full.formula=full.formula.5mo,
#   mb.feats=mb.feats)
```

```{r}
pheno.res.df.5mo <- foreach(
  this.pheno = this.age.phenos, 
  .combine="rbind",
  .packages = c("tidyverse")) %dopar% {
    
    .GlobalEnv$this.age.pheno.and.mb.df <- this.age.pheno.and.mb.df
    .GlobalEnv$mb.feats <- mb.feats
    
    run_lm_one_pheno(
      this.pheno = this.pheno, 
      this.age.df = this.age.pheno.and.mb.df,
      reduced.formula = as.formula("pheno.scaled ~ 1"), 
      full.formula = as.formula("pheno.scaled ~ mb.scaled"),
      mb.feats = mb.feats) %>% 
      
      # add age to output df
      mutate(age = 5)
    
  }
```

## 10 months

```{r}
this.age <- 10

# subset to this age
this.age.pheno.and.mb.df <- pheno.and.mb.df %>% 
  dplyr::filter(age.approx.months == this.age) %>% 
  mutate(Diet = factor(Diet, levels=c("AL", "1D", "2D", "20", "40")))

# only include phenotypes with enough observations at this age
this.age.n.per.pheno.df <- data.frame(
  n.non.na = sapply(this.age.pheno.and.mb.df[phenos], FUN=function(x) {length(x[!is.na(x)])}),
  n.uniq = sapply(this.age.pheno.and.mb.df[phenos], FUN=function(x) {n_distinct(x[!is.na(x)])})) %>% 
  rownames_to_column("pheno") 

this.age.phenos <- this.age.n.per.pheno.df %>% 
  dplyr::filter(n.non.na > MIN.N) %>% 
  dplyr::filter(n.uniq > 1) %>% 
  pull(pheno)

length(this.age.phenos)
```

```{r}
pheno.res.df.10mo <- foreach(
  this.pheno = this.age.phenos, 
  .combine="rbind",
  .packages = c("tidyverse")) %dopar% {
    
    .GlobalEnv$this.age.pheno.and.mb.df <- this.age.pheno.and.mb.df
    .GlobalEnv$mb.feats <- mb.feats
    
    run_lm_one_pheno(
      this.pheno = this.pheno, 
      this.age.df = this.age.pheno.and.mb.df,
      reduced.formula = as.formula("pheno.scaled ~ Diet"), 
      full.formula = as.formula("pheno.scaled ~ mb.scaled + Diet"),
      mb.feats = mb.feats) %>% 
      
      # add age to output df
      mutate(age = 10)
    
  }
```

## 16 months

```{r}
this.age <- 16

# subset to this age
this.age.pheno.and.mb.df <- pheno.and.mb.df %>% 
  dplyr::filter(age.approx.months == this.age) %>% 
  mutate(Diet = factor(Diet, levels=c("AL", "1D", "2D", "20", "40")))

# only include phenotypes with enough observations at this age
this.age.n.per.pheno.df <- data.frame(
  n.non.na = sapply(this.age.pheno.and.mb.df[phenos], FUN=function(x) {length(x[!is.na(x)])}),
  n.uniq = sapply(this.age.pheno.and.mb.df[phenos], FUN=function(x) {n_distinct(x[!is.na(x)])})) %>% 
  rownames_to_column("pheno") 

this.age.phenos <- this.age.n.per.pheno.df %>% 
  dplyr::filter(n.non.na > MIN.N) %>% 
  dplyr::filter(n.uniq > 1) %>% 
  pull(pheno)

length(this.age.phenos)
```

```{r}
pheno.res.df.16mo <- foreach(
  this.pheno = this.age.phenos, 
  .combine="rbind",
  .packages = c("tidyverse")) %dopar% {
    
    .GlobalEnv$this.age.pheno.and.mb.df <- this.age.pheno.and.mb.df
    .GlobalEnv$mb.feats <- mb.feats
    
    run_lm_one_pheno(
      this.pheno = this.pheno, 
      this.age.df = this.age.pheno.and.mb.df,
      reduced.formula = as.formula("pheno.scaled ~ Diet"), 
      full.formula = as.formula("pheno.scaled ~ mb.scaled + Diet"),
      mb.feats = mb.feats) %>% 
      
      # add age to output df
      mutate(age = 16)
    
  }
```


## 22 months

```{r}
this.age <- 22

# subset to this age
this.age.pheno.and.mb.df <- pheno.and.mb.df %>% 
  dplyr::filter(age.approx.months == this.age) %>% 
  mutate(Diet = factor(Diet, levels=c("AL", "1D", "2D", "20", "40")))

# only include phenotypes with enough observations at this age
this.age.n.per.pheno.df <- data.frame(
  n.non.na = sapply(this.age.pheno.and.mb.df[phenos], FUN=function(x) {length(x[!is.na(x)])}),
  n.uniq = sapply(this.age.pheno.and.mb.df[phenos], FUN=function(x) {n_distinct(x[!is.na(x)])})) %>% 
  rownames_to_column("pheno") 

this.age.phenos <- this.age.n.per.pheno.df %>% 
  dplyr::filter(n.non.na > MIN.N) %>% 
  dplyr::filter(n.uniq > 1) %>% 
  pull(pheno)

length(this.age.phenos)
```

```{r}
pheno.res.df.22mo <- foreach(
  this.pheno = this.age.phenos, 
  .combine="rbind",
  .packages = c("tidyverse")) %dopar% {
    
    .GlobalEnv$this.age.pheno.and.mb.df <- this.age.pheno.and.mb.df
    .GlobalEnv$mb.feats <- mb.feats
    
    run_lm_one_pheno(
      this.pheno = this.pheno, 
      this.age.df = this.age.pheno.and.mb.df,
      reduced.formula = as.formula("pheno.scaled ~ Diet"), 
      full.formula = as.formula("pheno.scaled ~ mb.scaled + Diet"),
      mb.feats = mb.feats) %>% 
      
      # add age to output df
      mutate(age = 22)
    
  }
```

## 28 months

```{r}
this.age <- 28

# subset to this age
this.age.pheno.and.mb.df <- pheno.and.mb.df %>% 
  dplyr::filter(age.approx.months == this.age) %>% 
  mutate(Diet = factor(Diet, levels=c("AL", "1D", "2D", "20", "40")))

# only include phenotypes with enough observations at this age
this.age.n.per.pheno.df <- data.frame(
  n.non.na = sapply(this.age.pheno.and.mb.df[phenos], FUN=function(x) {length(x[!is.na(x)])}),
  n.uniq = sapply(this.age.pheno.and.mb.df[phenos], FUN=function(x) {n_distinct(x[!is.na(x)])})) %>% 
  rownames_to_column("pheno") 

this.age.phenos <- this.age.n.per.pheno.df %>% 
  dplyr::filter(n.non.na > MIN.N) %>% 
  dplyr::filter(n.uniq > 1) %>% 
  pull(pheno)

length(this.age.phenos)
```

```{r}
pheno.res.df.28mo <- foreach(
  this.pheno = this.age.phenos, 
  .combine="rbind",
  .packages = c("tidyverse")) %dopar% {
    
    .GlobalEnv$this.age.pheno.and.mb.df <- this.age.pheno.and.mb.df
    .GlobalEnv$mb.feats <- mb.feats
    
    run_lm_one_pheno(
      this.pheno = this.pheno, 
      this.age.df = this.age.pheno.and.mb.df,
      reduced.formula = as.formula("pheno.scaled ~ Diet"), 
      full.formula = as.formula("pheno.scaled ~ mb.scaled + Diet"),
      mb.feats = mb.feats) %>% 
      
      # add age to output df
      mutate(age = 28)
    
  }
```

## Combine outputs

```{r}
pheno.res.df <- rbind(
  pheno.res.df.5mo,
  pheno.res.df.10mo,
  pheno.res.df.16mo,
  pheno.res.df.22mo,
  pheno.res.df.28mo)
```

```{r}
nrow(pheno.res.df)
n_distinct(pheno.res.df$pheno)
n_distinct(pheno.res.df$feature)
unique(pheno.res.df$age)
```

# Export

```{r}
pheno.res.df %>%
  write.table(
    sprintf("../results/lm_cross_sectional_mb_pheno_assoc_diet_covar_n%i_kraken_feats.txt", length(mb.feats)),
    # sprintf("../results/lm_cross_sectional_mb_pheno_assoc_diet_covar_n%i_pathways.txt", length(mb.feats)),
    sep="\t", row.names=F, quote=F)
```

