We'll prepare 4 datasets at once: DO samples, DO AL samples, B6 samples, and human samples.

Input: genus relabs and pathway RPK
Output: genus log2(relabs) and pathway log2(TPM), plus community features, after prevalence filtering

# Load libraries

```{r, message=F, warning=F}
library(tidyverse)
library(foreach) # %do%
library(vegan) # diversity
```

# Import metadata

## DO

```{r}
stool.meta.df <- read.table(
  "../data/metadata/stool_metadata_after_QC_no_controls_n2997_240418.txt", 
  sep="\t", header=T)
mouse.meta.df <- read.csv(
  "../data/metadata/AnimalData_Processed_20230712.csv")

stool.meta.annot.df <- stool.meta.df %>%
  merge(mouse.meta.df, by.x="mouse.ID", by.y="MouseID")
```

## B6

```{r}
B6.meta.df <- read.table(
  "../data/metadata/B6_sample_metadata.txt",
  sep="\t", header=T, check.names=F)
```

## Humans

```{r}
human.meta.df <- read.table(
  "../data/metadata/CMD_human_sample_metadata_n4101.txt",
  sep="\t", header=T)
```

# Import data

## DO

```{r}
DO.genus.relab.df <- read.table(
  "../data/metaphlan_genus_agg_by_stool_ID_n376x2997.txt",
  sep="\t", header=T, row.names=1)

DO.pathway.tpm.df <- read.table(
  "../data/pathabundance_tpm_agg_by_stool_ID_n422x2997.txt",
  sep="\t", header=T, quote="", row.names=1)

dim(DO.genus.relab.df)
dim(DO.pathway.tpm.df)
```

* N.B. DO pathway abundance is already TPM because I needed to convert to TPM to aggregate by `stool.ID`.

## DO AL

```{r}
tmp.DO.AL.genus.relab.df <- DO.genus.relab.df %>% 
  dplyr::select(all_of(
    stool.meta.annot.df %>% 
      dplyr::filter(Diet == "AL") %>% 
      pull(stool.ID)))

tmp.DO.AL.pathway.tpm.df <- DO.pathway.tpm.df %>% 
  dplyr::select(all_of(
    stool.meta.annot.df %>% 
      dplyr::filter(Diet == "AL") %>% 
      pull(stool.ID)))
```

Remove features that are all zeros after subsetting to AL samples.

```{r}
DO.AL.genus.relab.df <- tmp.DO.AL.genus.relab.df[(rowSums(tmp.DO.AL.genus.relab.df != 0) > 0), ]
DO.AL.pathway.tpm.df <- tmp.DO.AL.pathway.tpm.df[(rowSums(tmp.DO.AL.pathway.tpm.df != 0) > 0), ]

dim(tmp.DO.AL.genus.relab.df)
dim(tmp.DO.AL.pathway.tpm.df)

dim(DO.AL.genus.relab.df)
dim(DO.AL.pathway.tpm.df)
```

## B6

```{r}
B6.mpa.df <- read.table(
  "../data/B6_metaphlan.txt", 
  sep="\t", header=T, check.names=F)
```

Subset to genus-level relative abundances.

```{r}
tmp.B6.genus.relab.df <- B6.mpa.df %>% 
  
  # subset to genera (implicitly excludes UNCLASSIFIED)
  dplyr::filter(str_detect(clade_name, "g__")) %>% 
  dplyr::filter(!str_detect(clade_name, "s__")) %>% 
  
  # make "genus" column
  mutate(genus = str_extract(clade_name, "g__(.*)")) %>% 
  mutate(genus = str_sub(genus, 4)) %>% # remove g__ prefix 
  dplyr::select(-clade_name) %>%
  column_to_rownames("genus")
```

Adjust relabs so they add to 100.

```{r}
B6.genus.relab.df <- 100 * sweep(
  tmp.B6.genus.relab.df, MARGIN=2, STATS=colSums(tmp.B6.genus.relab.df), FUN="/")
```

```{r}
summary(colSums(tmp.B6.genus.relab.df))
summary(colSums(B6.genus.relab.df))
```

```{r}
tmp.B6.pathway.rpk.df <- read.table(
  "../data/B6_humann_pathway_rpk.txt", 
  sep="\t", header=T,
  check.names=F, quote="") # sample IDs contain hyphens and pathway names contain apostrophes
```

Exclude UNMAPPED and UNINTEGRATED.

```{r}
B6.pathway.rpk.df <- tmp.B6.pathway.rpk.df %>% 
  dplyr::filter(!(Pathway %in% c("UNMAPPED", "UNINTEGRATED"))) %>% 
  column_to_rownames("Pathway")
```

## Human

```{r}
tmp.human.genus.relab.df <- read.table(
  "../data/CMD_human_genus_relab_n331x4101.txt", 
  sep="\t", header=T, row.names=1,
  check.names=F) # sample IDs contain hyphens
```

Adjust relabs so they add to 100.

```{r}
human.genus.relab.df <- 100 * sweep(
  tmp.human.genus.relab.df, MARGIN=2, STATS=colSums(tmp.human.genus.relab.df), FUN="/")
```

```{r}
summary(colSums(tmp.human.genus.relab.df))
summary(colSums(human.genus.relab.df))
```

```{r}
tmp.human.pathway.rpk.df <- read.table(
  "../data/CMD_human_pathway_rpk_n573x4101.txt", 
  sep="\t", header=T,
  check.names=F, quote="") # sample IDs contain hyphens and pathway names contain apostrophes
```

Exclude UNMAPPED and UNINTEGRATED.

```{r}
human.pathway.rpk.df <- tmp.human.pathway.rpk.df %>% 
  dplyr::filter(!(pathway_id %in% c("UNMAPPED", "UNINTEGRATED"))) %>% 
  column_to_rownames("pathway_id")
```

# Convert pathways to TPM

DO pathways already TPM.

## B6

```{r}
B6.pathway.tpm.df <- 1e6 * sweep(
  B6.pathway.rpk.df, MARGIN=2, STATS=colSums(B6.pathway.rpk.df), FUN="/")
```

## Human

```{r}
human.pathway.tpm.df <- 1e6 * sweep(
  human.pathway.rpk.df, MARGIN=2, STATS=colSums(human.pathway.rpk.df), FUN="/")
```

# Log transform

First need to replace zeros. Will use the feature-specific non-zero minimum value divided by 2.

## DO

### Genus

```{r}
DO.zero.replacement.per.genus <- apply(DO.genus.relab.df, MARGIN=1, FUN=function(x) {min(x[x != 0]) / 2})
```

```{r}
DO.genus.relab.mat.no.zeros <- foreach(this.genus = rownames(DO.genus.relab.df), .combine="rbind") %do% {
  
  this.genus.vec <- as.numeric(DO.genus.relab.df[this.genus, ])
  this.genus.vec[this.genus.vec == 0] <- DO.zero.replacement.per.genus[[this.genus]]
  
  return(this.genus.vec)
  
}

DO.genus.relab.df.no.zeros <- data.frame(DO.genus.relab.mat.no.zeros)
rownames(DO.genus.relab.df.no.zeros) <- rownames(DO.genus.relab.df)
colnames(DO.genus.relab.df.no.zeros) <- colnames(DO.genus.relab.df)
```

```{r}
DO.genus.log2relab.df <- log2(DO.genus.relab.df.no.zeros)
dim(DO.genus.log2relab.df)
```

### Pathways

```{r}
DO.zero.replacement.per.pathway <- apply(DO.pathway.tpm.df, MARGIN=1, FUN=function(x) {min(x[x != 0]) / 2})
```

```{r}
DO.pathway.tpm.mat.no.zeros <- foreach(this.pathway = rownames(DO.pathway.tpm.df), .combine="rbind") %do% {
  
  this.pathway.vec <- as.numeric(DO.pathway.tpm.df[this.pathway, ])
  this.pathway.vec[this.pathway.vec == 0] <- DO.zero.replacement.per.pathway[[this.pathway]]
  
  return(this.pathway.vec)
  
}

DO.pathway.tpm.df.no.zeros <- data.frame(DO.pathway.tpm.mat.no.zeros)
rownames(DO.pathway.tpm.df.no.zeros) <- rownames(DO.pathway.tpm.df)
colnames(DO.pathway.tpm.df.no.zeros) <- colnames(DO.pathway.tpm.df)
```

```{r}
DO.pathway.log2tpm.df <- log2(DO.pathway.tpm.df.no.zeros)
dim(DO.pathway.log2tpm.df)
```

## DO AL

### Genus

```{r}
DO.AL.zero.replacement.per.genus <- apply(DO.AL.genus.relab.df, MARGIN=1, FUN=function(x) {min(x[x != 0]) / 2})
```

```{r}
DO.AL.genus.relab.mat.no.zeros <- foreach(this.genus = rownames(DO.AL.genus.relab.df), .combine="rbind") %do% {
  
  this.genus.vec <- as.numeric(DO.AL.genus.relab.df[this.genus, ])
  this.genus.vec[this.genus.vec == 0] <- DO.AL.zero.replacement.per.genus[[this.genus]]
  
  return(this.genus.vec)
  
}

DO.AL.genus.relab.df.no.zeros <- data.frame(DO.AL.genus.relab.mat.no.zeros)
rownames(DO.AL.genus.relab.df.no.zeros) <- rownames(DO.AL.genus.relab.df)
colnames(DO.AL.genus.relab.df.no.zeros) <- colnames(DO.AL.genus.relab.df)
```

```{r}
DO.AL.genus.log2relab.df <- log2(DO.AL.genus.relab.df.no.zeros)
dim(DO.AL.genus.log2relab.df)
```

### Pathways

```{r}
DO.AL.zero.replacement.per.pathway <- apply(DO.AL.pathway.tpm.df, MARGIN=1, FUN=function(x) {min(x[x != 0]) / 2})
```

```{r}
DO.AL.pathway.tpm.mat.no.zeros <- foreach(this.pathway = rownames(DO.AL.pathway.tpm.df), .combine="rbind") %do% {
  
  this.pathway.vec <- as.numeric(DO.AL.pathway.tpm.df[this.pathway, ])
  this.pathway.vec[this.pathway.vec == 0] <- DO.AL.zero.replacement.per.pathway[[this.pathway]]
  
  return(this.pathway.vec)
  
}

DO.AL.pathway.tpm.df.no.zeros <- data.frame(DO.AL.pathway.tpm.mat.no.zeros)
rownames(DO.AL.pathway.tpm.df.no.zeros) <- rownames(DO.AL.pathway.tpm.df)
colnames(DO.AL.pathway.tpm.df.no.zeros) <- colnames(DO.AL.pathway.tpm.df)
```

```{r}
DO.AL.pathway.log2tpm.df <- log2(DO.AL.pathway.tpm.df.no.zeros)
dim(DO.AL.pathway.log2tpm.df)
```

## B6

### Genus

```{r}
B6.zero.replacement.per.genus <- apply(B6.genus.relab.df, MARGIN=1, FUN=function(x) {min(x[x != 0]) / 2})
```

```{r}
B6.genus.relab.mat.no.zeros <- foreach(this.genus = rownames(B6.genus.relab.df), .combine="rbind") %do% {
  
  this.genus.vec <- as.numeric(B6.genus.relab.df[this.genus, ])
  this.genus.vec[this.genus.vec == 0] <- B6.zero.replacement.per.genus[[this.genus]]
  
  return(this.genus.vec)
  
}

B6.genus.relab.df.no.zeros <- data.frame(B6.genus.relab.mat.no.zeros)
rownames(B6.genus.relab.df.no.zeros) <- rownames(B6.genus.relab.df)
colnames(B6.genus.relab.df.no.zeros) <- colnames(B6.genus.relab.df)
```

```{r}
B6.genus.log2relab.df <- log2(B6.genus.relab.df.no.zeros)
dim(B6.genus.log2relab.df)
```

### Pathways

```{r}
B6.zero.replacement.per.pathway <- apply(B6.pathway.tpm.df, MARGIN=1, FUN=function(x) {min(x[x != 0]) / 2})
```

```{r}
B6.pathway.tpm.mat.no.zeros <- foreach(this.pathway = rownames(B6.pathway.tpm.df), .combine="rbind") %do% {
  
  this.pathway.vec <- as.numeric(B6.pathway.tpm.df[this.pathway, ])
  this.pathway.vec[this.pathway.vec == 0] <- B6.zero.replacement.per.pathway[[this.pathway]]
  
  return(this.pathway.vec)
  
}

B6.pathway.tpm.df.no.zeros <- data.frame(B6.pathway.tpm.mat.no.zeros)
rownames(B6.pathway.tpm.df.no.zeros) <- rownames(B6.pathway.tpm.df)
colnames(B6.pathway.tpm.df.no.zeros) <- colnames(B6.pathway.tpm.df)
```

```{r}
B6.pathway.log2tpm.df <- log2(B6.pathway.tpm.df.no.zeros)
dim(B6.pathway.log2tpm.df)
```

## Human

### Genus

```{r}
human.zero.replacement.per.genus <- apply(human.genus.relab.df, MARGIN=1, FUN=function(x) {min(x[x != 0]) / 2})
```

```{r}
human.genus.relab.mat.no.zeros <- foreach(this.genus = rownames(human.genus.relab.df), .combine="rbind") %do% {
  
  this.genus.vec <- as.numeric(human.genus.relab.df[this.genus, ])
  this.genus.vec[this.genus.vec == 0] <- human.zero.replacement.per.genus[[this.genus]]
  
  return(this.genus.vec)
  
}

human.genus.relab.df.no.zeros <- data.frame(human.genus.relab.mat.no.zeros)
rownames(human.genus.relab.df.no.zeros) <- rownames(human.genus.relab.df)
colnames(human.genus.relab.df.no.zeros) <- colnames(human.genus.relab.df)
```

```{r}
human.genus.log2relab.df <- log2(human.genus.relab.df.no.zeros)
```

### Pathways

```{r}
human.zero.replacement.per.pathway <- apply(human.pathway.tpm.df, MARGIN=1, FUN=function(x) {min(x[x != 0]) / 2})
```

```{r}
human.pathway.tpm.mat.no.zeros <- foreach(this.pathway = rownames(human.pathway.tpm.df), .combine="rbind") %do% {
  
  this.pathway.vec <- as.numeric(human.pathway.tpm.df[this.pathway, ])
  this.pathway.vec[this.pathway.vec == 0] <- human.zero.replacement.per.pathway[[this.pathway]]
  
  return(this.pathway.vec)
  
}

human.pathway.tpm.df.no.zeros <- data.frame(human.pathway.tpm.mat.no.zeros)
rownames(human.pathway.tpm.df.no.zeros) <- rownames(human.pathway.tpm.df)
colnames(human.pathway.tpm.df.no.zeros) <- colnames(human.pathway.tpm.df)
```

```{r}
human.pathway.log2tpm.df <- log2(human.pathway.tpm.df.no.zeros)
```

# Calculate alpha diversity

## DO

### Genus

```{r}
DO.genus.alpha.div.long.df <- data.frame(
  Shannon = diversity(DO.genus.relab.df, index="shannon", MARGIN=2),
  Simpson = diversity(DO.genus.relab.df, index="simpson", MARGIN=2),
  InvSimpson = diversity(DO.genus.relab.df, index="invsimpson", MARGIN=2),
  Observed = specnumber(DO.genus.relab.df, MARGIN=2)) %>% 
  rownames_to_column("stool.ID") %>% 
  pivot_longer(-stool.ID, names_to="metric")
```

## DO AL

### Genus

```{r}
DO.AL.genus.alpha.div.long.df <- data.frame(
  Shannon = diversity(DO.AL.genus.relab.df, index="shannon", MARGIN=2),
  Simpson = diversity(DO.AL.genus.relab.df, index="simpson", MARGIN=2),
  InvSimpson = diversity(DO.AL.genus.relab.df, index="invsimpson", MARGIN=2),
  Observed = specnumber(DO.AL.genus.relab.df, MARGIN=2)) %>% 
  rownames_to_column("stool.ID") %>% 
  pivot_longer(-stool.ID, names_to="metric")
```

## B6

### Genus

```{r}
B6.genus.alpha.div.long.df <- data.frame(
  Shannon = diversity(B6.genus.relab.df, index="shannon", MARGIN=2),
  Simpson = diversity(B6.genus.relab.df, index="simpson", MARGIN=2),
  InvSimpson = diversity(B6.genus.relab.df, index="invsimpson", MARGIN=2),
  Observed = specnumber(B6.genus.relab.df, MARGIN=2)) %>% 
  rownames_to_column("id") %>% 
  pivot_longer(-id, names_to="metric")
```

## Human

### Genus

```{r}
human.genus.alpha.div.long.df <- data.frame(
  Shannon = diversity(human.genus.relab.df, index="shannon", MARGIN=2),
  Simpson = diversity(human.genus.relab.df, index="simpson", MARGIN=2),
  InvSimpson = diversity(human.genus.relab.df, index="invsimpson", MARGIN=2),
  Observed = specnumber(human.genus.relab.df, MARGIN=2)) %>% 
  rownames_to_column("sample_id") %>% 
  pivot_longer(-sample_id, names_to="metric")
```

# Calculate uniqueness

For taxonomy, we'll use Bray-Curtis on relabs.

For pathways, we'll use Euclidean on log2(TPM).

## DO

### Genus

```{r}
DO.genus.relab.bc.dist.mat <- DO.genus.relab.df %>% 
  t() %>% vegdist(method="bray") %>% 
  as.matrix()
diag(DO.genus.relab.bc.dist.mat) <- NA
DO.genus.uniq.df <- data.frame(
  uniqueness=apply(DO.genus.relab.bc.dist.mat, MARGIN=2, FUN=function(x) {min(x, na.rm=T)})) %>% 
  rownames_to_column("stool.ID")
```

### Pathways

```{r}
DO.pathway.dist.mat <- DO.pathway.log2tpm.df %>% 
  t() %>% dist() %>% as.matrix()
diag(DO.pathway.dist.mat) <- NA
DO.pathway.uniq.df <- data.frame(
  uniqueness=apply(DO.pathway.dist.mat, MARGIN=2, FUN=function(x) {min(x, na.rm=T)})) %>% 
  rownames_to_column("stool.ID")
```

## DO AL

### Genus

```{r}
DO.AL.genus.relab.bc.dist.mat <- DO.AL.genus.relab.df %>% 
  t() %>% vegdist(method="bray") %>% 
  as.matrix()
diag(DO.AL.genus.relab.bc.dist.mat) <- NA
DO.AL.genus.uniq.df <- data.frame(
  uniqueness=apply(DO.AL.genus.relab.bc.dist.mat, MARGIN=2, FUN=function(x) {min(x, na.rm=T)})) %>% 
  rownames_to_column("stool.ID")
```

### Pathways

```{r}
DO.AL.pathway.dist.mat <- DO.AL.pathway.log2tpm.df %>% 
  t() %>% dist() %>% as.matrix()
diag(DO.AL.pathway.dist.mat) <- NA
DO.AL.pathway.uniq.df <- data.frame(
  uniqueness=apply(DO.AL.pathway.dist.mat, MARGIN=2, FUN=function(x) {min(x, na.rm=T)})) %>% 
  rownames_to_column("stool.ID")
```

## B6

### Genus

```{r}
B6.genus.relab.bc.dist.mat <- B6.genus.relab.df %>% 
  t() %>% vegdist(method="bray") %>% 
  as.matrix()
diag(B6.genus.relab.bc.dist.mat) <- NA
B6.genus.uniq.df <- data.frame(
  uniqueness=apply(B6.genus.relab.bc.dist.mat, MARGIN=2, FUN=function(x) {min(x, na.rm=T)})) %>% 
  rownames_to_column("id")
```

### Pathways

```{r}
B6.pathway.dist.mat <- B6.pathway.log2tpm.df %>% 
  t() %>% dist() %>% as.matrix()
diag(B6.pathway.dist.mat) <- NA
B6.pathway.uniq.df <- data.frame(
  uniqueness=apply(B6.pathway.dist.mat, MARGIN=2, FUN=function(x) {min(x, na.rm=T)})) %>% 
  rownames_to_column("id")
```

## Human

Will compute uniqueness separately within study.

### Genus

```{r}
genus.uniq.df.across.studies <- foreach(this.study = unique(human.meta.df$study_name), 
                                        .combine="rbind") %do% {
  
  # subset to this study
  sample.ids.for.this.study <- human.meta.df %>% 
    dplyr::filter(study_name == this.study) %>% 
    pull(sample_id)
  this.study.tmp.genus.relab.df <- human.genus.relab.df[, sample.ids.for.this.study]
  
  # remove genera with all zero relabs
  this.study.genus.relab.df <- this.study.tmp.genus.relab.df[rowSums(this.study.tmp.genus.relab.df) > 0, ]
  
  # compute Bray-Curtis distance
  this.study.bc.dist.mat <- this.study.genus.relab.df %>% 
    t() %>% vegdist(method="bray") %>% 
    as.matrix()
  
  # compute uniqueness
  diag(this.study.bc.dist.mat) <- NA
  this.uniq.df <- data.frame(
    uniqueness=apply(this.study.bc.dist.mat, MARGIN=2, FUN=function(x) {min(x, na.rm=T)})) %>% 
    rownames_to_column("sample_id")

  return(this.uniq.df)
}
```

### Pathways

```{r}
pathway.uniq.df.across.studies <- foreach(this.study = unique(human.meta.df$study_name), 
                                        .combine="rbind") %do% {
                                          
  # subset to this study
  sample.ids.for.this.study <- human.meta.df %>% 
    dplyr::filter(study_name == this.study) %>% 
    pull(sample_id)
  this.study.pathway.log2tpm.df <- human.pathway.log2tpm.df[, sample.ids.for.this.study]
  
  # compute Euclidean distance
  this.study.dist.mat <- this.study.pathway.log2tpm.df %>% 
    t() %>% dist() %>% as.matrix()
  
  # compute uniqueness
  diag(this.study.dist.mat) <- NA
  this.uniq.df <- data.frame(
    uniqueness=apply(this.study.dist.mat, MARGIN=2, FUN=function(x) {min(x, na.rm=T)})) %>% 
    rownames_to_column("sample_id")

  return(this.uniq.df)
}
```

# Prevalence filtering

Relab filtration: 0.001% in at least 10% of samples
TPM filtration: 100 in at least 10% of samples

```{r}
MIN.RELAB.ABUNDANCE <- 0.001
MIN.TPM.ABUNDANCE <- 100
MIN.PREVALENCE <- 0.1
```

Use relab and TPM data.

## DO

```{r}
DO.genera.to.keep <- rownames(DO.genus.relab.df)[(rowSums(DO.genus.relab.df > MIN.RELAB.ABUNDANCE) / ncol(DO.genus.relab.df)) > MIN.PREVALENCE]

nrow(DO.genus.relab.df)
length(DO.genera.to.keep)
```

* 376 --> 254 genera

```{r}
DO.pathways.to.keep <- rownames(DO.pathway.tpm.df)[(rowSums(DO.pathway.tpm.df > MIN.TPM.ABUNDANCE) / ncol(DO.pathway.tpm.df)) > MIN.PREVALENCE]

nrow(DO.pathway.tpm.df)
length(DO.pathways.to.keep)
```

* 422 --> 272 pathways

## DO AL

```{r}
DO.AL.genera.to.keep <- rownames(DO.AL.genus.relab.df)[(rowSums(DO.AL.genus.relab.df > MIN.RELAB.ABUNDANCE) / ncol(DO.AL.genus.relab.df)) > MIN.PREVALENCE]

nrow(DO.AL.genus.relab.df)
length(DO.AL.genera.to.keep)
```

* 311 --> 248 genera

```{r}
DO.AL.pathways.to.keep <- rownames(DO.AL.pathway.tpm.df)[(rowSums(DO.AL.pathway.tpm.df > MIN.TPM.ABUNDANCE) / ncol(DO.AL.pathway.tpm.df)) > MIN.PREVALENCE]

nrow(DO.AL.pathway.tpm.df)
length(DO.AL.pathways.to.keep)
```

* 399 --> 262 pathways

## B6

```{r}
B6.genera.to.keep <- rownames(B6.genus.relab.df)[(rowSums(B6.genus.relab.df > MIN.RELAB.ABUNDANCE) / ncol(B6.genus.relab.df)) > MIN.PREVALENCE]

nrow(B6.genus.relab.df)
length(B6.genera.to.keep)
```

* 310 --> 262 genera

```{r}
B6.pathways.to.keep <- rownames(B6.pathway.tpm.df)[(rowSums(B6.pathway.tpm.df > MIN.TPM.ABUNDANCE) / ncol(B6.pathway.tpm.df)) > MIN.PREVALENCE]

nrow(B6.pathway.tpm.df)
length(B6.pathways.to.keep)
```

* 329 --> 233 pathways

## Human

```{r}
human.genera.to.keep <- rownames(human.genus.relab.df)[(rowSums(human.genus.relab.df > MIN.RELAB.ABUNDANCE) / ncol(human.genus.relab.df)) > MIN.PREVALENCE]

nrow(human.genus.relab.df)
length(human.genera.to.keep)
```

* 331 --> 90 genera

```{r}
human.pathways.to.keep <- rownames(human.pathway.tpm.df)[(rowSums(human.pathway.tpm.df > MIN.TPM.ABUNDANCE) / ncol(human.pathway.tpm.df)) > MIN.PREVALENCE]

nrow(human.pathway.tpm.df)
length(human.pathways.to.keep)
```

* 573 --> 358 pathways

# Combine individual and community features

Easiest to do this when all dataframes are in long form. 

## DO

### Genus

```{r}
DO.genus.all.feats.long.df <- rbind(
  
  # log2(relab)
  DO.genus.log2relab.df %>% 
    rownames_to_column("feature") %>% 
    dplyr::filter(feature %in% DO.genera.to.keep) %>% 
    pivot_longer(-feature, names_to="stool.ID"),
  
  # alpha div
  DO.genus.alpha.div.long.df %>% 
    dplyr::rename(feature = metric) %>% 
    dplyr::select(feature, stool.ID, value),
  
  # uniq
  DO.genus.uniq.df %>% 
    mutate(feature = "Uniqueness") %>% 
    dplyr::rename(value = uniqueness) %>% 
    dplyr::select(feature, stool.ID, value)
  
)
```

```{r}
DO.genus.all.feats.wide.df <- DO.genus.all.feats.long.df %>% 
  pivot_wider(id_cols=feature, names_from=stool.ID, values_from=value)

dim(DO.genus.all.feats.wide.df)
```

### Pathways

```{r}
DO.pathway.all.feats.long.df <- rbind(
  
  # log2(TPM)
  DO.pathway.log2tpm.df %>% 
    rownames_to_column("feature") %>% 
    dplyr::filter(feature %in% DO.pathways.to.keep) %>% 
    pivot_longer(-feature, names_to="stool.ID"),
  
  # uniq
  DO.pathway.uniq.df %>% 
    mutate(feature = "Uniqueness") %>% 
    dplyr::rename(value = uniqueness) %>% 
    dplyr::select(feature, stool.ID, value)
  
)
```

```{r}
DO.pathway.all.feats.wide.df <- DO.pathway.all.feats.long.df %>% 
  pivot_wider(id_cols=feature, names_from=stool.ID, values_from=value)

dim(DO.pathway.all.feats.wide.df)
```

## DO AL

### Genus

```{r}
DO.AL.genus.all.feats.long.df <- rbind(
  
  # log2(relab)
  DO.AL.genus.log2relab.df %>% 
    rownames_to_column("feature") %>% 
    dplyr::filter(feature %in% DO.AL.genera.to.keep) %>% 
    pivot_longer(-feature, names_to="stool.ID"),
  
  # alpha div
  DO.AL.genus.alpha.div.long.df %>% 
    dplyr::rename(feature = metric) %>% 
    dplyr::select(feature, stool.ID, value),
  
  # uniq
  DO.AL.genus.uniq.df %>% 
    mutate(feature = "Uniqueness") %>% 
    dplyr::rename(value = uniqueness) %>% 
    dplyr::select(feature, stool.ID, value)
  
)
```

```{r}
DO.AL.genus.all.feats.wide.df <- DO.AL.genus.all.feats.long.df %>% 
  pivot_wider(id_cols=feature, names_from=stool.ID, values_from=value)

dim(DO.AL.genus.all.feats.wide.df)
```

### Pathways

```{r}
DO.AL.pathway.all.feats.long.df <- rbind(
  
  # log2(TPM)
  DO.AL.pathway.log2tpm.df %>% 
    rownames_to_column("feature") %>% 
    dplyr::filter(feature %in% DO.AL.pathways.to.keep) %>% 
    pivot_longer(-feature, names_to="stool.ID"),
  
  # uniq
  DO.AL.pathway.uniq.df %>% 
    mutate(feature = "Uniqueness") %>% 
    dplyr::rename(value = uniqueness) %>% 
    dplyr::select(feature, stool.ID, value)
  
)
```

```{r}
DO.AL.pathway.all.feats.wide.df <- DO.AL.pathway.all.feats.long.df %>% 
  pivot_wider(id_cols=feature, names_from=stool.ID, values_from=value)

dim(DO.AL.pathway.all.feats.wide.df)
```

## B6

### Genus

```{r}
B6.genus.all.feats.long.df <- rbind(
  
  # log2(relab)
  B6.genus.log2relab.df %>% 
    rownames_to_column("feature") %>% 
    dplyr::filter(feature %in% B6.genera.to.keep) %>% 
    pivot_longer(-feature, names_to="id"),
  
  # alpha div
  B6.genus.alpha.div.long.df %>% 
    dplyr::rename(feature = metric) %>% 
    dplyr::select(feature, id, value),
  
  # uniq
  B6.genus.uniq.df %>% 
    mutate(feature = "Uniqueness") %>% 
    dplyr::rename(value = uniqueness) %>% 
    dplyr::select(feature, id, value)
  
)
```

```{r}
B6.genus.all.feats.wide.df <- B6.genus.all.feats.long.df %>% 
  pivot_wider(id_cols=feature, names_from=id, values_from=value)

dim(B6.genus.all.feats.wide.df)
```

### Pathways

```{r}
B6.pathway.all.feats.long.df <- rbind(
  
  # log2(TPM)
  B6.pathway.log2tpm.df %>% 
    rownames_to_column("feature") %>% 
    dplyr::filter(feature %in% B6.pathways.to.keep) %>% 
    pivot_longer(-feature, names_to="id"),
  
  # uniq
  B6.pathway.uniq.df %>% 
    mutate(feature = "Uniqueness") %>% 
    dplyr::rename(value = uniqueness) %>% 
    dplyr::select(feature, id, value)
  
)
```

```{r}
B6.pathway.all.feats.wide.df <- B6.pathway.all.feats.long.df %>% 
  pivot_wider(id_cols=feature, names_from=id, values_from=value)

dim(B6.pathway.all.feats.wide.df)
```

## Human

### Genus

```{r}
human.genus.all.feats.long.df <- rbind(
  
  # log2(relab)
  human.genus.log2relab.df %>% 
    rownames_to_column("feature") %>% 
    dplyr::filter(feature %in% human.genera.to.keep) %>% 
    pivot_longer(-feature, names_to="sample_id"),
  
  # alpha div
  human.genus.alpha.div.long.df %>% 
    dplyr::rename(feature = metric) %>% 
    dplyr::select(feature, sample_id, value),
  
  # uniq
  genus.uniq.df.across.studies %>% 
    mutate(feature = "Uniqueness") %>% 
    dplyr::rename(value = uniqueness) %>% 
    dplyr::select(feature, sample_id, value)
  
)
```

```{r}
human.genus.all.feats.wide.df <- human.genus.all.feats.long.df %>% 
  pivot_wider(id_cols=feature, names_from=sample_id, values_from=value)

dim(human.genus.all.feats.wide.df)
```

### Pathways

```{r}
human.pathway.all.feats.long.df <- rbind(
  
  # log2(TPM)
  human.pathway.log2tpm.df %>% 
    rownames_to_column("feature") %>% 
    dplyr::filter(feature %in% human.pathways.to.keep) %>% 
    pivot_longer(-feature, names_to="sample_id"),
  
  # uniq
  pathway.uniq.df.across.studies %>% 
    mutate(feature = "Uniqueness") %>% 
    dplyr::rename(value = uniqueness) %>% 
    dplyr::select(feature, sample_id, value)
  
)
```

```{r}
human.pathway.all.feats.wide.df <- human.pathway.all.feats.long.df %>% 
  pivot_wider(id_cols=feature, names_from=sample_id, values_from=value)

dim(human.pathway.all.feats.wide.df)
```

# Clean up pathway names so they don't cause trouble later

```{r}
all.pathways <- union(union(
  rownames(DO.pathway.tpm.df),
  rownames(B6.pathway.rpk.df)),
  rownames(human.pathway.rpk.df))

length(all.pathways)
```

* 703 pathways across all 3 datasets

```{r}
pathway.names.df <- data.frame(
  original = all.pathways) %>% 
  
  # extract the pathway ID (before the colon)
  mutate(short = str_extract(original, "(.*):")) %>% 
  mutate(short = str_sub(short, 1, nchar(short)-1)) %>% 
  
  # remove punctuation
  mutate(short.clean = str_replace_all(short, "[:punct:]", "")) %>% 
  
  # for some reason, + not included in punct?
  mutate(short.clean = str_replace_all(short.clean, "\\+", "")) %>%
  
  # prefix with X if it starts with a number
  mutate(short.clean = case_when(
    str_detect(short.clean, "^[:digit:]") ~ paste0("X", short.clean),
    TRUE ~ short.clean)) %>% 
  
  arrange(short.clean) %>% 
  relocate(short.clean) 
```

```{r}
DO.pathway.all.feats.wide.out.df <- DO.pathway.all.feats.wide.df %>% 
  merge(pathway.names.df %>% dplyr::select(original, short.clean),
        by.x="feature", by.y="original",
        all.x=T) %>% # to keep Uniqueness
  mutate(short.clean = ifelse(is.na(short.clean), feature, short.clean)) %>% # for Uniqueness
  dplyr::select(-feature) %>% 
  relocate(short.clean) %>% 
  dplyr::rename(feature = short.clean)

DO.pathway.all.feats.wide.out.df[1:3, 1:3]
```

```{r}
DO.AL.pathway.all.feats.wide.out.df <- DO.AL.pathway.all.feats.wide.df %>% 
  merge(pathway.names.df %>% dplyr::select(original, short.clean),
        by.x="feature", by.y="original",
        all.x=T) %>% # to keep Uniqueness
  mutate(short.clean = ifelse(is.na(short.clean), feature, short.clean)) %>% # for Uniqueness
  dplyr::select(-feature) %>% 
  relocate(short.clean) %>% 
  dplyr::rename(feature = short.clean)

DO.AL.pathway.all.feats.wide.out.df[1:3, 1:3]
```

```{r}
B6.pathway.all.feats.wide.out.df <- B6.pathway.all.feats.wide.df %>% 
  merge(pathway.names.df %>% dplyr::select(original, short.clean),
        by.x="feature", by.y="original",
        all.x=T) %>% # to keep Uniqueness
  mutate(short.clean = ifelse(is.na(short.clean), feature, short.clean)) %>% # for Uniqueness
  dplyr::select(-feature) %>% 
  relocate(short.clean) %>% 
  dplyr::rename(feature = short.clean)

B6.pathway.all.feats.wide.out.df[1:3, 1:3]
```

```{r}
human.pathway.all.feats.wide.out.df <- human.pathway.all.feats.wide.df %>% 
  merge(pathway.names.df %>% dplyr::select(original, short.clean),
        by.x="feature", by.y="original",
        all.x=T) %>% # to keep Uniqueness
  mutate(short.clean = ifelse(is.na(short.clean), feature, short.clean)) %>% # for Uniqueness
  dplyr::select(-feature) %>% 
  relocate(short.clean) %>% 
  dplyr::rename(feature = short.clean)

human.pathway.all.feats.wide.out.df[1:3, 1:3]
```

Make sure we don't export the wrong ones:

```{r}
rm(
  DO.pathway.all.feats.wide.df,
  DO.AL.pathway.all.feats.wide.df,
  B6.pathway.all.feats.wide.df, 
  human.pathway.all.feats.wide.df)
```

# Export

```{r}
dim(DO.AL.genus.all.feats.wide.df)
dim(B6.genus.all.feats.wide.df)
dim(human.genus.all.feats.wide.df)
```

```{r}
dim(DO.pathway.all.feats.wide.out.df)
dim(DO.AL.pathway.all.feats.wide.out.df)
dim(B6.pathway.all.feats.wide.out.df)
dim(human.pathway.all.feats.wide.out.df)
```

## DO

```{r}
DO.genus.all.feats.wide.df %>%
  write.table(sprintf(
    "../results/metaphlan_genus_log2relab_filt_w_comm_n%ix%i.txt",
    nrow(.), ncol(.)-1),
    sep="\t", row.names=F, quote=F)
```

```{r}
DO.pathway.all.feats.wide.out.df %>%
  write.table(sprintf(
    "../results/pathway_log2tpm_filt_w_comm_n%ix%i.txt",
    nrow(.), ncol(.)-1),
    sep="\t", row.names=F, quote=F)
```

## DO AL

```{r}
DO.AL.genus.all.feats.wide.df %>%
  write.table(sprintf(
    "../results/DO_AL_genus_log2relab_filt_w_comm_n%ix%i.txt",
    nrow(.), ncol(.)-1),
    sep="\t", row.names=F, quote=F)
```

```{r}
DO.AL.pathway.all.feats.wide.out.df %>%
  write.table(sprintf(
    "../results/DO_AL_pathway_log2tpm_filt_w_comm_n%ix%i.txt",
    nrow(.), ncol(.)-1),
    sep="\t", row.names=F, quote=F)
```

## B6

```{r}
B6.genus.all.feats.wide.df %>%
  write.table(sprintf(
    "../results/B6_genus_log2relab_filt_w_comm_n%ix%i.txt",
    nrow(.), ncol(.)-1),
    sep="\t", row.names=F, quote=F)
```

```{r}
B6.pathway.all.feats.wide.out.df %>%
  write.table(sprintf(
    "../results/B6_pathway_log2tpm_filt_w_comm_n%ix%i.txt",
    nrow(.), ncol(.)-1),
    sep="\t", row.names=F, quote=F)
```

## Human

```{r}
human.genus.all.feats.wide.df %>%
  write.table(sprintf(
    "../results/CMD_human_genus_log2relab_filt_w_comm_n%ix%i.txt",
    nrow(.), ncol(.)-1),
    sep="\t", row.names=F, quote=F)
```

```{r}
human.pathway.all.feats.wide.out.df %>%
  write.table(sprintf(
    "../results/CMD_human_pathway_log2tpm_filt_w_comm_n%ix%i.txt",
    nrow(.), ncol(.)-1),
    sep="\t", row.names=F, quote=F)
```
