Input: genus relabs (n=252)
Output: genus CLR (n=100) plus 7 community features: Chao1, Shannon, Simpson, uniqueness, PCoA1-3

# Load libraries

```{r, message=F, warning=F}
library(tidyverse)
library(phyloseq)
library(speedyseq) # faster tax_glom
library(foreach)
library(compositions) # clr
library(vegan) # diversity
```

# Import metadata

```{r}
stool.meta.df <- read.table(
  "../data/metadata/stool_metadata_after_QC_no_controls_n2997_230620.txt", 
  sep="\t", header=T)
mouse.meta.df <- read.csv("../data/metadata/AnimalData_Processed_20230712.csv")

stool.meta.annot.df <- stool.meta.df %>%
  merge(mouse.meta.df, by.x="mouse.ID", by.y="MouseID")
```

# Import Kraken data

```{r}
agg.kraken.df <- read.table(
  "../data/kraken_matrix_agg_by_stool_ID_n1303x2997.txt", 
  sep="\t", header=T, row.names=1)
dim(agg.kraken.df)
```

# Import Kraken taxonomy

```{r}
kraken.tax.df <- read.table("../data/kraken_taxonomy_n1303.txt", sep="\t", header=T, quote="") %>%
  column_to_rownames("taxon.ID")
dim(kraken.tax.df)
```

# Create phyloseq object

```{r}
kraken.sample.meta.df <- stool.meta.annot.df %>% column_to_rownames("stool.ID")

agg.physeq <- phyloseq(
  agg.kraken.df %>% as.matrix %>% otu_table(taxa_are_rows=T),
  kraken.sample.meta.df %>% sample_data,
  kraken.tax.df %>% as.matrix %>% tax_table
)
nsamples(agg.physeq)
```

* 2997 samples

# Aggregate to genera

```{r}
agg.physeq.genus <- agg.physeq %>% tax_glom(taxrank="genus")
```

# Identify top 100 genera

```{r}
N.GENERA <- 100
TOP.N.GENERA.IDs <- names(sort(taxa_sums(agg.physeq.genus), decreasing=T))[1:N.GENERA]
TOP.N.GENERA <- data.frame(tax_table(agg.physeq.genus))[TOP.N.GENERA.IDs, "genus"]
TOP.N.GENERA.no.special <- str_replace_all(TOP.N.GENERA, "[[:punct:]]", "")

# if genus starts with a number, prefix with X
# this helps avoid problems down the line, like providing the genus name within a lm formula
TOP.N.GENERA.CLEAN <- as.character(sapply(
  TOP.N.GENERA.no.special, function(x) {
    ifelse(str_detect(x, "^[:digit:]"), paste0("X", x), x)}))
```

```{r}
top.n.genera.df <- data.frame(
  genus = TOP.N.GENERA,
  clean = TOP.N.GENERA.CLEAN)
rownames(top.n.genera.df) <- TOP.N.GENERA.IDs
```

# Export relab for top 100 genera

We use this in `prediction.Rmd`, but not in linear modeling.

```{r}
genus.relab.df <- agg.physeq.genus %>%
  transform_sample_counts(function(x) {100 * x/sum(x)}) %>% 
  prune_taxa(taxa = TOP.N.GENERA.IDs) %>% 
  otu_table() %>% data.frame()

rownames(genus.relab.df) <- top.n.genera.df[rownames(genus.relab.df), "clean"]
```

```{r}
# genus.relab.df %>%
#   rownames_to_column("genus") %>%
#   write.table(sprintf(
#     "../data/kraken_genus_relab_n%ix%i.txt",
#     nrow(.), ncol(.)-1),
#     sep="\t", row.names=F, quote=F)
```

# Convert to CLR

```{r}
# convert to CLR
genus.clr.df <- agg.physeq.genus %>% 
  otu_table() %>% t() %>% 
  compositions::clr() %>% 
  t() %>% data.frame()

# subset to top genera
genus.clr.filt.df <- genus.clr.df[TOP.N.GENERA.IDs, ]

# replace IDs with names
rownames(genus.clr.filt.df) <- TOP.N.GENERA.CLEAN

# pivot longer
genus.clr.long.df <- genus.clr.filt.df %>% 
  rownames_to_column("taxon") %>% 
  pivot_longer(-taxon, names_to="stool.ID")
```

# Calculate alpha diversity

```{r}
alpha.div.df.genus <- estimate_richness(
  agg.physeq.genus, 
  measures=c("Observed", "Chao1", "Shannon", "Simpson")) %>% 
  rownames_to_column("stool.ID")

alpha.div.long.df.genus <- alpha.div.df.genus %>% 
  pivot_longer(c(Observed, Chao1, Shannon, Simpson), names_to="metric")
```

# Calculate uniqueness

Bray-Curtis on genus-level relative abundances.

```{r}
agg.genus.bc.dist <- agg.physeq.genus %>%
  transform_sample_counts(function(x) x/sum(x)) %>%
  phyloseq::distance(method="bray")
```

```{r}
agg.genus.bc.dist.mat <- agg.genus.bc.dist %>% as.matrix
diag(agg.genus.bc.dist.mat) <- NA
agg.genus.uniq.bc.df <- data.frame(
  uniqueness=apply(agg.genus.bc.dist.mat, MARGIN=2, FUN=function(x) {min(x, na.rm=T)}),
  stool.ID=sample_names(agg.physeq.genus))
```

# PCoA

```{r}
agg.pcoa.genus.bc <- agg.genus.bc.dist %>% 
  cmdscale(eig=T, k=3)

agg.pcoa.genus.bc.df <- data.frame(agg.pcoa.genus.bc$points) %>% 
  setNames(c("PCoA1", "PCoA2", "PCoA3")) %>% 
  rownames_to_column("stool.ID")
```

# Combine

```{r}
full.genus.clr.long.df <- rbind(
  
  # CLR
  genus.clr.long.df %>% 
    dplyr::rename(feature="taxon"),
  
  # alpha diversity
  alpha.div.long.df.genus %>% 
    dplyr::filter(metric %in% c("Chao1", "Simpson", "Shannon")) %>% 
    dplyr::rename(feature="metric") %>% 
    dplyr::select(stool.ID, feature, value),
  
  # PCoA
  agg.pcoa.genus.bc.df %>% 
    dplyr::select(stool.ID, PCoA1, PCoA2, PCoA3) %>% 
    pivot_longer(-stool.ID, names_to="feature"),
  
  # uniqueness
  agg.genus.uniq.bc.df %>% 
    dplyr::select(stool.ID, uniqueness) %>% 
    dplyr::rename(value="uniqueness") %>% 
    mutate(feature="uniq")
  
)
```

```{r}
dim(full.genus.clr.long.df)
n_distinct(full.genus.clr.long.df$feature)
n_distinct(full.genus.clr.long.df$stool.ID)
```

* 320k values
* 107 features
* 2997 stool IDs

# Export

## Features in rows

```{r}
genus.clr.df.feats.in.rows <- full.genus.clr.long.df %>% 
  pivot_wider(id_cols=feature, names_from="stool.ID")
dim(genus.clr.df.feats.in.rows)
```

```{r}
# genus.clr.df.feats.in.rows %>%
#   write.table(
#     sprintf("../results/kraken_genus_clr_filt_w_comm_n%ix%i_231017.txt", nrow(.), ncol(.)-1),
#     sep="\t", quote=F, row.names=F)
```

## Stool ID in rows

```{r}
genus.clr.df.stool.ID.in.rows <- full.genus.clr.long.df %>% 
  pivot_wider(id_cols=stool.ID, names_from="feature")
dim(genus.clr.df.stool.ID.in.rows)
```

```{r}
# genus.clr.df.stool.ID.in.rows %>%
#   write.table(
#     sprintf("../results/kraken_genus_clr_filt_w_comm_n%ix%i_231017.txt", nrow(.), ncol(.)-1),
#     sep="\t", quote=F, row.names=F)
```
