LMM separately per age versus longitudinally.

```{r, warning=F, message=F}
library(tidyverse)

PADJ.THRESH <- 0.01 
```

# Import cross-sectional LMM results

```{r}
herit.df.5mo <- read.table(
  "../results/split_by_age/asreml_kraken_genus_5mo/herit.txt",
  sep="\t", header=T) %>% 
  mutate(age = 5)
herit.df.10mo <- read.table(
  "../results/split_by_age/asreml_kraken_genus_10mo/herit.txt",
  sep="\t", header=T) %>% 
  mutate(age = 10)
herit.df.16mo <- read.table(
  "../results/split_by_age/asreml_kraken_genus_16mo/herit.txt",
  sep="\t", header=T) %>% 
  mutate(age = 16)
herit.df.22mo <- read.table(
  "../results/split_by_age/asreml_kraken_genus_5mo/herit.txt",
  sep="\t", header=T) %>% 
  mutate(age = 22)
herit.df.28mo <- read.table(
  "../results/split_by_age/asreml_kraken_genus_28mo/herit.txt",
  sep="\t", header=T) %>% 
  mutate(age = 28)

herit.df.xs <- rbind(
  herit.df.5mo,
  herit.df.10mo,
  herit.df.16mo,
  herit.df.22mo,
  herit.df.28mo)
```

# Import longitudinal LMM results

```{r}
herit.df.longitud <- read.table(
  "../results/asreml_kraken_genus_w_time_ranef/herit.txt",
  sep="\t", header=T)
```

# Import longitudinal, downsampled LMM results

```{r}
herit.df.longitud.downsamp <- read.table(
  "../results/asreml_kraken_genus_downsampled_w_time_ranef/herit.txt",
  sep="\t", header=T)
```

# Merge

```{r}
herit.df <- rbind(
  
  herit.df.xs %>%
    dplyr::select(feature, herit, LRT.pval, age),
  
  herit.df.longitud %>% 
    dplyr::select(feature, herit, LRT.pval) %>% 
    mutate(age = "longitudinal"),
  
  herit.df.longitud.downsamp %>% 
    dplyr::select(feature, herit, LRT.pval) %>% 
    mutate(age = "longitudinal_downsampled")
)
```

# Adjust p-values separately per age

```{r}
herit.df <- herit.df %>% 
  
  # exclude community features
  dplyr::filter(!(feature %in% c("PCoA1", "PCoA2", "PCoA3", "uniq", "Simpson", "Shannon", "Chao1"))) %>%
  
  group_by(age) %>% 
  mutate(LRT.padj = p.adjust(LRT.pval, method="BH")) %>% 
  ungroup() %>% 
  mutate(is.sig = LRT.padj < PADJ.THRESH)
```

```{r}
herit.df %>% 
  group_by(age) %>% 
  dplyr::count(is.sig) %>% 
  arrange(desc(is.sig))
```

These numbers are just genera, no community features.

# p-value histograms

Indicate # samples and # mice at each timepoint (added these numbers manually).

```{r}
n.per.age.df <- data.frame(
  n.samples = c(
    `5` = 563, `10` = 509, `16` = 641, `22` = 519, `28` = 367,
    longitudinal = 2974, longitudinal_downsampled = 516),
  
  n.mice = c(
    `5` = 563, `10` = 509, `16` = 641, `22` = 519, `28` = 367,
    longitudinal = 904, longitudinal_downsampled = 110))
```

```{r}
(longitud.v.cross.sectional.herit.pval.hists <- herit.df %>% 
   merge(n.per.age.df, by.x="age", by.y="row.names") %>% 
   mutate(age = factor(age, levels=c("5", "10", "16", "22", "28", "longitudinal", "longitudinal_downsampled"))) %>% 
   mutate(is.sig = factor(is.sig, levels=c(T, F))) %>% 
   ggplot(aes(LRT.pval, fill=is.sig)) +
   geom_histogram(bins=20) +
   facet_wrap(~age + n.mice + n.samples, labeller=labeller(
     n.mice = function(x){paste0("# mice = ", x)},
     n.samples = function(x){paste0("# samples = ", x)}), nrow=1) +
   labs(x="p-value", y="# genera") +
   scale_fill_manual(values=c(`TRUE`="royalblue", `FALSE`="gray80")) +
   scale_x_continuous(breaks=c(0, 0.5, 1)) +
   theme_classic(base_size=8) +
   theme(legend.position="none"))
```

```{r}
pdf("../plots/longitudinal_v_cross_sectional_heritability_genera_pval_histograms.pdf", height=2, width=7)
longitud.v.cross.sectional.herit.pval.hists
dev.off()
```

# % significantly heritable

Exclude community features.

```{r}
herit.df %>% 
  # dplyr::filter(is.sig) %>%
  dplyr::count(age, is.sig)
```

* Very few heritable taxa with cross-sectional analysis
