---
title: "Effect of chronological time"
---

What is the effect on our linear mixed models when we add chronological time (either as a fixed effect or random effect)? Is the model "improved"?

We will address this question in two ways.

First, we will just report how many features are associated with age when time is included as a fixed effect, as a random effect, or not at all.

Next, we will compare these 3 models against "cross-sectional" models, i.e. models run on samples collected at roughly the same time. In these models, chronological time is removed from the equation, so we consider age coefficients estimated from these models to be a kind of gold standard.

```{r, warning=F, message=F}
library(tidyverse)
library(patchwork) # plot_layout

PADJ.THRESH <- 0.01

age.color <- "seagreen"
theme_set(theme_classic(base_size=8))
```

# Import longitudinal results 

Full model with time as random effect, time as fixed effect, and without time.

```{r}
fixef.df.w.time.ranef <- read.table(
  "../results/asreml_kraken_genus_w_time_ranef/fixef.txt",
  sep="\t", header=T)

fixef.df.w.time.fixef <- read.table(
  "../results/asreml_kraken_genus_w_time_fixef/fixef.txt",
  sep="\t", header=T)

fixef.df.wo.time <- read.table(
  "../results/asreml_kraken_genus_wo_time/fixef.txt",
  sep="\t", header=T)
```

# Import cross-sectional results

Only perform cross-sectional models for genera, not pathways.

```{r}
fixef.df.slice1 <- read.table(
  "../results/cross_sectional_slices/asreml_kraken_genus_slice1/fixef.txt",
  sep="\t", header=T)

fixef.df.slice2 <- read.table(
  "../results/cross_sectional_slices/asreml_kraken_genus_slice2/fixef.txt",
  sep="\t", header=T)

fixef.df.slice3 <- read.table(
  "../results/cross_sectional_slices/asreml_kraken_genus_slice3/fixef.txt",
  sep="\t", header=T)
```

# Extract age coefficients and adjust p-values

```{r}
age.df.w.time.ranef <- fixef.df.w.time.ranef %>% 
  dplyr::filter(fixef == "age.wks.scaled") %>% 
  mutate(padj = p.adjust(Pr.chisq..con., method="BH")) %>% 
  mutate(is.sig = padj < PADJ.THRESH)

age.df.w.time.fixef <- fixef.df.w.time.fixef %>% 
  dplyr::filter(fixef == "age.wks.scaled") %>% 
  mutate(padj = p.adjust(Pr.chisq..con., method="BH")) %>% 
  mutate(is.sig = padj < PADJ.THRESH)

age.df.wo.time <- fixef.df.wo.time %>% 
  dplyr::filter(fixef == "age.wks.scaled") %>% 
  mutate(padj = p.adjust(Pr.chisq..con., method="BH")) %>% 
  mutate(is.sig = padj < PADJ.THRESH)

age.df.slice1 <- fixef.df.slice1 %>% 
  dplyr::filter(fixef == "age.wks.scaled") %>% 
  mutate(padj = p.adjust(Pr.chisq..con., method="BH")) %>% 
  mutate(is.sig = padj < PADJ.THRESH)

age.df.slice2 <- fixef.df.slice2 %>% 
  dplyr::filter(fixef == "age.wks.scaled") %>% 
  mutate(padj = p.adjust(Pr.chisq..con., method="BH")) %>% 
  mutate(is.sig = padj < PADJ.THRESH)

age.df.slice3 <- fixef.df.slice3 %>% 
  dplyr::filter(fixef == "age.wks.scaled") %>% 
  mutate(padj = p.adjust(Pr.chisq..con., method="BH")) %>% 
  mutate(is.sig = padj < PADJ.THRESH)
```

# 1) Number of age-associated features

```{r}
table(age.df.slice1$is.sig)
table(age.df.slice2$is.sig)
table(age.df.slice3$is.sig)
table(age.df.w.time.fixef$is.sig)
table(age.df.w.time.ranef$is.sig)
table(age.df.wo.time$is.sig)
```

* Longitudinal model without time has the most age-associated features
* Longitudinal model with time as ranef has fewer, but still a lot of, age-associated features
* Longitudinal model with time as fixef has few age-associated features
* All cross-sectional models have 5 age-associated features, and they're all different across the 3 models

```{r}
num.age.feats.df <- data.frame(
  model = factor(c("Fixed effect", "Random effect", "Omitted"), levels=c("Fixed effect", "Random effect", "Omitted")),
  n.age.feats = c(sum(age.df.w.time.fixef$is.sig), sum(age.df.w.time.ranef$is.sig), sum(age.df.wo.time$is.sig)))

(barplot.age.assoc.feats.w.or.wo.time <- num.age.feats.df %>% 
    ggplot(aes(x=model, y=n.age.feats)) +
    geom_col(fill=age.color) +
    labs(x="Time", y="# age-associated genera") +
    theme_classic(base_size=8))
```

~/Documents/phd/calico_aging/DRiDO_microbiome_github/analysis/effect_of_chronological_time.Rmd

```{r}
pdf("../plots/barplot_num_age_assoc_genera_w_time_fixef_ranef_omitted.pdf", height=2, width=1.5)
barplot.age.assoc.feats.w.or.wo.time
dev.off()
```

# 2) Correlations between models

```{r}
age.long.df <- rbind(
  age.df.slice1 %>% mutate(model="slice1"),
  age.df.slice2 %>% mutate(model="slice2"),
  age.df.slice3 %>% mutate(model="slice3"),
  age.df.w.time.ranef %>% mutate(model="full.w.time.ranef"),
  age.df.w.time.fixef %>% mutate(model="full.w.time.fixef"),
  age.df.wo.time %>% mutate(model="full.wo.time"))

age.wide.df <- age.long.df %>% 
  pivot_wider(id_cols = feature,
              values_from = solution,
              names_from = model)
```

```{r}
# Slice 1
(spearman.full.w.time.ranef.v.slice1 <- cor(age.wide.df$slice1, age.wide.df$full.w.time.ranef, method="spearman"))
(spearman.full.w.time.fixef.v.slice1 <- cor(age.wide.df$slice1, age.wide.df$full.w.time.fixef, method="spearman"))
(spearman.full.wo.time.v.slice1 <- cor(age.wide.df$slice1, age.wide.df$full.wo.time, method="spearman"))

spearman.pval.full.w.time.ranef.v.slice1 <- cor.test(age.wide.df$slice1, age.wide.df$full.w.time.ranef, method="spearman")$p.value
spearman.pval.full.w.time.fixef.v.slice1 <- cor.test(age.wide.df$slice1, age.wide.df$full.w.time.fixef, method="spearman")$p.value
spearman.pval.full.wo.time.v.slice1 <- cor.test(age.wide.df$slice1, age.wide.df$full.wo.time, method="spearman")$p.value

# Slice 2
(spearman.full.w.time.ranef.v.slice2 <- cor(age.wide.df$slice2, age.wide.df$full.w.time.ranef, method="spearman"))
(spearman.full.w.time.fixef.v.slice2 <- cor(age.wide.df$slice2, age.wide.df$full.w.time.fixef, method="spearman"))
(spearman.full.wo.time.v.slice2 <- cor(age.wide.df$slice2, age.wide.df$full.wo.time, method="spearman"))

spearman.pval.full.w.time.ranef.v.slice2 <- cor.test(age.wide.df$slice2, age.wide.df$full.w.time.ranef, method="spearman")$p.value
spearman.pval.full.w.time.fixef.v.slice2 <- cor.test(age.wide.df$slice2, age.wide.df$full.w.time.fixef, method="spearman")$p.value
spearman.pval.full.wo.time.v.slice2 <- cor.test(age.wide.df$slice2, age.wide.df$full.wo.time, method="spearman")$p.value

# Slice 3
(spearman.full.w.time.ranef.v.slice3 <- cor(age.wide.df$slice3, age.wide.df$full.w.time.ranef, method="spearman"))
(spearman.full.w.time.fixef.v.slice3 <- cor(age.wide.df$slice3, age.wide.df$full.w.time.fixef, method="spearman"))
(spearman.full.wo.time.v.slice3 <- cor(age.wide.df$slice3, age.wide.df$full.wo.time, method="spearman"))

spearman.pval.full.w.time.ranef.v.slice3 <- cor.test(age.wide.df$slice3, age.wide.df$full.w.time.ranef, method="spearman")$p.value
spearman.pval.full.w.time.fixef.v.slice3 <- cor.test(age.wide.df$slice3, age.wide.df$full.w.time.fixef, method="spearman")$p.value
spearman.pval.full.wo.time.v.slice3 <- cor.test(age.wide.df$slice3, age.wide.df$full.wo.time, method="spearman")$p.value
```

* Correlations are highest for the model with time as a fixed effect
* Correlation with the model that includes time as ranef is *slightly* better than for model without time

```{r}
# Slice 1 v. longitudinal w/ time ranef
full.w.time.ranef.v.slice1.scatter.p <- age.wide.df %>% 
  ggplot(aes(x=slice1, y=full.w.time.ranef)) +
  geom_abline(slope=1, lty=2) +
  geom_smooth(aes(label=NULL), method="lm") + 
  geom_point(shape=16) +
  labs(x="Slice 1", y="With time, ranef",
       title=sprintf(
         "rho = %.2f, p-value = %2.2e", 
         spearman.full.w.time.ranef.v.slice1, spearman.pval.full.w.time.ranef.v.slice1)) +
  theme_classic(base_size=8)

# Slice 2 v. longitudinal w/ time ranef
full.w.time.ranef.v.slice2.scatter.p <- age.wide.df %>% 
  ggplot(aes(x=slice2, y=full.w.time.ranef)) +
  geom_abline(slope=1, lty=2) +
  geom_smooth(aes(label=NULL), method="lm") + 
  geom_point(shape=16) +
  labs(x="Slice 2", y="With time, ranef",
       title=sprintf(
         "rho = %.2f, p-value = %2.2e", 
         spearman.full.w.time.ranef.v.slice2, spearman.pval.full.w.time.ranef.v.slice2)) +
  theme_classic(base_size=8)

# Slice 3 v. longitudinal w/ time ranef
full.w.time.ranef.v.slice3.scatter.p <- age.wide.df %>% 
  ggplot(aes(x=slice3, y=full.w.time.ranef)) +
  geom_abline(slope=1, lty=2) +
  geom_smooth(aes(label=NULL), method="lm") + 
  geom_point(shape=16) +
  labs(x="Slice 3", y="With time, ranef",
       title=sprintf(
         "rho = %.2f, p-value = %2.2e",
       spearman.full.w.time.ranef.v.slice3, spearman.pval.full.w.time.ranef.v.slice3)) +
  theme_classic(base_size=8)

# Slice 1 v. longitudinal w/ time fixef
full.w.time.fixef.v.slice1.scatter.p <- age.wide.df %>% 
  ggplot(aes(x=slice1, y=full.w.time.fixef)) +
  geom_abline(slope=1, lty=2) +
  geom_smooth(aes(label=NULL), method="lm") + 
  geom_point(shape=16) +
  labs(x="Slice 1", y="With time, fixef",
       title=sprintf(
         "rho = %.2f, p-value = %2.2e", 
         spearman.full.w.time.fixef.v.slice1, spearman.pval.full.w.time.fixef.v.slice1)) +
  theme_classic(base_size=8)

# Slice 2 v. longitudinal w/ time fixef
full.w.time.fixef.v.slice2.scatter.p <- age.wide.df %>% 
  ggplot(aes(x=slice2, y=full.w.time.fixef)) +
  geom_abline(slope=1, lty=2) +
  geom_smooth(aes(label=NULL), method="lm") + 
  geom_point(shape=16) +
  labs(x="Slice 2", y="With time, fixef",
       title=sprintf(
         "rho = %.2f, p-value = %2.2e", 
         spearman.full.w.time.fixef.v.slice2, spearman.pval.full.w.time.fixef.v.slice2)) +
  theme_classic(base_size=8)

# Slice 3 v. longitudinal w/ time fixef
full.w.time.fixef.v.slice3.scatter.p <- age.wide.df %>% 
  ggplot(aes(x=slice3, y=full.w.time.fixef)) +
  geom_abline(slope=1, lty=2) +
  geom_smooth(aes(label=NULL), method="lm") + 
  geom_point(shape=16) +
  labs(x="Slice 3", y="With time, fixef",
       title=sprintf(
         "rho = %.2f, p-value = %2.2e",
       spearman.full.w.time.fixef.v.slice3, spearman.pval.full.w.time.fixef.v.slice3)) +
  theme_classic(base_size=8)

# Slice 1 v. longitudinal w/o time
full.wo.time.v.slice1.scatter.p <- age.wide.df %>% 
  ggplot(aes(x=slice1, y=full.wo.time)) +
  geom_abline(slope=1, lty=2) +
  geom_smooth(aes(label=NULL), method="lm") + 
  geom_point(shape=16) +
  labs(x="Slice 1", y="Without time",
       title=sprintf(
         "rho = %.2f, p-value = %2.2e", 
         spearman.full.wo.time.v.slice1, spearman.pval.full.wo.time.v.slice1)) +
  theme_classic(base_size=8)

# Slice 2 v. longitudinal w/o time
full.wo.time.v.slice2.scatter.p <- age.wide.df %>% 
  ggplot(aes(x=slice2, y=full.wo.time)) +
  geom_abline(slope=1, lty=2) +
  geom_smooth(aes(label=NULL), method="lm") + 
  geom_point(shape=16) +
  labs(x="Slice 2", y="Without time",
       title=sprintf(
         "rho = %.2f, p-value = %2.2e", 
         spearman.full.wo.time.v.slice2, spearman.pval.full.wo.time.v.slice2)) +
  theme_classic(base_size=8)

# Slice 3 v. longitudinal w/o time
full.wo.time.v.slice3.scatter.p <- age.wide.df %>% 
  ggplot(aes(x=slice3, y=full.wo.time)) +
  geom_abline(slope=1, lty=2) +
  geom_smooth(aes(label=NULL), method="lm") + 
  geom_point(shape=16) +
  labs(x="Slice 3", y="Without time",
       title=sprintf(
         "rho = %.2f, p-value = %2.2e", 
         spearman.full.wo.time.v.slice3, spearman.pval.full.wo.time.v.slice3)) +
  theme_classic(base_size=8)
```

```{r, message=F}
min.x <- min(
  layer_scales(full.w.time.ranef.v.slice1.scatter.p, 1, 1)$x$range$range[1],
  layer_scales(full.w.time.ranef.v.slice2.scatter.p, 1, 1)$x$range$range[1],
  layer_scales(full.w.time.ranef.v.slice3.scatter.p, 1, 1)$x$range$range[1],
  layer_scales(full.w.time.fixef.v.slice1.scatter.p, 1, 1)$x$range$range[1],
  layer_scales(full.w.time.fixef.v.slice2.scatter.p, 1, 1)$x$range$range[1],
  layer_scales(full.w.time.fixef.v.slice3.scatter.p, 1, 1)$x$range$range[1],
  layer_scales(full.wo.time.v.slice1.scatter.p, 1, 1)$x$range$range[1],
  layer_scales(full.wo.time.v.slice2.scatter.p, 1, 1)$x$range$range[1],
  layer_scales(full.wo.time.v.slice3.scatter.p, 1, 1)$x$range$range[1])

max.x <- max(
  layer_scales(full.w.time.ranef.v.slice1.scatter.p, 1, 1)$x$range$range[2],
  layer_scales(full.w.time.ranef.v.slice2.scatter.p, 1, 1)$x$range$range[2],
  layer_scales(full.w.time.ranef.v.slice3.scatter.p, 1, 1)$x$range$range[2],
  layer_scales(full.w.time.fixef.v.slice1.scatter.p, 1, 1)$x$range$range[1],
  layer_scales(full.w.time.fixef.v.slice2.scatter.p, 1, 1)$x$range$range[1],
  layer_scales(full.w.time.fixef.v.slice3.scatter.p, 1, 1)$x$range$range[1],
  layer_scales(full.wo.time.v.slice1.scatter.p, 1, 1)$x$range$range[2],
  layer_scales(full.wo.time.v.slice2.scatter.p, 1, 1)$x$range$range[2],
  layer_scales(full.wo.time.v.slice3.scatter.p, 1, 1)$x$range$range[2])

min.y <- min(
  layer_scales(full.w.time.ranef.v.slice1.scatter.p, 1, 1)$y$range$range[1],
  layer_scales(full.w.time.ranef.v.slice2.scatter.p, 1, 1)$y$range$range[1],
  layer_scales(full.w.time.ranef.v.slice3.scatter.p, 1, 1)$y$range$range[1],
  layer_scales(full.w.time.fixef.v.slice1.scatter.p, 1, 1)$y$range$range[1],
  layer_scales(full.w.time.fixef.v.slice2.scatter.p, 1, 1)$y$range$range[1],
  layer_scales(full.w.time.fixef.v.slice3.scatter.p, 1, 1)$y$range$range[1],
  layer_scales(full.wo.time.v.slice1.scatter.p, 1, 1)$y$range$range[1],
  layer_scales(full.wo.time.v.slice2.scatter.p, 1, 1)$y$range$range[1],
  layer_scales(full.wo.time.v.slice3.scatter.p, 1, 1)$y$range$range[1])

max.y <- max(
  layer_scales(full.w.time.ranef.v.slice1.scatter.p, 1, 1)$y$range$range[2],
  layer_scales(full.w.time.ranef.v.slice2.scatter.p, 1, 1)$y$range$range[2],
  layer_scales(full.w.time.ranef.v.slice3.scatter.p, 1, 1)$y$range$range[2],
  layer_scales(full.w.time.fixef.v.slice1.scatter.p, 1, 1)$y$range$range[2],
  layer_scales(full.w.time.fixef.v.slice2.scatter.p, 1, 1)$y$range$range[2],
  layer_scales(full.w.time.fixef.v.slice3.scatter.p, 1, 1)$y$range$range[2],
  layer_scales(full.wo.time.v.slice1.scatter.p, 1, 1)$y$range$range[2],
  layer_scales(full.wo.time.v.slice2.scatter.p, 1, 1)$y$range$range[2],
  layer_scales(full.wo.time.v.slice3.scatter.p, 1, 1)$y$range$range[2])

cor.cross.sectional.v.longitudinal.age.coef.manual.p <- (
  full.w.time.fixef.v.slice1.scatter.p + ylim(min.y, max.y) + xlim(min.x, max.x)) +
  (full.w.time.fixef.v.slice2.scatter.p + ylim(min.y, max.y) + xlim(min.x, max.x)) +
  (full.w.time.fixef.v.slice3.scatter.p + ylim(min.y, max.y) + xlim(min.x, max.x)) +
  (full.w.time.ranef.v.slice1.scatter.p + ylim(min.y, max.y) + xlim(min.x, max.x)) +
  (full.w.time.ranef.v.slice2.scatter.p + ylim(min.y, max.y) + xlim(min.x, max.x)) +
  (full.w.time.ranef.v.slice3.scatter.p + ylim(min.y, max.y) + xlim(min.x, max.x)) +
  (full.wo.time.v.slice1.scatter.p + ylim(min.y, max.y) + xlim(min.x, max.x)) +
  (full.wo.time.v.slice2.scatter.p + ylim(min.y, max.y) + xlim(min.x, max.x)) +
  (full.wo.time.v.slice3.scatter.p + ylim(min.y, max.y) + xlim(min.x, max.x)) +
  plot_layout(nrow=3)
```

```{r}
pdf("../plots/age_coef_cross_sectional_v_longitudinal_scatter.pdf", height=6, width=6)
cor.cross.sectional.v.longitudinal.age.coef.manual.p
dev.off()
```
