We have three timepoints:

1. Baseline: two weeks after germ-free mice received a young or old FMT and were housed in isocages
2. 2 weeks cohousing: two weeks of Y FMT recipients being cohoused with O FMT recipients
3. 1 month cohousing

# Load libraries

```{r, warning=F, message=F}
library(tidyverse)
library(phyloseq)
library(speedyseq) # faster tax_glom

separation.palette <- c(
  Y = "skyblue",
  O = "firebrick",
  CY = "mediumpurple",
  CO = "violetred"
)
```

# Import metadata

```{r}
meta.df <- read.table(
  "../data/metadata/germfree_cohousing_metadata.txt", sep="\t", header=T)
head(meta.df)
```

```{r}
meta.df <- meta.df %>% 
  
  # convert timepoint.short to a factor
  mutate(timepoint.short = factor(
    timepoint.short, levels=c("B", "C2", "C4"))) %>% 
  
  # create metadata that does not know about future treatment status
  # (e.g. at baseline, CY mice are just Y)
  mutate(current.group = case_when(
    timepoint.short == "B" & group == "CY" ~ "Y",
    timepoint.short == "B" & group == "CO" ~ "O",
    TRUE ~ group)) %>% 
  mutate(current.group = factor(current.group, levels=c(
      "Y", "O", "CY", "CO")))
```

# Import taxonomy

```{r}
tax.df <- read.table("../data/germfree_cohousing_taxonomy.tsv", sep="\t", header=T)
```

# Import counts

```{r}
counts.df <- read.table(
  "../data/germfree_cohousing_feature_table.tsv", 
  sep="\t", header=T, row.names=1)
dim(counts.df)
```

* 5041 taxa x 131 samples

# Create phyloseq object

Arrange samples in the same order in counts and metadata.

```{r}
# # confirm that all samples from metadata are also in counts, and vice versa
# meta.df$sample.id %in% colnames(counts.df)
# colnames(counts.df) %in% meta.df$sample.id
```

```{r}
physeq <- phyloseq(
  counts.df %>% as.matrix() %>% otu_table(taxa_are_rows=T),
  meta.df %>% column_to_rownames("sample.id") %>% sample_data(),
  tax.df %>% column_to_rownames("Feature.ID") %>% as.matrix() %>% tax_table()
)
```

# Aggregate to genera

```{r}
physeq.genus <- physeq %>% tax_glom(taxrank="genus")
ntaxa(physeq.genus)
sum(otu_table(physeq.genus)) / sum(otu_table(physeq))
```

* 191 genera
* 92.6% of counts could be assigned to a genus

# PCoA

Bray-Curtis distance on genus-level relabs.

```{r}
pcoa.res <- physeq.genus %>%
  
  # compute relative abundance
  transform_sample_counts(function(x) 100*x/sum(x)) %>%
  
  # perform PCoA using Bray-Curtis distance
  ordinate(method="MDS", distance="bray")

pcoa.genus.df <- plot_ordination(physeq.genus, pcoa.res, type="samples", justDF=T)
```

Calculate centroids.

```{r}
pcoa.centroids.df <- pcoa.genus.df %>% 
  group_by(current.group, timepoint.short) %>% 
  summarise(Axis.1 = mean(Axis.1),
            Axis.2 = mean(Axis.2),
            .groups="drop")
```

Add lines between centroids for Y to CY and for O to CO.

```{r}
(pcoa.lines.bw.centroids.df <- pcoa.centroids.df %>% 
   dplyr::filter(str_detect(timepoint.short, "^C")) %>% 
   group_by(timepoint.short) %>% 
   summarise(Axis.1_Y = Axis.1[current.group == "Y"],
             Axis.1.end_Y = Axis.1[current.group == "CY"],
             Axis.1_O = Axis.1[current.group == "O"],
             Axis.1.end_O = Axis.1[current.group == "CO"],
             Axis.2_Y = Axis.2[current.group == "Y"],
             Axis.2.end_Y = Axis.2[current.group == "CY"],
             Axis.2_O = Axis.2[current.group == "O"],
             Axis.2.end_O = Axis.2[current.group == "CO"]) %>% 
   pivot_longer(starts_with("Axis"), names_to=c("Axis", "age"), names_sep="_") %>% 
   pivot_wider(id_cols=c(timepoint.short, age), names_from="Axis"))
```

```{r}
(germfree.cohousing.pcoa.p <- pcoa.genus.df %>% 
   ggplot(aes(x=Axis.1, y=Axis.2, color=current.group)) +
   geom_point(shape=16, alpha=0.3) +
   geom_point(data=pcoa.centroids.df, size=5, shape="+") +
   facet_wrap(~timepoint.short, ncol=1) +
   theme_classic(base_size=8) +
   labs(x=sprintf("PCoA1 (%.1f%%)", 100*pcoa.res$values$Relative_eig[1]), 
        y=sprintf("PCoA2 (%.1f%%)", 100*pcoa.res$values$Relative_eig[2]),
        color="Group") +
   scale_color_manual(values=separation.palette))
```

```{r}
pdf("../plots/germfree_cohousing_pcoa.pdf", width=2, height=4.5)
germfree.cohousing.pcoa.p + theme(legend.position="none")
dev.off()
```

