---
title: "Analysis of pathway results from HUMAnN"
---

# Load libraries

```{r, warning=F, message=F}
library(tidyverse)
library(foreach)
library(patchwork) # for mega PCoA
library(ggpubr) # for t-tests
library(ggrepel) # for labels
library(corrr) # stretch, DR scatterplots
library(GGally) # ggpairs, DR scatterplots

PADJ.THRESH <- 0.01

age.color <- "seagreen"
genetics.color <- "gold"
diet.color <- "mediumorchid"

diet.palette <- c(AL="seashell4",
                  `1D`="skyblue",
                  `2D`="royalblue4",
                  `20`="orange",
                  `40`="firebrick")
CR.v.fasting.palette <- c(fasting="royalblue4", CR="firebrick")
pathway.type.palette <- c(`community-wide`="royalblue", `specialized`="firebrick")
```

# Import metadata

```{r}
stool.meta.df <- read.table(
  "../data/metadata/stool_metadata_after_QC_no_controls_n2997_240418.txt", 
  sep="\t", header=T)
mouse.meta.df <- read.csv("../data/metadata/AnimalData_Processed_20230712.csv")
stool.meta.annot.df <- stool.meta.df %>%
  merge(mouse.meta.df, by.x="mouse.ID", by.y="MouseID")

# identify pre-randomization samples as AL
stool.meta.annot.df <- stool.meta.annot.df %>%
  mutate(Diet.5mo.as.AL=factor(case_when(
    age.wks < 25 ~ "AL",
    TRUE ~ as.character(Diet)),
    levels=c("AL", "1D", "2D", "20", "40")))
```


# Import pathway data

```{r}
pathway.tpm.df <- read.table(
  "../data/pathabundance_tpm_agg_by_stool_ID_n422x2997.txt",
  sep="\t", header=T, quote="", row.names=1)

dim(pathway.tpm.df)
```

## log2(TPM)

```{r}
zero.replacement.per.pathway <- apply(pathway.tpm.df, MARGIN=1, FUN=function(x) {min(x[x != 0]) / 2})
```

```{r}
pathway.tpm.mat.no.zeros <- foreach(this.pathway = rownames(pathway.tpm.df), .combine="rbind") %do% {
  
  this.pathway.vec <- as.numeric(pathway.tpm.df[this.pathway, ])
  this.pathway.vec[this.pathway.vec == 0] <- zero.replacement.per.pathway[[this.pathway]]
  
  return(this.pathway.vec)
  
}

pathway.tpm.df.no.zeros <- data.frame(pathway.tpm.mat.no.zeros)
rownames(pathway.tpm.df.no.zeros) <- rownames(pathway.tpm.df)
colnames(pathway.tpm.df.no.zeros) <- colnames(pathway.tpm.df)
```

```{r}
pathway.log2tpm.df <- log2(pathway.tpm.df.no.zeros)
dim(pathway.log2tpm.df)
```

## Import short and long pathway names

```{r}
pathway.names.df <- read.table(
  "../data/metadata/pathway_long_and_short_names_n703.txt",
  sep="\t", header=T, quote="")
```

Add additional columns for plotting.

```{r}
pathway.names.df <- pathway.names.df %>% 
  mutate(after.colon = str_sub(str_extract(original, "(?=:)(.*)"), 3)) %>% 
  mutate(after.colon.short = ifelse(
    nchar(after.colon) > 30, 
    paste0(str_sub(after.colon, 1, 27), "..."),
    after.colon)) %>% 
  mutate(after.colon.short.w.ID = paste0(
    after.colon.short, " (", short, ")")) %>% 
  relocate(after.colon.short.w.ID)
```

We need `short.clean` and `after.colon.short.w.ID`.

# Import ASReml results

## Age and DR as fixed effects

Model: y_mb ~ age (f) + diet (f) + [genetics] (r) + mouse (r) + batch (r) + cohort (r) + cage (r)

```{r}
pathway.fixef.df <- read.table(
  "../results/asreml_humann_pathways/fixef.txt",
  sep="\t", header=T)

pathway.ranef.df <- read.table(
  "../results/asreml_humann_pathways/ranef.txt",
  sep="\t", header=T)

pathway.herit.df <- read.table(
  "../results/asreml_humann_pathways/herit.txt",
  sep="\t", header=T)
```

## All random effects

Model: y_mb ~ age (r) + diet (r) + [genetics] (r) + mouse (r) + batch (r) + cohort (r) + cage (r)

```{r}
pathway.ranef.df.all.ranef.model <- read.table(
  "../results/asreml_humann_pathways_all_ranef/ranef.txt",
  sep="\t", header=T)
```

# Import taxonomic data

To be able to determine which pathways are "community-wide" versus "specialized".

```{r}
genus.df <- read.table(
  "../results/kraken_genus_clr_filt_w_comm_n107x2997_231017.txt",
  sep="\t", header=T, row.names=1)
dim(genus.df)
```

# Define "community-wide" and "specialized" pathways

First, correlate pathways with genera. log2(TPM) for pathways, CLR for genera.

```{r}
# identical(
#   pathway.tpm.df %>% t() %>% data.frame() %>% rownames(),
#   genus.df %>% t() %>% data.frame() %>% rownames())

genus.v.pathway.cor.df <- cor(
  genus.df %>% t() %>% data.frame() %>% dplyr::select(-c(Chao1, Shannon, Simpson, PCoA1, PCoA2, PCoA3, uniq)),
  pathway.log2tpm.df %>% t() %>% data.frame(check.names=F)) %>% data.frame(check.names=F)

max.cor.per.pathway <- apply(genus.v.pathway.cor.df, MARGIN=2, FUN=max)
closest.genus.per.pathway <- apply(
  genus.v.pathway.cor.df, MARGIN=2, FUN=function(x) {names(x)[which.max(x)]})
```

**I will define a specialized pathway as having 1-4 correlations over 0.5**.

```{r}
genus.v.pathway.cor.long.df <- genus.v.pathway.cor.df %>% 
  rownames_to_column("genus") %>% 
  pivot_longer(-genus, names_to="pathway", values_to="cor") %>% 
  group_by(pathway) %>% 
  mutate(max.genus.cor = max(cor),
         closest.genus = genus[which.max(cor)],
         n.gt.0p5 = sum(cor > 0.5)) %>% 
  mutate(pathway.type = ifelse(n.gt.0p5 %in% c(1, 2, 3, 4), "specialized", "community-wide")) %>% 
  ungroup()
```

```{r}
pathway.type.df <- genus.v.pathway.cor.long.df %>% 
  distinct(pathway, pathway.type, max.genus.cor, closest.genus)
```

Also indicate if this pathway passed prevalence filtration.

```{r}
pathway.type.df <- pathway.type.df %>% 
  dplyr::rename(original = pathway) %>% 
  merge(pathway.names.df, by="original") %>% 
  mutate(passed.prev.filt = short.clean %in% unique(pathway.fixef.df$feature))
```

```{r}
table(pathway.type.df$passed.prev.filt)
table(pathway.type.df$pathway.type)
pathway.type.df %>% dplyr::filter(passed.prev.filt == T) %>% pull(pathway.type) %>% table()
```

* 272 of 422 pathways passed prevalence filtering
* 213 of all 422 pathways are community-wide
* 140 of 272 pathways that passed prevalence filtering are community-wide

## Plot two examples

```{r}
one.comm.and.one.spec.path <- c(
  
  # community-wide
  "BRANCHED-CHAIN-AA-SYN-PWY: superpathway of branched chain amino acid biosynthesis",
  
  # specialized
  "PWY-8004: Entner-Doudoroff pathway I"
)

(example.comm.v.specialized.histograms <- genus.v.pathway.cor.long.df %>% 
    dplyr::filter(pathway %in% one.comm.and.one.spec.path) %>% 
    ggplot(aes(cor, fill=pathway.type)) +
    geom_histogram() +
    geom_vline(xintercept=0.5, lty=2) +
    facet_wrap(~pathway.type + pathway, ncol=1, scales="free_y") +
    labs(x="Correlation with a genus", y="# correlations") +
    scale_fill_manual(values=pathway.type.palette) +
    theme_classic(base_size=8) +
    theme(legend.position="none"))
```

```{r}
# pdf("../plots/community_wide_v_specialized_pathway_example_histograms.pdf", height=3, width=1.5)
# example.comm.v.specialized.histograms
# dev.off()
```

# Uniqueness

```{r}
pathway.dist.mat <- pathway.log2tpm.df %>% 
  t() %>% dist() %>% as.matrix()
diag(pathway.dist.mat) <- NA
pathway.uniq.df <- data.frame(
  uniqueness=apply(pathway.dist.mat, MARGIN=2, FUN=function(x) {min(x, na.rm=T)})) %>% 
  rownames_to_column("stool.ID") %>% 
  merge(stool.meta.annot.df, by="stool.ID")
```

```{r}
(pathway.uniq.v.age.p <- pathway.uniq.df %>% 
   dplyr::filter(age.approx.months <= 40) %>%
   ggplot(aes(x=age.approx.months, y=uniqueness)) +
   geom_vline(xintercept=6, lty=2) +
   geom_smooth(method="lm", color=age.color, linewidth=0.5) +
   geom_boxplot(aes(group=age.approx.months), fill=age.color, 
                outlier.shape=16, outlier.size=0.5) +
   labs(x="Age (months)", y="Uniqueness", 
        title="Functional uniqueness") +
   theme_classic(base_size=8) +
   theme(legend.position="none")
)
```

```{r}
# pdf("../plots/fxn_uniq_v_age_boxplots.pdf", width=2, height=2)
# pathway.uniq.v.age.p
# dev.off()
```

# PCA

```{r}
euclid.dist.pathway.log2tpm <- pathway.log2tpm.df %>% 
  as.matrix() %>% 
  t() %>% 
  dist()

pca.pathway.log2tpm <- euclid.dist.pathway.log2tpm %>% 
  cmdscale(eig=T, k=3)

pca.pathway.log2tpm.df <- merge(
  data.frame(pca.pathway.log2tpm$points) %>% 
    setNames(c("PC1", "PC2", "PC3")) %>% 
    rownames_to_column("stool.ID"),
  stool.meta.annot.df,
  by="stool.ID") %>% 
  mutate(Diet=factor(Diet, levels=c("AL", "1D", "2D", "20", "40")),
         Diet.5mo.as.AL=factor(Diet.5mo.as.AL, levels=c("AL", "1D", "2D", "20", "40")))
```

```{r}
frac.var.explained.by.pcs <- pca.pathway.log2tpm$eig / sum(pca.pathway.log2tpm$eig)
frac.var.explained.by.pcs[1:3]
```

```{r}
(pathway.pca.p <- pca.pathway.log2tpm.df %>% 
   arrange(desc(age.approx.months)) %>% {
     ggplot(., aes(x=PC1, y=PC2, color=Diet.5mo.as.AL, size=age.approx.months)) +
       geom_point(shape=16, alpha=0.8) +
       scale_color_manual(values=diet.palette) +
       scale_size_continuous(breaks=c(5,10,16,22,28,34,40,46), range=c(0.25,2))
   })
```

## With barplots

```{r}
pc1.lims <- layer_scales(pathway.pca.p, 1, 1)$x$range$range
pc1.intervals <- seq(from=pc1.lims[1], to=pc1.lims[2], length.out=20)

pc1.age.barplots <- pca.pathway.log2tpm.df %>% 
  mutate(PC1.discrete=cut(PC1, pc1.intervals, include.lowest=T)) %>%
  group_by(PC1.discrete) %>%
  summarise(mean.age.mo=mean(age.approx.months), .groups="drop") %>%
  ggplot(aes(x=PC1.discrete, y=mean.age.mo)) +
  geom_col() +
  theme_classic(base_size=8) +
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
```

```{r}
pc2.lims <- layer_scales(pathway.pca.p, 1, 1)$y$range$range
pc2.intervals <- seq(from=pc2.lims[1], to=pc2.lims[2], length.out=20)

pc2.age.barplots <- pca.pathway.log2tpm.df %>% 
  mutate(PC2.discrete=cut(PC2, pc2.intervals, include.lowest=T)) %>%
  group_by(PC2.discrete) %>%
  summarise(mean.age.mo=mean(age.approx.months), .groups="drop") %>%
  ggplot(aes(x=mean.age.mo, y=PC2.discrete)) +
  geom_col() +
  theme_classic(base_size=8) +
  theme(axis.text.y=element_blank(),
        axis.ticks.y=element_blank())
```

```{r}
pc1.diet.boxplots <- pca.pathway.log2tpm.df %>% 
  ggplot(aes(x=PC1, y=fct_rev(Diet.5mo.as.AL), fill=Diet.5mo.as.AL)) +
  geom_boxplot(outlier.shape=16, outlier.size=0.5) +
  scale_fill_manual(values=diet.palette)
```

```{r}
pc2.diet.boxplots <- pca.pathway.log2tpm.df %>% 
  ggplot(aes(x=Diet.5mo.as.AL, y=PC2, fill=Diet.5mo.as.AL)) +
  geom_boxplot(outlier.shape=16, outlier.size=0.5) +
  scale_fill_manual(values=diet.palette)
```

```{r}
mega.pca.layout <- "
#AAAAAA#
BCCCCCCD
BCCCCCCD
BCCCCCCD
BCCCCCCD
BCCCCCCD
BCCCCCCD
#EEEEEE#
"

(mega.pca <- (
  
  # A = PC1 diet boxplots
  pc1.diet.boxplots + 
    scale_x_continuous(position="top") + 
    labs(x=sprintf("PC1 (%.0f%%)", 100*frac.var.explained.by.pcs[1])) +
    theme_classic(base_size=8) +
    theme(legend.position="none", 
          axis.title.y=element_blank()) +
    
    # B = PC2 diet boxplots
    pc2.diet.boxplots + 
    labs(y=sprintf("PC2 (%.0f%%)", 100*frac.var.explained.by.pcs[2])) +
    theme_classic(base_size=8) +
    theme(legend.position="none", 
          axis.title.x=element_blank()) +
    
    
    # C = main PCA
    pathway.pca.p + 
    theme_classic(base_size=8) +
    theme(legend.position="none",
          axis.title.x=element_blank(),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank(),
          axis.title.y=element_blank(),
          axis.text.y=element_blank(),
          axis.ticks.y=element_blank()) +
    
    # D = PC2 age barplots
    pc2.age.barplots + 
    scale_y_discrete(position="right") +
    scale_x_continuous(position="top") +
    labs(x="Mean age (mo)", y="Binned PC2") +
    
    # E = PC1 age barplots
    pc1.age.barplots + 
    scale_y_reverse(position="right") +
    labs(x="Binned PC1", y="Mean age (mo)")) +
    
    plot_layout(design=mega.pca.layout))
```

```{r}
# pdf("../plots/humann_pathway_pca_w_barplots.pdf", height=4, width=4)
# mega.pca
# dev.off()
```

# Effect of age

```{r}
pathway.age.df <- pathway.fixef.df %>% 
  dplyr::filter(fixef == "age.wks.scaled") %>% 
  merge(pathway.names.df, by.x="feature", by.y="short.clean", all.x=T) # all.x=T for uniqueness
```

Adjust p-values.

```{r}
pathway.age.df <- pathway.age.df %>% 
  mutate(padj = p.adjust(Pr.chisq..con., method="BH")) %>% 
  mutate(is.sig = padj < PADJ.THRESH)
```

## Community-wide versus specialized

```{r}
pathway.age.df.w.max.genus.cor <- pathway.age.df %>% 
  merge(pathway.type.df, by.x="feature", by.y="short.clean")
```

```{r}
age.assoc.pathways <- c(
  
  # specialized, increases with age
  "PWY-622: starch biosynthesis",
  
  # community-wide, decreases with age
  "PWY-1042: glycolysis IV"
  
)

(age.assoc.comm.v.specialized.histograms <- genus.v.pathway.cor.long.df %>% 
    dplyr::filter(pathway %in% age.assoc.pathways) %>% 
    ggplot(aes(cor, fill=pathway.type)) +
    geom_histogram() +
    geom_vline(xintercept=0.5, lty=2) +
    facet_wrap(~pathway.type + pathway, ncol=1, scales="free_y") +
    labs(x="Correlation with a genus", y="# correlations") +
    scale_fill_manual(values=pathway.type.palette) +
    theme_classic(base_size=8) +
    theme(legend.position="none"))
```

```{r}
# pdf("../plots/community_wide_v_specialized_pathway_histograms_age_assoc.pdf", height=3, width=1.5)
# age.assoc.comm.v.specialized.histograms
# dev.off()
```

## Barplot

```{r}
age.assoc.pathways.short <- c(
  
  "Uniqueness",
  
  # specialized, increases with age
  "PWY622", 
  
  # community-wide, decreases with age
  "PWY1042"
  
)
```

```{r}
(age.effect.pathways.p <- pathway.age.df %>% 
   mutate(label = case_when(
     feature == "Uniqueness" ~ "Uniqueness",
     feature %in% age.assoc.pathways.short ~ after.colon.short.w.ID,
     TRUE ~ ""
   )) %>%
   ggplot(aes(y=reorder(feature, solution), x=solution, 
              xmin=solution-std.error, xmax=solution+std.error,
              label=label)) +
   geom_col(aes(fill=is.sig)) +
   geom_linerange(linewidth=0.25) +
   geom_vline(xintercept=0, lty=2) +
   geom_text_repel(max.overlaps=Inf, nudge_x=0.01, min.segment.length=0, size=0.36*6) +
   labs(y="", x="Age effect", 
        color=sprintf("padj < %.2f?", PADJ.THRESH)) +
   theme_void(base_size=8) +
   theme(axis.text.y=element_blank(),
         legend.position="none") + 
   scale_fill_manual(values=c(`TRUE`=age.color, `FALSE`="gray80"))
)
```

```{r}
# pdf("../plots/asreml_age_effect_pathway_barplot.pdf", height=3.5, width=2)
# age.effect.pathways.p
# dev.off()
```

## Examples

```{r}
(glycolysis4.v.age.boxplots.p <- pathway.log2tpm.df["PWY-1042: glycolysis IV", ] %>% 
   t() %>% 
   data.frame() %>% 
   setNames("log2tpm") %>% 
   rownames_to_column("stool.ID") %>% 
   merge(stool.meta.annot.df, by="stool.ID") %>% 
   ggplot(aes(x=age.approx.months, y=log2tpm)) +
   geom_smooth(method="lm", color=age.color, linewidth=0.5) +
   geom_boxplot(aes(group=age.approx.months), fill=age.color, 
                outlier.shape=16, outlier.size=0.5) +
   geom_vline(xintercept=6, lty=2) +
   labs(x="Age (months)", y="log2(TPM)", title="glycolysis IV (PWY-1042)") +
   theme_classic(base_size=8))
```

```{r}
# pdf("../plots/glycolysis_IV_v_age_boxplots.pdf", height=2, width=2)
# glycolysis4.v.age.boxplots.p
# dev.off()
```

# Effect of diet

```{r}
pathway.DR.df <- pathway.fixef.df %>% 
  dplyr::filter(fixef == "Diet.5mo.as.AL") %>% 
  dplyr::filter(fixef.value != "AL") %>% # the AL row doesn't add anything
  merge(pathway.names.df, by.x="feature", by.y="short.clean", all.x=T) # all.x=T for uniqueness

dim(pathway.DR.df)
n_distinct(pathway.DR.df$feature)
```

Adjust p-values. I think I could call `p.adjust` directly even though each p-value is repeated 4x, but to be safe, I'll compute adjusted p-values for just one diet and then merge them back in.

```{r}
DR.padj.df <- pathway.DR.df %>% 
  dplyr::filter(fixef.value == "40") %>% 
  mutate(padj = p.adjust(Pr.chisq..con., method="BH")) %>% 
  dplyr::select(feature, padj)

pathway.DR.df <- pathway.DR.df %>% 
  merge(DR.padj.df, by="feature") %>% 
  mutate(is.sig = padj < PADJ.THRESH)

dim(pathway.DR.df)
```

## Community-wide versus specialized

```{r}
DR.pathways <- c(
  
  "OANTIGEN-PWY: O-antigen building blocks biosynthesis (E. coli)",
  "PWY-2941: L-lysine biosynthesis II",
  "CITRULBIO-PWY: L-citrulline biosynthesis",
  "PWY-4984: urea cycle"

  )
```


```{r}
(DR.comm.v.specialized.histograms <- genus.v.pathway.cor.long.df %>%
   dplyr::filter(pathway %in% DR.pathways) %>% 
   ggplot(aes(cor, fill=pathway.type)) +
   geom_histogram() +
   geom_vline(xintercept=0.5, lty=2) +
   facet_wrap(~pathway.type + pathway, nrow=1, scales="free_y") +
   labs(x="Correlation with a genus", y="# correlations") +
   scale_fill_manual(values=pathway.type.palette) +
   scale_x_continuous(breaks=c(-0.5, 0, 0.5)) +
   theme_classic(base_size=8) +
   theme(legend.position="none"))
```

```{r}
# pdf("../plots/community_wide_v_specialized_pathway_histograms_DR.pdf", width=5, height=1.5)
# DR.comm.v.specialized.histograms
# dev.off()
```

## Barplots

Order based on 40% CR.

```{r}
pathway.DR.effects.feature.order.df <- pathway.DR.df %>% 
  dplyr::filter(fixef.value == "40") %>%
  arrange(solution) %>% 
  mutate(feature.order = row_number()) %>% 
  dplyr::select(feature, feature.order)
```

```{r}
DR.pathways.short <- c(
  
  "OANTIGENPWY",
  "PWY2941",
  "CITRULBIOPWY",
  "PWY4984"
  )

DR.assoc.pathways.to.highlight.ii <- pathway.DR.effects.feature.order.df %>% 
  dplyr::filter(feature %in% DR.pathways.short) %>% 
  pull(feature.order)
```


```{r}
(DR.effect.pathways.p <- pathway.DR.df %>% 
   
   # specify order based on 40% CR
   merge(pathway.DR.effects.feature.order.df, by="feature") %>%
   
   # reorder diet
   mutate(DR = factor(fixef.value, levels=c("1D", "2D", "20", "40"))) %>% 
   
   ggplot(aes(y=reorder(feature, feature.order), x=solution, 
              xmin=solution-std.error, xmax=solution+std.error,
              alpha=is.sig, fill=DR)) +
   geom_col() + 
   geom_linerange(linewidth=0.25) +
   geom_vline(xintercept=0, lty=2) +
   geom_hline(yintercept=DR.assoc.pathways.to.highlight.ii, lty=2, color="gray50") +
   facet_wrap(~DR, nrow=1) +
   theme_void(base_size=8) +
   theme(legend.position="none") + 
   scale_fill_manual(values=diet.palette) +
   scale_alpha_manual(values=c(`TRUE`=1, `FALSE`=0.2))
)
```

```{r}
# pdf("../plots/asreml_DR_effect_pathway_barplot.pdf", height=3, width=3)
# DR.effect.pathways.p
# dev.off()
```

## Magnitude of effects

```{r}
(pathway.DR.mean.abs.effect.df <- pathway.DR.df %>% 
   dplyr::rename(DR = fixef.value) %>% 
   group_by(DR) %>% 
   summarise(mean = mean(abs(solution)),
             median = median(abs(solution)), .groups="drop") %>% 
   mutate(solution = sprintf("%.2f", mean)))
```

* 40% > 2D ≥ 20% > 1D

```{r}
pd <- position_dodge2(width=0.1)

(diet.effect.size.p <- pathway.DR.df %>% 
    mutate(DR=factor(fixef.value, levels=c("1D", "2D", "20", "40"))) %>% 
    ggplot(aes(x=DR, y=abs(solution))) +
    geom_path(aes(group=feature), alpha=0.1, position=pd) +
    geom_point(aes(color=DR), shape=16, position=pd) +
    geom_text(aes(y=-0.1, label=solution, color=DR), data=pathway.DR.mean.abs.effect.df) +
    stat_summary(aes(color=DR), fun="mean", geom="crossbar", width=0.5) +
    scale_color_manual(values=diet.palette) +
    stat_compare_means(method="t.test", label="p.signif", paired=T, comparisons=list(
      c("1D", "2D"), c("1D", "20"), c("1D", "40"), 
      c("2D", "20"), c("2D", "40"), c("20", "40"))) +
    labs(y="Absolute DR coefficient") +
    theme_classic(base_size=8) +
    theme(legend.position = "none"))
```

```{r}
# pdf("../plots/asreml_diet_overall_effect_pathway_boxplots.pdf", height=2, width=2)
# diet.effect.size.p
# dev.off()
```

## Similarity of diet coefficients

It will be useful to first pivot wider.

```{r}
DR.coef.df <- pathway.DR.df %>%
  pivot_wider(id_cols=feature, names_from=fixef.value, values_from=solution)
```

```{r}
DR.cor.df <- DR.coef.df %>% 
  dplyr::select(-feature) %>% 
  correlate() %>% 
  shave() %>%
  stretch(na.rm=T) %>% 
  mutate(x = factor(x, levels=c("1D", "2D", "20", "40")),
         y = factor(y, levels=c("1D", "2D", "20", "40")),
         r.chr = sprintf("%.2f", r)) 
```

```{r}
# make sure we use linear fit
my_fn <- function(data, mapping, ...){
  p <- ggplot(data = data, mapping = mapping) + 
    geom_point(size=0.5) + 
    geom_smooth(method="lm", color="mediumorchid", ...)
  p
}

(cor.DR.effects.p <- DR.coef.df %>% 
  dplyr::select(`1D`, `2D`, `20`, `40`) %>%  
  ggpairs(lower = list(continuous = my_fn), 
          diag = NULL,
          progress = F) +
  theme_classic(base_size=8) +
  labs(x="DR coef", y="DR coef"))
```

```{r}
# pdf("../plots/asreml_diet_similarity_pathway_scatter.pdf", height=4, width=4)
# cor.DR.effects.p
# dev.off()
```

## Pathways different between CR and fasting

```{r}
pathway.CR.minus.fasting.mean.solution.df <- pathway.DR.df %>% 
  pivot_wider(id_cols=feature, names_from=fixef.value, values_from=solution) %>% 
  rowwise() %>% mutate(CR.mean = mean(c(`20`, `40`)), fasting.mean = mean(c(`1D`, `2D`))) %>% ungroup() %>% 
  mutate(CR.minus.fasting.mean = CR.mean - fasting.mean) %>% 
  mutate(CR.and.fasting.in.diff.dir = sign(CR.mean) != sign(fasting.mean))
```

```{r}
pathway.CR.or.fasting.mean.coef.df <- pathway.DR.df %>% 
  mutate(CR.or.fasting = ifelse(fixef.value %in% c("20", "40"), "CR", "fasting")) %>% 
  group_by(feature, CR.or.fasting) %>% 
  summarise(solution = mean(solution), .groups="drop") %>% 
  group_by()

pathway.CR.or.fasting.mean.coef.order.df <- pathway.CR.or.fasting.mean.coef.df %>% 
  dplyr::filter(CR.or.fasting == "CR") %>% 
  arrange(solution) %>%
  mutate(ii = row_number()) %>% 
  dplyr::select(feature, ii)
  
(pathway.CR.v.fasting.mean.coef.p <- pathway.CR.or.fasting.mean.coef.df %>% 
    merge(pathway.CR.minus.fasting.mean.solution.df %>% 
            dplyr::select(feature, CR.and.fasting.in.diff.dir), by="feature") %>% 
    merge(pathway.CR.or.fasting.mean.coef.order.df, by="feature") %>% 
    ggplot(aes(x=reorder(feature, ii), y=solution, color=CR.or.fasting, alpha=CR.and.fasting.in.diff.dir)) +
    geom_hline(yintercept=0, lty=2) +
    geom_line(aes(group=feature, color=NULL)) +
    geom_point(shape=16, size=0.5) +
    scale_alpha_manual(values=c(`TRUE`=1, `FALSE`=0.2)) +
    scale_color_manual(values=CR.v.fasting.palette) +
    theme_classic(base_size=8) +
    theme(axis.text.x=element_blank(),
          axis.ticks.x=element_blank(),
          legend.position="none") +
    labs(x="Pathway", y="Coefficient"))
```

```{r}
# pdf("../plots/pathway_CR_v_fasting_mean_coef.pdf", height=2, width=3)
# pathway.CR.v.fasting.mean.coef.p
# dev.off()
```

## Examples

```{r}
# str_subset(rownames(pathway.log2tpm.df), "urea")

(urea.v.diet.p <- pathway.log2tpm.df["PWY-4984: urea cycle", ] %>% 
   t() %>% 
   data.frame() %>% 
   setNames("log2tpm") %>% 
   rownames_to_column("stool.ID") %>% 
   merge(stool.meta.annot.df, by="stool.ID") %>%  
   mutate(Diet.5mo.as.AL=factor(Diet.5mo.as.AL, levels=c("AL", "1D", "2D", "20", "40"))) %>% 
   dplyr::filter(age.approx.months <= 40) %>% 
   ggplot(aes(x=Diet.5mo.as.AL, y=log2tpm, fill=Diet.5mo.as.AL)) +
   geom_boxplot(outlier.shape=16, outlier.size=0.5) +
   stat_compare_means(method="t.test", label="p.signif", ref.group="AL") +
   labs(x="", y="log2(TPM)", title="urea cycle (PWY-4984)") +
   scale_fill_manual(values=diet.palette) +
   theme_classic(base_size=8) +
   theme(legend.position="none"))
```

```{r}
# pdf("../plots/urea_cycle_v_diet.pdf", height=1.5, width=1.5)
# urea.v.diet.p
# dev.off()
```


```{r}
# str_subset(rownames(pathway.log2tpm.df), "citrulline")

(citrulline.v.diet.p <- pathway.log2tpm.df["CITRULBIO-PWY: L-citrulline biosynthesis", ] %>% 
  t() %>% 
  data.frame() %>% 
  setNames("log2tpm") %>% 
  rownames_to_column("stool.ID") %>% 
  merge(stool.meta.annot.df, by="stool.ID") %>% 
  mutate(Diet.5mo.as.AL=factor(Diet.5mo.as.AL, levels=c("AL", "1D", "2D", "20", "40"))) %>% 
  dplyr::filter(age.approx.months <= 40) %>% 
  ggplot(aes(x=Diet.5mo.as.AL, y=log2tpm, fill=Diet.5mo.as.AL)) +
  geom_boxplot(outlier.shape=16, outlier.size=0.5) +
  # facet_wrap(~age.approx.months, nrow=1) +
   stat_compare_means(method="t.test", label="p.signif", ref.group="AL") +
  labs(x="", y="log2(TPM)", title="L-citrulline biosynthesis (CITRULBIO-PWY)") +
  scale_fill_manual(values=diet.palette) +
  theme_classic(base_size=8) +
  theme(legend.position="none"))
```

```{r}
# pdf("../plots/citrulline_biosynth_v_diet.pdf", height=1.5, width=1.5)
# citrulline.v.diet.p
# dev.off()
```


# Heritability

Add long pathway names.

```{r}
pathway.herit.df <- pathway.herit.df %>% 
  merge(pathway.names.df, by.x="feature", by.y="short.clean", all.x=T) # all.x=T for uniqueness
```

Adjust p-values.

```{r}
pathway.herit.df <- pathway.herit.df %>% 
  mutate(padj = p.adjust(LRT.pval, method="BH")) %>% 
  mutate(is.sig = padj < PADJ.THRESH)
```

```{r}
pathway.herit.df %>% 
  dplyr::filter(feature != "Uniqueness") %>%
  dplyr::count(is.sig) %>%
  pull(n) %>% prop.table()
```

* 51% of pathways are heritable

```{r}
(mean.herit.for.heritable.pathways <- pathway.herit.df %>% 
   dplyr::filter(feature != "Uniqueness") %>%
   dplyr::filter(is.sig) %>% 
   summarise(mean = mean(herit)) %>% 
   pull(mean))
```

* Mean heritability is 8.5%

## Community-wide versus specialized

```{r}
herit.pathways <- c(
  
  # heritable, specialized
  "LACTOSECAT-PWY: lactose and galactose degradation I",
  
  # heritable, community-wide
  "PWY-6700: queuosine biosynthesis I (de novo)", 
  "PWY66-429: fatty acid biosynthesis initiation (mitochondria)"

  )
```


```{r}
(herit.comm.v.specialized.histograms <- genus.v.pathway.cor.long.df %>%
   dplyr::filter(pathway %in% herit.pathways) %>% 
   ggplot(aes(cor, fill=pathway.type)) +
   geom_histogram() +
   geom_vline(xintercept=0.5, lty=2) +
   facet_wrap(~pathway.type + pathway, nrow=1, scales="free_y") +
   labs(x="Correlation with a genus", y="# correlations") +
   scale_fill_manual(values=pathway.type.palette) +
   scale_x_continuous(breaks=c(-0.5, 0, 0.5)) +
   theme_classic(base_size=8) +
   theme(legend.position="none"))
```

```{r}
pdf("../plots/community_wide_v_specialized_pathway_histograms_heritable.pdf", width=3, height=1.5)
herit.comm.v.specialized.histograms
dev.off()
```

## Barplot

```{r}
herit.pathways.short <- c(
  
  # heritable, specialized
  "LACTOSECATPWY", 
  
  # heritable, community-wide
  "PWY6700", "PWY66429"

  )
```

```{r}
(humann.herit.p <- pathway.herit.df %>% 
   dplyr::filter(herit > 5e-7) %>% # omit 9 pathways with zero herit and small error for visual purposes
   
   mutate(label = case_when(
     feature %in% herit.pathways ~ after.colon.short.w.ID,
     TRUE ~ ""
   )) %>%
   
   ggplot(aes(x=herit, xmin=herit-herit.se, xmax=herit+herit.se,
              y=reorder(feature, herit), label=label, fill=is.sig)) +
   geom_col() +
   geom_linerange(linewidth=0.25) +
   geom_vline(xintercept = mean.herit.for.heritable.pathways, 
              color=genetics.color, lty=2) +
   geom_text_repel(max.overlaps=Inf, nudge_x=0.01, min.segment.length=0, size=0.36*6) +
   labs(x="Heritability", y="Pathway", fill="padj < 0.01?") +
   scale_fill_manual(values=c(`TRUE`=genetics.color, `FALSE`="gray80")) +
   theme_classic(base_size=8) +
   theme(legend.position="none",
         axis.text.y = element_blank(),
         axis.ticks.y = element_blank()))
```

```{r}
# pdf("../plots/pathway_heritability_barplot.pdf", height=3, width=1.5)
# humann.herit.p
# dev.off()
```

# All random effects

```{r}
pathway.ranef.df.all.ranef.model %>% 
  group_by(term.clean) %>% 
  summarise(PVE = mean(PVE)) %>% 
  arrange(desc(PVE))
```

* 76% of variance remains unexplained
* Genetics explains mean 5% of variance with all random effects model

Adjust p-values within each experimental variable.

```{r}
pathway.ranef.df.all.ranef.model <- pathway.ranef.df.all.ranef.model %>% 
  group_by(term.clean) %>% 
  mutate(padj = p.adjust(LRT.pval, method="BH")) %>% 
  mutate(is.sig = padj < PADJ.THRESH)
```

```{r}
(pathway.all.ranef.PVE.p <- pathway.ranef.df.all.ranef.model %>% 
   dplyr::filter(term.clean != "Residual") %>% 
   arrange(is.sig) %>% 
   ggplot(aes(x=reorder(term.clean, -PVE), y=PVE, color=is.sig)) +
   geom_jitter(width=0.2, shape=16) +
   stat_summary(aes(color=NULL), fun="mean", geom="crossbar", width=0.75) +
   labs(x="Experimental variable", y="Proportion of variance explained") +
   theme_classic(base_size=8) +
   theme(legend.position="none") +
   scale_color_manual(values=c(`TRUE`="gray20", `FALSE`="gray80")))
```

```{r}
# pdf("../plots/PVE_all_ranef_pathways.pdf", height=2, width=2)
# pathway.all.ranef.PVE.p
# dev.off()
```

# Appendix: universal age-associated pathways

We previously identified 5 pathways that change with age in both mice and humans. Are these specialized or community-wide pathways?

```{r}
universal.age.assoc.pathways <- c(
  "PWY-7237: myo-, chiro- and scyllo-inositol degradation",
  "RIBOSYN2-PWY: flavin biosynthesis I (bacteria and plants)",
  "PWY-7357: thiamine phosphate formation from pyrithiamine and oxythiamine (yeast)",
  "PWY-7234: inosine-5'-phosphate biosynthesis III",
  "PWY-6897: thiamine diphosphate salvage II"
  )
```

```{r}
(univ.age.assoc.comm.v.specialized.histograms <- genus.v.pathway.cor.long.df %>%
   dplyr::filter(pathway %in% universal.age.assoc.pathways) %>% 
   ggplot(aes(cor, fill=pathway.type)) +
   geom_histogram() +
   geom_vline(xintercept=0.5, lty=2) +
   facet_wrap(~pathway.type + pathway, nrow=1, scales="free_y") +
   labs(x="Correlation with a genus", y="# correlations") +
   scale_fill_manual(values=pathway.type.palette) +
   theme_classic(base_size=8) +
   theme(legend.position="none"))
```

```{r}
# pdf("../plots/community_wide_v_specialized_pathway_histograms_univ_age_assoc.pdf", width=6, height=1.5)
# univ.age.assoc.comm.v.specialized.histograms
# dev.off()
```
