---
title: "Analysis of taxonomic results based on Kraken"
---

**2024-10-19 updates:**
- Time is a random effect instead of fixed effect
- Plotting individual genera against age using mean ± SEM, instead of boxplots

# Load libraries

```{r, warning=F, message=F}
library(tidyverse)
library(foreach)
library(phyloseq)
library(cowplot) # plot_grid
library(speedyseq) # faster tax_glom
library(patchwork) # for mega PCoA
library(ggpubr) # for t-tests
library(ggrepel) # for labels
library(corrr) # stretch, DR scatterplots
library(GGally) # ggpairs, DR scatterplots

PADJ.THRESH <- 0.01

age.color <- "seagreen"
genetics.color <- "gold"
diet.color <- "mediumorchid"

diet.palette <- c(AL="seashell4",
                  `1D`="skyblue",
                  `2D`="royalblue4",
                  `20`="orange",
                  `40`="firebrick")
CR.v.fasting.palette <- c(fasting="royalblue4", CR="firebrick")
```

# Import Kraken data

```{r}
agg.kraken.df <- read.table(
  "../data/kraken_matrix_agg_by_stool_ID_n1303x2997.txt", 
  sep="\t", header=T, row.names=1)
dim(agg.kraken.df)
```

* 1303 taxons x 2997 stool IDs

# Import metadata

```{r}
stool.meta.df <- read.table(
  "../data/metadata/stool_metadata_after_QC_no_controls_n2997_240418.txt", 
  sep="\t", header=T)
mouse.meta.df <- read.csv("../data/metadata/AnimalData_Processed_20230712.csv")
stool.meta.annot.df <- stool.meta.df %>%
  merge(mouse.meta.df, by.x="mouse.ID", by.y="MouseID")

dim(stool.meta.annot.df)
n_distinct(stool.meta.annot.df$stool.ID)
```

* 2997 (unique) stool IDs, as expected

Pre-randomization timepoints are AL.

```{r}
stool.meta.annot.df <- stool.meta.annot.df %>% 
  mutate(Diet.5mo.as.AL=case_when(
    age.wks < 25 ~ "AL",
    TRUE ~ as.character(Diet)))
```

Create character version of cage.

```{r}
stool.meta.annot.df <- stool.meta.annot.df %>% 
  mutate(Cage=paste0("c", HID))
```

# Import Kraken taxonomy

```{r}
kraken.tax.df <- read.table(
  "../data/kraken_taxonomy_n1303.txt", 
  sep="\t", header=T, quote="", row.names=1)
dim(kraken.tax.df)
```

# Import ASReml results

## Age and DR as fixed effects

Model: y_mb ~ age (f) + diet (f) + time(r) + [genetics] (r) + mouse (r) + batch (r) + cohort (r) + cage (r)

```{r}
fixef.df <- read.table(
  "../results/asreml_kraken_genus_w_time_ranef/fixef.txt",
  sep="\t", header=T)

ranef.df <- read.table(
  "../results/asreml_kraken_genus_w_time_ranef/ranef.txt",
  sep="\t", header=T)

herit.df <- read.table(
  "../results/asreml_kraken_genus_w_time_ranef/herit.txt",
  sep="\t", header=T)
```

## All random effects

Model: y_mb ~ age (r) + diet (r) + time (r) + [genetics] (r) + mouse (r) + batch (r) + cohort (r) + cage (r)

```{r}
ranef.df.all.ranef.model <- read.table(
  "../results/asreml_kraken_genus_all_ranef_w_time/ranef.txt",
  sep="\t", header=T)
```

# Create phyloseq object

```{r}
kraken.sample.meta.df <- stool.meta.annot.df %>% column_to_rownames("stool.ID")

agg.physeq <- phyloseq(
  agg.kraken.df %>% as.matrix %>% otu_table(taxa_are_rows=T),
  kraken.sample.meta.df %>% sample_data,
  kraken.tax.df %>% as.matrix %>% tax_table
)
nsamples(agg.physeq)
```

* 2997 samples

# Aggregate

```{r}
agg.physeq.genus <- agg.physeq %>% tax_glom(taxrank="genus")
agg.physeq.species <- agg.physeq %>% tax_glom(taxrank="species")
```

```{r}
ntaxa(agg.physeq)
ntaxa(agg.physeq.genus)
ntaxa(agg.physeq.species)
```

* 1303 taxa --> 252 genera
* 1303 taxa --> 1093 species

```{r}
sum(otu_table(agg.physeq.genus)) / sum(otu_table(agg.physeq))
sum(otu_table(agg.physeq.species)) / sum(otu_table(agg.physeq))
```

* 85% of reads could be assigned to genus level
* 80% of reads could be assigned to species level

# Compute distances

Bray-Curtis on genus-level relative abundances.

```{r}
agg.genus.bc.dist <- agg.physeq.genus %>%
  transform_sample_counts(function(x) x/sum(x)) %>%
  phyloseq::distance(method="bray")
```

# Alpha diversity

```{r}
alpha.div.df.genus <- estimate_richness(
  agg.physeq.genus, 
  measures=c("Observed", "Chao1", "Shannon", "Simpson")) %>% 
  rownames_to_column("stool.ID") %>% 
  merge(stool.meta.annot.df, by="stool.ID")

alpha.div.long.df.genus <- alpha.div.df.genus %>% 
  pivot_longer(c(Observed, Chao1, Shannon, Simpson), names_to="metric")
```

## Shannon

Versus age.

```{r}
# summarize per diet 
shannon.plot.df.per.diet <- alpha.div.long.df.genus %>% 
   dplyr::filter(metric == "Shannon") %>% 
   dplyr::filter(age.approx.months <= 40) %>% # omit the few 46 month samples
  mutate(Diet=factor(Diet.5mo.as.AL, levels=c("AL", "1D", "2D", "20", "40"))) %>%
  
  # compute mean and SD for each age and diet combination
  group_by(Diet, age.approx.months) %>% 
  summarise(n=n(),
            mean = mean(value),
            sd = sd(value),
            sem = sd(value)/sqrt(n()), .groups="drop") %>% 
  
  # make sure we have at least 3 observations at this age combination
  dplyr::filter(n >= 3)

# summarize across diets
shannon.plot.df.across.diets <- alpha.div.long.df.genus %>% 
   dplyr::filter(metric == "Shannon") %>% 
   dplyr::filter(age.approx.months <= 40) %>% # omit the few 46 month samples
  
  # compute mean and SD across diets
  group_by(age.approx.months) %>% 
  summarise(n=n(),
            mean = mean(value),
            sd = sd(value),
            sem = sd(value)/sqrt(n()), .groups="drop") %>% 
  
  # make sure we have at least 3 observations at this age combination
  dplyr::filter(n >= 3)
```

```{r}
(shannon.v.age.mean.SEM.p <- shannon.plot.df.per.diet %>% 
   ggplot(aes(x=age.approx.months, y=mean, ymin=mean-sem, ymax=mean+sem,
              color=Diet)) +
   
   # per diet
   geom_path(alpha=0.3, linewidth=0.5) +
   geom_pointrange(alpha=0.3, size=0.1, shape=16) +
   
   # across diets
   geom_path(aes(color=NULL), linewidth=0.5, data=shannon.plot.df.across.diets) +
   geom_pointrange(aes(color=NULL), size=0.1, shape=16, data=shannon.plot.df.across.diets) +
   
   scale_color_manual(values=diet.palette) +
   labs(x="Age (months)", y="Shannon index", title="Shannon") +
   geom_vline(xintercept=6, lty=2) +
   theme_classic(base_size=8) +
   theme(legend.position="none"))
```

```{r}
pdf("../plots/shannon_v_age_mean_SEM.pdf", height=2, width=2)
shannon.v.age.mean.SEM.p
dev.off()
```

## Simpson

Versus age.

```{r}
# summarize per diet 
simpson.plot.df.per.diet <- alpha.div.long.df.genus %>% 
   dplyr::filter(metric == "Simpson") %>% 
   dplyr::filter(age.approx.months <= 40) %>% # omit the few 46 month samples
  mutate(Diet=factor(Diet.5mo.as.AL, levels=c("AL", "1D", "2D", "20", "40"))) %>%
  
  # compute mean and SD for each age and diet combination
  group_by(Diet, age.approx.months) %>% 
  summarise(n=n(),
            mean = mean(value),
            sd = sd(value),
            sem = sd(value)/sqrt(n()), .groups="drop") %>% 
  
  # make sure we have at least 3 observations at this age combination
  dplyr::filter(n >= 3)

# summarize across diets
simpson.plot.df.across.diets <- alpha.div.long.df.genus %>% 
   dplyr::filter(metric == "Simpson") %>% 
   dplyr::filter(age.approx.months <= 40) %>% # omit the few 46 month samples
  
  # compute mean and SD across diets
  group_by(age.approx.months) %>% 
  summarise(n=n(),
            mean = mean(value),
            sd = sd(value),
            sem = sd(value)/sqrt(n()), .groups="drop") %>% 
  
  # make sure we have at least 3 observations at this age combination
  dplyr::filter(n >= 3)
```

```{r}
(simpson.v.age.mean.SEM.p <- simpson.plot.df.per.diet %>% 
   ggplot(aes(x=age.approx.months, y=mean, ymin=mean-sem, ymax=mean+sem,
              color=Diet)) +
   
   # per diet
   geom_path(alpha=0.3, linewidth=0.5) +
   geom_pointrange(alpha=0.3, size=0.1, shape=16) +
   
   # across diets
   geom_path(aes(color=NULL), linewidth=0.5, data=simpson.plot.df.across.diets) +
   geom_pointrange(aes(color=NULL), size=0.1, shape=16, data=simpson.plot.df.across.diets) +
   
   scale_color_manual(values=diet.palette) +
   labs(x="Age (months)", y="Simpson index", title="Simpson") +
   geom_vline(xintercept=6, lty=2) +
   theme_classic(base_size=8) +
   theme(legend.position="none"))
```

```{r}
pdf("../plots/simpson_v_age_mean_SEM.pdf", height=2, width=2)
simpson.v.age.mean.SEM.p
dev.off()
```

# PCoA

```{r}
agg.pcoa.genus.bc <- agg.genus.bc.dist %>% 
  cmdscale(eig=T, k=3)

agg.pcoa.genus.bc.df <- merge(
  data.frame(agg.pcoa.genus.bc$points) %>% setNames(c("PCoA1", "PCoA2", "PCoA3")),
  kraken.sample.meta.df,
  by="row.names") %>% 
  mutate(Diet=factor(Diet, levels=c("AL", "1D", "2D", "20", "40")),
         Diet.5mo.as.AL=factor(Diet.5mo.as.AL, levels=c("AL", "1D", "2D", "20", "40")))
```

```{r}
frac.var.explained.by.pcos <- agg.pcoa.genus.bc$eig / sum(agg.pcoa.genus.bc$eig)
frac.var.explained.by.pcos[1:3]
```

```{r}
full.pcoa.p <- agg.pcoa.genus.bc.df %>% 
  arrange(desc(age.approx.months)) %>%
  ggplot(aes(x=PCoA1, y=PCoA2, color=Diet.5mo.as.AL, size=age.approx.months)) +
  geom_point(shape=16, alpha=0.8) +
  scale_color_manual(values=diet.palette) +
  scale_size_continuous(breaks=c(5,10,16,22,28,34,40,46), range=c(0.25,2))
```

## With barplots

```{r}
pcoa1.lims <- layer_scales(full.pcoa.p, 1, 1)$x$range$range
pcoa1.intervals <- seq(from=pcoa1.lims[1], to=pcoa1.lims[2], length.out=20)

pcoa1.age.barplots <- agg.pcoa.genus.bc.df %>% 
  mutate(PCoA1.discrete=cut(PCoA1, pcoa1.intervals, include.lowest=T)) %>%
  group_by(PCoA1.discrete) %>%
  summarise(mean.age.mo=mean(age.approx.months), .groups="drop") %>%
  ggplot(aes(x=PCoA1.discrete, y=mean.age.mo)) +
  geom_col() +
  theme_classic(base_size=8) +
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
```

```{r}
pcoa2.lims <- layer_scales(full.pcoa.p, 1, 1)$y$range$range
pcoa2.intervals <- seq(from=pcoa2.lims[1], to=pcoa2.lims[2], length.out=20)

pcoa2.age.barplots <- agg.pcoa.genus.bc.df %>% 
  mutate(PCoA2.discrete=cut(PCoA2, pcoa2.intervals, include.lowest=T)) %>%
  group_by(PCoA2.discrete) %>%
  summarise(mean.age.mo=mean(age.approx.months), .groups="drop") %>%
  ggplot(aes(x=mean.age.mo, y=PCoA2.discrete)) +
  geom_col() +
  theme_classic(base_size=8) +
  theme(axis.text.y=element_blank(),
        axis.ticks.y=element_blank())
```

```{r}
pcoa1.diet.boxplots <- agg.pcoa.genus.bc.df %>% 
  ggplot(aes(x=PCoA1, y=fct_rev(Diet.5mo.as.AL), fill=Diet.5mo.as.AL)) +
  geom_boxplot(outlier.shape=16, outlier.size=0.5) +
  scale_fill_manual(values=diet.palette)
```

```{r}
pcoa2.diet.boxplots <- agg.pcoa.genus.bc.df %>% 
  ggplot(aes(x=Diet.5mo.as.AL, y=PCoA2, fill=Diet.5mo.as.AL)) +
  geom_boxplot(outlier.shape=16, outlier.size=0.5) +
  scale_fill_manual(values=diet.palette)
```

```{r}
mega.pcoa.layout <- "
#AAAAAA#
BCCCCCCD
BCCCCCCD
BCCCCCCD
BCCCCCCD
BCCCCCCD
BCCCCCCD
#EEEEEE#
"

(mega.pcoa <- (
  
  # A = PCoA1 diet boxplots
  pcoa1.diet.boxplots + 
    scale_x_continuous(position="top") + 
    labs(x=sprintf("PCoA1 (%.0f%%)", 100*frac.var.explained.by.pcos[1])) +
    theme_classic(base_size=8) +
    theme(legend.position="none", 
          axis.title.y=element_blank()) +
    
    # B = PCoA2 diet boxplots
    pcoa2.diet.boxplots + 
    labs(y=sprintf("PCoA2 (%.0f%%)", 100*frac.var.explained.by.pcos[2])) +
    theme_classic(base_size=8) +
    theme(legend.position="none", 
          axis.title.x=element_blank()) +
    
    # C = PCoA
    full.pcoa.p + 
    theme_classic(base_size=8) +
    theme(legend.position="none",
          axis.title.x=element_blank(),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank(),
          axis.title.y=element_blank(),
          axis.text.y=element_blank(),
          axis.ticks.y=element_blank()) +
    
    # D = PCoA2 age barplots
    pcoa2.age.barplots + 
    scale_y_discrete(position="right") +
    scale_x_continuous(position="top") +
    labs(x="Mean age (mo)", y="Binned PCoA2") +
    
    # E = PCoA1 age barplots
    pcoa1.age.barplots + 
    scale_y_reverse(position="right") +
    labs(x="Binned PCoA1", y="Mean age (mo)")) +
    
    plot_layout(design=mega.pcoa.layout))
```

```{r}
pdf("../plots/full_pcoa_w_barplots.pdf", height=5, width=5)
mega.pcoa
dev.off()
```

```{r}
pdf("../plots/full_pcoa_legend.pdf", height=3, width=1)
plot_grid(get_legend(full.pcoa.p))
dev.off()
```

## Just young and old, AL versus 40% CR

Ordination with just these samples.
  
```{r}
agg.genus.bc.dist.just.young.v.old.AL.v.40 <- agg.physeq.genus %>%
  subset_samples(age.approx.months %in% c(10, 28)) %>% 
  subset_samples(Diet %in% c("AL", "40")) %>%
  transform_sample_counts(function(x) x/sum(x)) %>%
  phyloseq::distance(method="bray")

agg.pcoa.genus.bc.res.just.young.v.old.AL.v.40 <- agg.genus.bc.dist.just.young.v.old.AL.v.40 %>% 
  cmdscale(eig=T, k=3)

frac.var.explained.by.pcos.just.young.v.old.AL.v.40 <- (
  agg.pcoa.genus.bc.res.just.young.v.old.AL.v.40$eig) / (
    sum(agg.pcoa.genus.bc.res.just.young.v.old.AL.v.40$eig))

agg.pcoa.genus.bc.df.just.young.v.old.AL.v.40 <- merge(
  data.frame(agg.pcoa.genus.bc.res.just.young.v.old.AL.v.40$points) %>% setNames(c("PCoA1", "PCoA2", "PCoA3")),
  kraken.sample.meta.df,
  by="row.names") %>% 
  mutate(Diet=factor(Diet, levels=c("AL", "40")))
```

```{r}
AL.CR.10mo.28mo.new.ordination.centroids.df <- agg.pcoa.genus.bc.df.just.young.v.old.AL.v.40 %>% 
  dplyr::filter(age.approx.months %in% c(10, 28)) %>% 
  dplyr::filter(Diet %in% c("AL", "40")) %>% 
  group_by(age.approx.months, Diet) %>% 
  summarise(PCoA1=mean(PCoA1), PCoA2=mean(PCoA2), .groups="drop")
```

```{r}
(AL.CR.10mo.28mo.new.ordination.p <- agg.pcoa.genus.bc.df.just.young.v.old.AL.v.40 %>% 
   ggplot(aes(x=PCoA1, y=PCoA2, color=Diet, shape=factor(age.approx.months))) +
   geom_point(size=0.5, alpha=0.3) +
   geom_point(data=AL.CR.10mo.28mo.new.ordination.centroids.df, size=2) +
   stat_ellipse(aes(linetype=factor(age.approx.months))) +
   scale_shape_manual(values=c(`10`=16, `28`=1)) +
   scale_color_manual(values=diet.palette) +
   labs(x=sprintf("PCoA1 (%.1f%%)", 100*frac.var.explained.by.pcos.just.young.v.old.AL.v.40[1]),
        y=sprintf("PCoA2 (%.1f%%)", 100*frac.var.explained.by.pcos.just.young.v.old.AL.v.40[2]),
        shape="Age (mo)", 
        linetype="Age (mo)", 
        title="Ordination based on just these samples") +
   theme_classic(base_size=8))
```

We see that 40% CR pulls both young and old samples to the left and up, and age also pulls samples left and up
  
```{r}
pdf("../plots/pcoa_AL_v_40CR_10mo_v_28mo.pdf", height=2, width=2)
AL.CR.10mo.28mo.new.ordination.p + theme(legend.position="none")
dev.off()
```

# Uniqueness

```{r}
agg.genus.bc.dist.mat <- agg.genus.bc.dist %>% as.matrix
diag(agg.genus.bc.dist.mat) <- NA
agg.genus.uniq.bc.df <- data.frame(
  uniqueness=apply(agg.genus.bc.dist.mat, MARGIN=2, FUN=function(x) {min(x, na.rm=T)}),
  stool.ID=sample_names(agg.physeq.genus)) %>%
  merge(stool.meta.annot.df, by="stool.ID")
```

## Versus age

```{r}
agg.genus.uniq.bc.lm.summary <- agg.genus.uniq.bc.df %>%
  dplyr::filter(age.approx.months <= 40) %>% # omit few 46 month samples
  lm(uniqueness ~ age.approx.months, data=.) %>% summary

agg.genus.uniq.bc.lm.pval <- agg.genus.uniq.bc.lm.summary$coefficients["age.approx.months", "Pr(>|t|)"]
agg.genus.uniq.bc.lm.beta <- agg.genus.uniq.bc.lm.summary$coefficients["age.approx.months", "Estimate"]
```

```{r}
# summarize per diet 
uniq.plot.df.per.diet <- agg.genus.uniq.bc.df %>% 
  dplyr::filter(age.approx.months <= 40) %>% # omit few 46 month samples
  mutate(Diet=factor(Diet.5mo.as.AL, levels=c("AL", "1D", "2D", "20", "40"))) %>%
  
  # compute mean and SD for each age and diet combination
  group_by(Diet, age.approx.months) %>% 
  summarise(n=n(),
            mean = mean(uniqueness),
            sd = sd(uniqueness),
            sem = sd(uniqueness)/sqrt(n()), .groups="drop") %>% 
  
  # make sure we have at least 3 observations at this date and age combination
  dplyr::filter(n >= 3)

# summarize across diets
uniq.plot.df.across.diets <- agg.genus.uniq.bc.df %>% 
  dplyr::filter(age.approx.months <= 40) %>% # omit few 46 month samples
  
  # compute mean and SD across diets
  group_by(age.approx.months) %>% 
  summarise(n=n(),
            mean = mean(uniqueness),
            sd = sd(uniqueness),
            sem = sd(uniqueness)/sqrt(n()), .groups="drop") %>% 
  
  # make sure we have at least 3 observations at this date and age combination
  dplyr::filter(n >= 3)
```

```{r}
# plot
(uniq.v.age.mean.SEM.p <- uniq.plot.df.per.diet %>% 
    ggplot(aes(x=age.approx.months, y=mean, ymin=mean-sem, ymax=mean+sem,
               color=Diet)) +
    
    # per diet
    geom_path(alpha=0.3, linewidth=0.5) +
    geom_pointrange(alpha=0.3, size=0.01) +
    
    # across diets
    geom_path(aes(color=NULL), linewidth=0.5, data=uniq.plot.df.across.diets) +
    geom_pointrange(aes(color=NULL), size=0.01, data=uniq.plot.df.across.diets) +
    
    scale_color_manual(values=diet.palette) +
    labs(x="Age (months)", y="Uniqueness", color="Diet",
         title="Uniqueness") +
    geom_vline(xintercept=6, lty=2) +
    theme_classic(base_size=8))
```

```{r}
pdf("../plots/tax_uniq_v_age_mean_SEM.pdf", width=2, height=2)
uniq.v.age.mean.SEM.p + theme(legend.position="none")
dev.off()
```

## Versus diet

```{r}
(uniq.v.diet.across.ages.p <- agg.genus.uniq.bc.df %>% 
   dplyr::filter(age.approx.months <= 40) %>% 
   mutate(Diet.5mo.as.AL=factor(Diet.5mo.as.AL, levels=c("AL", "1D", "2D", "20", "40"))) %>%
   ggplot(aes(x=Diet.5mo.as.AL, y=uniqueness, fill=Diet.5mo.as.AL)) +
   geom_boxplot(outlier.shape=16, outlier.size=0.5) +
   stat_compare_means(method="t.test", label="p.signif", ref.group="AL") +
   labs(x="Dietary group", y="Uniqueness", 
        title="Taxonomic uniqueness") +
   theme_classic(base_size=8) +
   theme(legend.position="none") +
   scale_fill_manual(values=diet.palette)
)
```

```{r}
# pdf("../plots/tax_uniq_v_diet_boxplots.pdf", width=1.5, height=1.5)
# uniq.v.diet.across.ages.p
# dev.off()
```

## Versus age, equal number of mice per cage

```{r}
uniq.df.n.mice.per.cage <- foreach(n=c(4,5,6,7,8), .combine="rbind") %do% {
  
  this.n.per.cage.stool.IDs <- stool.meta.annot.df %>%
    
    # only keep cage-age combinations with at least n mice
    group_by(Cage, age.approx.months) %>%
    mutate(n.per.group=n()) %>% 
    ungroup() %>% 
    dplyr::filter(n.per.group >= n) %>% 
    
    # if more than n mice per cage, randomly select n mice
    group_by(Cage, age.approx.months) %>%
    slice_sample(n=n) %>%
    pull(stool.ID)
  
  this.n.per.cage.dist <- agg.physeq.genus %>% 
    prune_samples(samples=this.n.per.cage.stool.IDs) %>% 
    transform_sample_counts(function(x) x/sum(x)) %>%
    phyloseq::distance(method="bray")
  
  this.n.per.cage.dist.mat <- this.n.per.cage.dist %>% as.matrix()
  diag(this.n.per.cage.dist.mat) <- NA
  this.n.per.cage.uniq.df <- data.frame(
    uniqueness=apply(this.n.per.cage.dist.mat, MARGIN=2, FUN=function(x) {min(x, na.rm=T)}),
    stool.ID=rownames(this.n.per.cage.dist.mat)) %>% 
    mutate(n.per.cage = n)
  
  return(this.n.per.cage.uniq.df)
}

uniq.annot.df.n.mice.per.cage <- uniq.df.n.mice.per.cage %>% 
  merge(stool.meta.annot.df, by="stool.ID")
```

Mean ± SEM.

```{r}
# summarize per diet 
uniq.n.mice.per.cage.plot.df.per.diet <- uniq.annot.df.n.mice.per.cage %>% 
  mutate(Diet=factor(Diet.5mo.as.AL, levels=c("AL", "1D", "2D", "20", "40"))) %>%
  
  # compute mean and SD for each age, diet, and n combination
  group_by(Diet, age.approx.months, n.per.cage) %>% 
  summarise(n=n(),
            mean = mean(uniqueness),
            sd = sd(uniqueness),
            sem = sd(uniqueness)/sqrt(n()), .groups="drop") %>% 
  
  # make sure we have at least 3 observations at this date and age combination
  dplyr::filter(n >= 3)

# summarize across diets
uniq.n.mice.per.cage.plot.df.across.diets <- uniq.annot.df.n.mice.per.cage %>% 
  
  # compute mean and SD across diets and n.per.cage
  group_by(age.approx.months, n.per.cage) %>% 
  summarise(n=n(),
            mean = mean(uniqueness),
            sd = sd(uniqueness),
            sem = sd(uniqueness)/sqrt(n()), .groups="drop") %>% 
  
  # make sure we have at least 3 observations at this date and age combination
  dplyr::filter(n >= 3)
```

Wide plot.

```{r}
(uniq.v.age.fixed.n.per.cage.mean.SEM.p <- uniq.n.mice.per.cage.plot.df.per.diet %>% 
   ggplot(aes(x=age.approx.months, y=mean, ymin=mean-sem, ymax=mean+sem,
              color=Diet)) +
   
   # per diet
   geom_path(alpha=0.3, linewidth=0.5) +
   geom_pointrange(alpha=0.3, size=0.1, shape=16) +
   
   # across diets
   geom_path(aes(color=NULL), linewidth=0.5, data=uniq.n.mice.per.cage.plot.df.across.diets) +
   geom_pointrange(aes(color=NULL), size=0.1, shape=16, data=uniq.n.mice.per.cage.plot.df.across.diets) +
   
   scale_color_manual(values=diet.palette) +
   labs(x="Age (months)", y="Uniqueness", 
        title="Taxonomic uniqueness, downsampled to n mice per cage") +
   facet_wrap(~n.per.cage, ncol=1,
              labeller=labeller(n.per.cage=function(x) {sprintf("n=%s", x)})) +
   geom_vline(xintercept=6, lty=2) +
   theme_classic(base_size=8) +
   theme(legend.position="none"))
```

```{r}
pdf("../plots/tax_uniq_v_age_fixed_n_per_cage_mean_SEM.pdf", height=6, width=2)
uniq.v.age.fixed.n.per.cage.mean.SEM.p
dev.off()
```

## Species

```{r}
agg.species.bc.dist <- agg.physeq.species %>%
  transform_sample_counts(function(x) x/sum(x)) %>%
  phyloseq::distance(method="bray")

agg.species.bc.dist.mat <- agg.species.bc.dist %>% as.matrix
diag(agg.species.bc.dist.mat) <- NA
agg.species.uniq.bc.df <- data.frame(
  uniqueness=apply(agg.species.bc.dist.mat, MARGIN=2, FUN=function(x) {min(x, na.rm=T)}),
  stool.ID=sample_names(agg.physeq.genus)) %>%
  merge(stool.meta.annot.df, by="stool.ID")
```

```{r}
# summarize per diet 
species.uniq.plot.df.per.diet <- agg.species.uniq.bc.df %>% 
  dplyr::filter(age.approx.months <= 40) %>% # omit few 46 month samples
  mutate(Diet=factor(Diet.5mo.as.AL, levels=c("AL", "1D", "2D", "20", "40"))) %>%
  
  # compute mean and SD for each age and diet combination
  group_by(Diet, age.approx.months) %>% 
  summarise(n=n(),
            mean = mean(uniqueness),
            sd = sd(uniqueness),
            sem = sd(uniqueness)/sqrt(n()), .groups="drop") %>% 
  
  # make sure we have at least 3 observations at this date and age combination
  dplyr::filter(n >= 3)

# summarize across diets
species.uniq.plot.df.across.diets <- agg.species.uniq.bc.df %>% 
  dplyr::filter(age.approx.months <= 40) %>% # omit few 46 month samples
  
  # compute mean and SD across diets
  group_by(age.approx.months) %>% 
  summarise(n=n(),
            mean = mean(uniqueness),
            sd = sd(uniqueness),
            sem = sd(uniqueness)/sqrt(n()), .groups="drop") %>% 
  
  # make sure we have at least 3 observations at this date and age combination
  dplyr::filter(n >= 3)
```

```{r}
# plot
(species.uniq.v.age.mean.SEM.p <- species.uniq.plot.df.per.diet %>% 
    ggplot(aes(x=age.approx.months, y=mean, ymin=mean-sem, ymax=mean+sem,
               color=Diet)) +
    
    # per diet
    geom_path(alpha=0.3, linewidth=0.5) +
    geom_pointrange(alpha=0.3, size=0.1, shape=16) +
    
    # across diets
    geom_path(aes(color=NULL), linewidth=0.5, data=species.uniq.plot.df.across.diets) +
    geom_pointrange(aes(color=NULL), size=0.1, shape=16, data=species.uniq.plot.df.across.diets) +
    
    scale_color_manual(values=diet.palette) +
    labs(x="Age (months)", y="Uniqueness", color="Diet",
         title="Uniqueness") +
    geom_vline(xintercept=6, lty=2) +
    theme_classic(base_size=8))
```

```{r}
pdf("../plots/tax_species_uniq_v_age_mean_SEM.pdf", width=2, height=2)
species.uniq.v.age.mean.SEM.p + theme(legend.position="none")
dev.off()
```

# Effect of age

```{r}
age.df <- fixef.df %>% 
  dplyr::filter(fixef == "age.wks.scaled")
```

Adjust p-values.

```{r}
age.df <- age.df %>% 
  mutate(padj = p.adjust(Pr.chisq..con., method="BH")) %>% 
  mutate(is.sig = padj < PADJ.THRESH)
```

```{r}
age.df %>% 
  dplyr::count(is.sig)
```

```{r}
78 / (78 + 29)
65 / (65 + 42)
```

* Majority (73%) of taxonomic features are significantly associated with age

```{r}
age.df %>% 
  ggplot(aes(Pr.chisq..con.)) +
  geom_histogram(bins=20)
```

* p-value histogram makes sense

## Volcano plot

```{r}
age.genera.to.label <- c(
  "Bifidobacterium", "EnterococcusD", "Ligilactobacillus", 
  "Turicibacter", "Escherichia", "Dubosiella",
  "uniq", "Shannon", "Simpson",
  "ASF356", "CAG475", "UBA3700NOV",
  "UMGS268", "UBA7182", "CAG273"
)
```

```{r}
(age.effect.genera.volcano <- age.df %>% 
   mutate(neg.log10padj = case_when(padj == 0 ~ 16, TRUE ~ -log10(padj))) %>% 
   # mutate(label=case_when(is.sig == TRUE ~ feature, TRUE ~ "")) %>%
   mutate(label = case_when(
     feature %in% age.genera.to.label ~ feature,
     TRUE ~ ""
   )) %>%
   ggplot(aes(x=solution, y=neg.log10padj, color=is.sig, label=label)) +
   geom_hline(yintercept=2, lty=2) + # padj = 0.01
   geom_point() +
   geom_text_repel(nudge_x=0.01, min.segment.length=0, size=0.36*6, max.overlaps=Inf) +
   labs(x="<-- Decreases with age | Increases with age -->",
        y="-log10(padj)", color="padj < 0.01?") +
   scale_color_manual(values=c(`TRUE`=age.color, `FALSE`="gray80")) +
   theme_classic(base_size=8))
```

```{r}
pdf("../plots/asreml_age_effect_genera_volcano.pdf", height=2.5, width=2.5)
age.effect.genera.volcano + theme(legend.position="none")
dev.off()
```

# Effect of DR

```{r}
DR.df <- fixef.df %>% 
  dplyr::filter(fixef == "Diet.5mo.as.AL") %>% 
  dplyr::filter(fixef.value != "AL") # the AL row doesn't add anything

dim(DR.df)
n_distinct(DR.df$feature)
```

Adjust p-values. I think I could call `p.adjust` directly even though each p-value is repeated 4x, but to be safe, I'll compute adjusted p-values for just one diet and then merge them back in.

```{r}
DR.padj.df <- DR.df %>% 
  dplyr::filter(fixef.value == "40") %>% 
  mutate(padj = p.adjust(Pr.chisq..con., method="BH")) %>% 
  dplyr::select(feature, padj)

DR.df <- DR.df %>% 
  merge(DR.padj.df, by="feature") %>% 
  mutate(is.sig = padj < PADJ.THRESH)

dim(DR.df)
```

## Barplots

Order based on 40% CR.

```{r}
DR.effects.feature.order.df <- DR.df %>% 
  dplyr::filter(fixef.value == "40") %>%
  arrange(solution) %>% 
  mutate(feature.order = row_number()) %>% 
  dplyr::select(feature, feature.order)
```

Instead of labeling genera, I will draw a gray line at the right locations and manually do the labels.

```{r}
DR.genera.to.highlight <- c(
  
  # increased by DR
  "Lactobacillus", "Ligilactobacillus", "Limosilactobacillus",
  "uniq", "Simpson", "Shannon",
  
  # uniquely affected by CR or fasting, or vice versa
  "Roseburia", "Emergencia"
  
)

DR.assoc.genera.to.highlight.ii <- DR.effects.feature.order.df %>% 
  dplyr::filter(feature %in% DR.genera.to.highlight) %>% 
  pull(feature.order)
```

```{r}
(DR.effect.p <- DR.df %>% 
   
   # specify order based on 40% CR
   merge(DR.effects.feature.order.df, by="feature") %>%
   
   # reorder diet
   mutate(DR = factor(fixef.value, levels=c("1D", "2D", "20", "40"))) %>% 
   
   ggplot(aes(y=reorder(feature, feature.order), x=solution, 
              xmin=solution-std.error, xmax=solution+std.error,
              alpha=is.sig, fill=DR)) +
   geom_col() + 
   geom_linerange(linewidth=0.25) +
   geom_vline(xintercept=0, lty=2) +
   geom_hline(yintercept=DR.assoc.genera.to.highlight.ii, lty=2, color="gray50") +
   facet_wrap(~DR, nrow=1) +
   theme_void(base_size=8) +
   theme(legend.position="none") + 
   scale_fill_manual(values=diet.palette) +
   scale_alpha_manual(values=c(`TRUE`=1, `FALSE`=0.2))
)
```

```{r}
pdf("../plots/asreml_DR_effect_genera_barplot.pdf", height=3.5, width=3)
DR.effect.p
dev.off()
```


## Magnitude of effects

```{r}
(DR.mean.abs.effect.df <- DR.df %>% 
   dplyr::rename(DR = fixef.value) %>% 
   group_by(DR) %>% 
   summarise(mean = mean(abs(solution)),
             median = median(abs(solution)), .groups="drop") %>% 
   mutate(solution = sprintf("%.2f", mean)))
```

```{r}
pd <- position_dodge2(width=0.1)

(diet.effect.size.p <- DR.df %>% 
    mutate(DR=factor(fixef.value, levels=c("1D", "2D", "20", "40"))) %>% 
    ggplot(aes(x=DR, y=abs(solution))) +
    geom_path(aes(group=feature), alpha=0.1, position=pd) +
    geom_point(aes(color=DR), shape=16, position=pd) +
    geom_text(aes(y=-0.1, label=solution, color=DR), data=DR.mean.abs.effect.df) +
    stat_summary(aes(color=DR), fun="mean", geom="crossbar", width=0.5) +
    scale_color_manual(values=diet.palette) +
    stat_compare_means(method="t.test", label="p.signif", paired=T, comparisons=list(
      c("1D", "2D"), c("1D", "20"), c("1D", "40"), 
      c("2D", "20"), c("2D", "40"), c("20", "40"))) +
    labs(y="Absolute DR coefficient") +
    theme_classic(base_size=8) +
    theme(legend.position = "none"))
```

40% > 20% = 2D > 1D

```{r}
pdf("../plots/asreml_diet_overall_effect_genus_boxplots.pdf", height=2, width=2)
diet.effect.size.p
dev.off()
```

## Similarity of diet coefficients

It will be useful to first pivot wider.

```{r}
DR.coef.df <- DR.df %>%
  pivot_wider(id_cols=feature, names_from=fixef.value, values_from=solution)
```

```{r}
DR.cor.df <- DR.coef.df %>% 
  dplyr::select(-feature) %>% 
  correlate() %>% 
  shave() %>%
  stretch(na.rm=T) %>% 
  mutate(x = factor(x, levels=c("1D", "2D", "20", "40")),
         y = factor(y, levels=c("1D", "2D", "20", "40")),
         r.chr = sprintf("%.2f", r)) 
```

```{r}
# make sure we use linear fit
my_fn <- function(data, mapping, ...){
  p <- ggplot(data = data, mapping = mapping) + 
    geom_point(size=0.5) + 
    geom_smooth(method="lm", color="mediumorchid", ...)
  p
}

(cor.DR.effects.p <- DR.coef.df %>% 
  dplyr::select(`1D`, `2D`, `20`, `40`) %>%  
  ggpairs(lower = list(continuous = my_fn), 
          diag = NULL,
          progress = F) +
  theme_classic(base_size=8) +
  labs(x="DR coef", y="DR coef"))
```

```{r}
pdf("../plots/asreml_diet_similarity_genus_scatter.pdf", height=4, width=4)
cor.DR.effects.p
dev.off()
```

## Affected by CR but not fasting, and vice versa

```{r}
genus.CR.or.fasting.mean.coef.df <- DR.df %>%
  mutate(CR.or.fasting = ifelse(fixef.value %in% c("20", "40"), "CR", "fasting")) %>% 
  group_by(feature, CR.or.fasting) %>% 
  summarise(solution = mean(solution), .groups="drop") %>% 
  group_by(feature) %>% 
  mutate(CR.and.fasting.in.diff.dir = sign(solution[CR.or.fasting == "CR"]) != sign(solution[CR.or.fasting == "fasting"])) %>% 
  ungroup()

genus.CR.or.fasting.mean.coef.order.df <- genus.CR.or.fasting.mean.coef.df %>% 
  dplyr::filter(CR.or.fasting == "CR") %>% 
  arrange(solution) %>%
  mutate(ii = row_number()) %>% 
  dplyr::select(feature, ii)
  
(genus.CR.v.fasting.mean.coef.p <- genus.CR.or.fasting.mean.coef.df %>% 
    merge(genus.CR.or.fasting.mean.coef.order.df, by="feature") %>% 
    mutate(label = ifelse(feature %in% c("Emergencia", "Roseburia"), feature, "")) %>% 
    ggplot(aes(x=reorder(feature, ii), y=solution, color=CR.or.fasting, alpha=CR.and.fasting.in.diff.dir)) +
    geom_hline(yintercept=0, lty=2) +
    geom_line(aes(group=feature, color=NULL)) +
    geom_point(shape=16, size=0.5) +
    geom_text(aes(label=label)) +
    scale_alpha_manual(values=c(`TRUE`=1, `FALSE`=0.2)) +
    scale_color_manual(values=CR.v.fasting.palette) +
    theme_classic(base_size=8) +
    theme(axis.text.x=element_blank(),
          axis.ticks.x=element_blank(),
          legend.position="none") +
    labs(x="Genus", y="Coefficient"))
```

```{r}
pdf("../plots/genus_CR_v_fasting_mean_coef.pdf", height=2, width=3)
genus.CR.v.fasting.mean.coef.p
dev.off()
```

# Heritability

Adjust p-values.

```{r}
herit.df <- herit.df %>% 
  mutate(padj = p.adjust(LRT.pval, method="BH")) %>% 
  mutate(is.sig = padj < PADJ.THRESH)
```

```{r}
herit.df %>% 
  dplyr::filter(feature %in% c("uniq", "PCoA1", "PCoA2", "PCoA3", "Shannon", "Simpson", "Chao1"))
```

* For the community features, only the PCoA coordinates are significantly heritable

```{r}
herit.df %>% 
  dplyr::filter(!(feature %in% c("uniq", "PCoA1", "PCoA2", "PCoA3", "Shannon", "Simpson", "Chao1"))) %>%
  dplyr::count(is.sig) %>%
  pull(n) %>% prop.table()

herit.df %>% 
  dplyr::count(is.sig) %>%
  pull(n) %>% prop.table()
```

* 66% of genera are heritable
* 64.5% of features are heritable (including the community features)

```{r}
herit.df %>%
  dplyr::filter(!(feature %in% c("uniq", "PCoA1", "PCoA2", "PCoA3", "Shannon", "Simpson", "Chao1"))) %>%
  dplyr::filter(is.sig) %>%
  summarise(mean = mean(herit)) %>%
  pull(mean)

(mean.herit.for.heritable.feats <- herit.df %>% 
   dplyr::filter(is.sig) %>% 
   summarise(mean = mean(herit)) %>% 
   pull(mean))
```

* Mean heritability (whether or not we keep community features is 11.1%)

Volcano plot.

```{r}
herit.genera.to.label <- c(
  "Lactobacillus", "Parvibacter", "ErysipelatoclostridiaceaeNOV",
  "ASF356", "Ileibacterium", "EnterococcusD")
```

```{r}
(genus.herit.scatter.p <- herit.df %>% 
   mutate(label = case_when(
     feature %in% herit.genera.to.label ~ feature,
     TRUE ~ ""
   )) %>%
   ggplot(aes(x=herit, y=-log10(padj), color=is.sig, label=label)) +
   geom_hline(yintercept=2, lty=2) + # padj = 0.01
   geom_vline(xintercept = mean.herit.for.heritable.feats, 
              color = genetics.color, lty=2) +
   geom_point(shape=16) +
   geom_text_repel(nudge_x=0.01, min.segment.length=0, size=0.36*6, max.overlaps=Inf) +
   labs(x="Heritability", fill="padj < 0.01?") +
   scale_color_manual(values=c(`TRUE`=genetics.color, `FALSE`="gray80")) +
   theme_classic(base_size=8) +
   theme(legend.position="none"))
```

```{r}
# pdf("../plots/genus_heritability_scatter.pdf", height=2.5, width=2.5)
# genus.herit.scatter.p
# dev.off()
```


# All random effects

```{r}
ranef.df.all.ranef.model %>% 
  group_by(term.clean) %>% 
  summarise(PVE = mean(PVE)) %>% 
  arrange(desc(PVE))
```

* 63% of variance remains unexplained
* Genetics explain mean 7.5% of variance

Adjust p-values within each experimental variable.

```{r}
ranef.df.all.ranef.model <- ranef.df.all.ranef.model %>% 
  group_by(term.clean) %>% 
  mutate(padj = p.adjust(LRT.pval, method="BH")) %>% 
  mutate(is.sig = padj < PADJ.THRESH)
```

```{r}
(genus.all.ranef.PVE.p <- ranef.df.all.ranef.model %>% 
   dplyr::filter(term.clean != "Residual") %>% 
   arrange(is.sig) %>% 
   ggplot(aes(x=reorder(term.clean, -PVE), y=PVE, color=is.sig)) +
   geom_jitter(width=0.2, shape=16) +
   stat_summary(aes(color=NULL), fun="mean", geom="crossbar", width=0.75) +
   labs(x="Experimental variable", y="Proportion of variance explained") +
   theme_classic(base_size=8) +
   theme(legend.position="none") +
   scale_color_manual(values=c(`TRUE`="gray20", `FALSE`="gray80")))
```

```{r}
pdf("../plots/PVE_all_ranef_genera.pdf", height=2, width=2)
genus.all.ranef.PVE.p
dev.off()
```

# Example plots

## Versus age (mean ± SEM)

```{r}
plot_one_genus_versus_age <- function(this.genus, type.of.error.bar="sem") {
  
  # summarize per diet 
  plot.df.per.diet <- agg.physeq.genus %>% 
    transform_sample_counts(function(x) 100*x/sum(x)) %>% 
    psmelt %>% 
    dplyr::filter(age.approx.months <= 40) %>% 
    dplyr::filter(genus == this.genus) %>% 
    mutate(Diet=factor(Diet.5mo.as.AL, levels=c("AL", "1D", "2D", "20", "40"))) %>%
    
    # compute mean, SD, and SEM for each age and diet combination
    group_by(Diet, age.approx.months) %>% 
    summarise(n=n(),
              mean = mean(Abundance),
              sd = sd(Abundance),
              sem = sd(Abundance)/sqrt(n()), .groups="drop") %>% 
    
    # make sure we have at least 3 observations at this date and age combination
    dplyr::filter(n >= 3)
  
  # summarize across diets
  plot.df.across.diets <- agg.physeq.genus %>% 
    transform_sample_counts(function(x) 100*x/sum(x)) %>% 
    psmelt %>% 
    dplyr::filter(age.approx.months <= 40) %>% 
    dplyr::filter(genus == this.genus) %>% 
    
    # compute mean, SD, and SEM across diets
    group_by(age.approx.months) %>% 
    summarise(n=n(),
              mean = mean(Abundance),
              sd = sd(Abundance),
              sem = sd(Abundance)/sqrt(n()), .groups="drop") %>% 
    
    # make sure we have at least 3 observations at this date and age combination
    dplyr::filter(n >= 3)
  
  # indicate what to use for error bars
  plot.df.per.diet <- plot.df.per.diet %>% 
    mutate(error = case_when(
      type.of.error.bar == "2sem" ~ 2*sem,
      type.of.error.bar == "sem" ~ sem,
      type.of.error.bar == "sd" ~ sd)) 
  
  plot.df.across.diets <- plot.df.across.diets %>% 
    mutate(error = case_when(
      type.of.error.bar == "2sem" ~ 2*sem,
      type.of.error.bar == "sem" ~ sem,
      type.of.error.bar == "sd" ~ sd)) 
  
  # plot
  plot.df.per.diet %>%
    ggplot(aes(x=age.approx.months, y=mean, ymin=mean-error, ymax=mean+error,
               color=Diet)) +
    
    # per diet
    geom_path(alpha=0.3, linewidth=0.5) +
    geom_pointrange(alpha=0.3, size=0.01, shape=16) +
    
    # across diets
    geom_path(aes(color=NULL), linewidth=0.5, data=plot.df.across.diets) +
    geom_pointrange(aes(color=NULL), shape=16, size=0.01, data=plot.df.across.diets) +
  
    scale_color_manual(values=diet.palette) +
    labs(x="Age (months)", y="Relative abundance (%)", color="Diet",
         title=sprintf("%s", this.genus)) +
    geom_vline(xintercept=6, lty=2) +
    theme_classic(base_size=8)
}
```

### Bifidobacterium

```{r}
pdf("../plots/bifido_versus_age_mean_SEM.pdf", height=2, width=2)
plot_one_genus_versus_age("Bifidobacterium", type.of.error.bar="sem") + theme(legend.position="none")
dev.off()
```

### Ligilactobacillus

```{r}
pdf("../plots/ligilacto_versus_age_mean_SEM.pdf", height=2, width=2)
plot_one_genus_versus_age("Ligilactobacillus", type.of.error.bar="sem") + theme(legend.position="none")
dev.off()
```

### UBA11957

```{r}
pdf("../plots/UBA11957_versus_age_mean_SEM.pdf", height=2, width=2)
plot_one_genus_versus_age("UBA11957", type.of.error.bar="sem") + theme(legend.position="none")
dev.off()
```

## Versus diet

### UMGS1815

```{r}
(UMGS1815.v.diet.p <- agg.physeq.genus %>% 
   transform_sample_counts(function(x) x/sum(x)) %>% 
   psmelt %>% 
   dplyr::filter(age.approx.months <= 40) %>% 
   mutate(Diet.5mo.as.AL=factor(Diet.5mo.as.AL, levels=c("AL", "1D", "2D", "20", "40"))) %>% 
   dplyr::filter(genus == "UMGS1815") %>% 
   ggplot(aes(x=Diet.5mo.as.AL, y=Abundance, fill=Diet.5mo.as.AL)) +
   geom_boxplot(outlier.shape=16, outlier.size=0.5) +
   stat_compare_means(ref.group="AL", label="p.signif", method="t.test") +
   labs(x="", y="Relative abundance (%)", title="UMGS1815") +
   scale_fill_manual(values=diet.palette) +
   theme_classic(base_size=8) +
   theme(legend.position="none"))
```

```{r}
pdf("../plots/UMGS1815_v_diet_boxplots.pdf", width=2, height=2)
UMGS1815.v.diet.p
dev.off()
```

### Ligilactobacillus

```{r}
(ligilacto.v.diet.p <- agg.physeq.genus %>% 
   transform_sample_counts(function(x) x/sum(x)) %>% 
   psmelt %>% 
   dplyr::filter(age.approx.months <= 40) %>% 
   mutate(Diet.5mo.as.AL=factor(Diet.5mo.as.AL, levels=c("AL", "1D", "2D", "20", "40"))) %>% 
   dplyr::filter(genus == "Ligilactobacillus") %>% 
   ggplot(aes(x=Diet.5mo.as.AL, y=Abundance, fill=Diet.5mo.as.AL)) +
   geom_boxplot(outlier.shape=16, outlier.size=0.5) +
   stat_compare_means(ref.group="AL", label="p.signif", method="t.test") +
   labs(x="", y="Relative abundance (%)", title="Ligilactobacillus") +
   scale_fill_manual(values=diet.palette) +
   theme_classic(base_size=8) +
   theme(legend.position="none"))
```

```{r}
pdf("../plots/ligilactobacillus_v_diet_boxplots.pdf", width=2, height=2)
ligilacto.v.diet.p
dev.off()
```

### Emergencia

```{r}
median.emergencia.relab.in.AL <- agg.physeq.genus %>% 
  transform_sample_counts(function(x) 100*x/sum(x)) %>% 
  psmelt %>% 
  dplyr::filter(age.approx.months <= 40) %>% 
  dplyr::filter(genus == "Emergencia") %>% 
  dplyr::filter(Diet.5mo.as.AL == "AL") %>% 
  mutate(log10relab=log10(Abundance)) %>%
  pull(log10relab) %>% median()

(emergencia.v.diet.p <- agg.physeq.genus %>% 
    transform_sample_counts(function(x) 100*x/sum(x)) %>% 
    psmelt %>% 
    dplyr::filter(age.approx.months <= 40) %>% 
    dplyr::filter(genus == "Emergencia") %>% 
    mutate(log10relab=log10(Abundance)) %>%
    mutate(Diet.5mo.as.AL=factor(Diet.5mo.as.AL, levels=c("AL", "1D", "2D", "20", "40"))) %>%
    ggplot(aes(x=Diet.5mo.as.AL, y=log10relab, fill=Diet.5mo.as.AL)) +
    geom_hline(yintercept=median.emergencia.relab.in.AL, lty=2, color="gray50",
               linewidth=0.5) +
    geom_boxplot(outlier.shape=16, outlier.size=0.5, linewidth=0.25) +
    stat_compare_means(ref.group="AL", label="p.signif", method="t.test") +
    labs(x="Diet", y="log10(relative abundance)", title="Emergencia") +
    scale_fill_manual(values=diet.palette) +
    theme_classic(base_size=8) +
    theme(legend.position="none"))
```

```{r}
pdf("../plots/emergencia_v_diet_boxplots.pdf", width=1.5, height=1.5)
emergencia.v.diet.p
dev.off()
```

### Roseburia

```{r}
median.roseburia.relab.in.AL <- agg.physeq.genus %>% 
  transform_sample_counts(function(x) 100*x/sum(x)) %>% 
  psmelt %>% 
  dplyr::filter(age.approx.months <= 40) %>% 
  dplyr::filter(genus == "Roseburia") %>% 
  dplyr::filter(Diet.5mo.as.AL == "AL") %>% 
  mutate(log10relab=log10(Abundance)) %>%
  pull(log10relab) %>% median()

(roseburia.v.diet.p <- agg.physeq.genus %>% 
    transform_sample_counts(function(x) 100*x/sum(x)) %>% 
    psmelt %>% 
    dplyr::filter(age.approx.months <= 40) %>% 
    dplyr::filter(genus == "Roseburia") %>%
    mutate(log10relab=log10(Abundance)) %>%
    mutate(Diet.5mo.as.AL=factor(Diet.5mo.as.AL, levels=c("AL", "1D", "2D", "20", "40"))) %>% 
    ggplot(aes(x=Diet.5mo.as.AL, y=log10relab, fill=Diet.5mo.as.AL)) +
    geom_hline(yintercept=median.roseburia.relab.in.AL, lty=2, color="gray50",
               linewidth=0.5) +
    geom_boxplot(outlier.shape=16, outlier.size=0.5, linewidth=0.25) +
    stat_compare_means(ref.group="AL", label="p.signif", method="t.test") +
    labs(x="Diet", y="log10(relative abundance)", title="Roseburia") +
    scale_fill_manual(values=diet.palette) +
    theme_classic(base_size=8) +
    theme(legend.position="none"))
```

```{r}
pdf("../plots/roseburia_v_diet_boxplots.pdf", width=1.5, height=1.5)
roseburia.v.diet.p
dev.off()
```
