Will compare MetaPhlAn versus Kraken taxonomic classification and LMM outputs.

1) MetaPhlAn = MetaPhlAn4, genus-level log2(relabs)
2) Kraken = Kraken2, genus-level CLR, MGBC as reference

* % unclassified
* Mean difference in relative abundance across all samples
* Age coefficients, DR coefficients, and heritability estimates

**2024-10-19 update: added comparison of LMM outputs**

# Load libraries

```{r, warning=F, message=F}
library(tidyverse)
library(phyloseq)
library(speedyseq) # faster tax_glom
library(foreach)
library(ggrepel)
library(patchwork)

PADJ.THRESH <- 0.01
```

# Import metadata

```{r}
seq.meta.df <- read.table(
  "../data/metadata/sequencing_metadata_after_QC_n3473_230620.txt", 
  sep="\t", header=T)
lib.meta.df <- read.table(
  "../data/metadata/library_metadata_after_QC_n3213_230620.txt",
  sep="\t", header=T)
stool.meta.df <- read.table(
  "../data/metadata/stool_metadata_after_QC_n3124_240418.txt", 
  sep="\t", header=T)
mouse.meta.df <- read.csv("../data/metadata/AnimalData_Processed_20230712.csv")
```

```{r}
seq.meta.annot.df <- seq.meta.df %>%
  merge(lib.meta.df, by="lib.ID") %>%
  merge(stool.meta.df, by="stool.ID") %>% 
  merge(mouse.meta.df, by.x="mouse.ID", by.y="MouseID", all.x=T)

stool.meta.annot.df <- stool.meta.df %>%
  merge(mouse.meta.df, by.x="mouse.ID", by.y="MouseID")
```

# Import data

## MetaPhlAn

Need MetaPhlAn output prior to aggregation in order to get % unclassified, but we'll use aggregated data for relative abundance comparison.

```{r}
mpa.df <- read.table(
  "../data/metaphlan.txt",
  sep="\t", header=T)

mpa.genus.df <- read.table(
  "../data/metaphlan_genus_agg_by_stool_ID_n376x2997.txt",
  sep="\t", header=T) %>% 
  column_to_rownames("genus")

dim(mpa.genus.df)
```

* 376 genera x 2997 samples

## Kraken

% unclassified stored separately.

```{r}
kraken.perc.unclass.df <- read.table(
  "../results/percent_kraken_unclassified_n4352_221102.txt",
  sep="\t", header=T)
```

Use phyloseq to create genus-level Kraken relative abundances.

```{r}
# import. data
agg.kraken.df <- read.table(
  "../data/kraken_matrix_agg_by_stool_ID_n1303x2997.txt", 
  sep="\t", header=T, row.names=1)

# import taxonomy
kraken.tax.df <- read.table(
  "../data/kraken_taxonomy_n1303.txt", 
  sep="\t", header=T, quote="", row.names=1)

# create phyloseq object
agg.physeq <- phyloseq(
  agg.kraken.df %>% as.matrix %>% otu_table(taxa_are_rows=T),
  stool.meta.annot.df %>% column_to_rownames("stool.ID") %>% sample_data,
  kraken.tax.df %>% as.matrix %>% tax_table)

# aggregate to genera
agg.physeq.genus <- agg.physeq %>% tax_glom(taxrank="genus")

# convert to relative abundances
agg.physeq.genus.relab <- agg.physeq.genus %>% 
  transform_sample_counts(function(x) 100*x/sum(x))

# convert to dataframe
kraken.genus.df <- agg.physeq.genus.relab %>% 
  otu_table() %>% data.frame()
rownames(kraken.genus.df) <- data.frame(tax_table(agg.physeq.genus.relab))$genus

dim(kraken.genus.df)
```

* 252 genera x 2997 samples

# Import LMM outputs

ASReml, time as a random effect.

## Fixed effects

```{r}
mpa.fixef.df <- read.table(
  "../results/asreml_metaphlan_genus_w_time_ranef/fixef.txt",
  sep="\t", header=T)

kraken.fixef.df <- read.table(
  "../results/asreml_kraken_genus_w_time_ranef/fixef.txt",
  sep="\t", header=T)

dim(mpa.fixef.df)
dim(kraken.fixef.df)
```

## Heritability

```{r}
mpa.herit.df <- read.table(
  "../results/asreml_metaphlan_genus_w_time_ranef/herit.txt",
  sep="\t", header=T)

kraken.herit.df <- read.table(
  "../results/asreml_kraken_genus_w_time_ranef/herit.txt",
  sep="\t", header=T)

dim(mpa.herit.df)
dim(kraken.herit.df)
```

## Rename "uniq" to "uniqueness"

In the Kraken results, uniqueness is "uniq". I will rename to "Uniqueness" to match the MetaPhlAn results.

```{r}
kraken.fixef.df <- kraken.fixef.df %>% 
  mutate(feature = case_when(
    feature == "uniq" ~ "Uniqueness",
    TRUE ~ feature))

kraken.herit.df <- kraken.herit.df %>% 
  mutate(feature = case_when(
    feature == "uniq" ~ "Uniqueness",
    TRUE ~ feature))
```

## Merge

How many features in common?

```{r}
common.feats <- intersect(mpa.fixef.df$feature, kraken.fixef.df$feature)
length(common.feats)
```

* 24, including 3 community features, so 21

```{r}
common.feats
```

```{r}
kraken.v.mpa.fixef.df <- inner_join(
  kraken.fixef.df %>% dplyr::filter(fixef != "(Intercept"), # ignore intercept
  mpa.fixef.df %>% dplyr::filter(fixef != "(Intercept"), # ignore intercept 
  by=join_by(feature, fixef, fixef.value),
  suffix=c(".kraken", ".mpa"))

dim(kraken.v.mpa.fixef.df)
```

```{r}
kraken.v.mpa.herit.df <- inner_join(
  kraken.herit.df,
  mpa.herit.df,
  by=join_by(feature),
  suffix=c(".kraken", ".mpa"))

dim(kraken.v.mpa.herit.df)
```

# Compare % unclassified

```{r}
perc.unclass.df <- rbind(
  
  # mpa
  mpa.df %>% 
    dplyr::filter(clade_name == "UNCLASSIFIED") %>% 
    dplyr::select(-clade_name) %>% 
    t() %>% data.frame() %>% setNames("perc.unclass") %>% 
    rownames_to_column("seq.ID") %>% 
    mutate(method = "MetaPhlAn4"),
  
  # kraken
  kraken.perc.unclass.df %>% 
    dplyr::rename(perc.unclass = perc.unclass.kraken) %>% 
    mutate(method = "Kraken2")
)
```

```{r}
perc.unclass.df %>% 
  dplyr::count(method)
```

* 4352 Kraken results but only 3473 MetaPhlAn results -- we'll subset to these 3473
* We should also subset to non-control samples

```{r}
seq.IDs.to.keep <- seq.meta.annot.df %>% 
  dplyr::filter(sample.type == "DO") %>% 
  dplyr::filter(seq.ID %in% colnames(mpa.df)) %>% 
  pull(seq.ID)
length(seq.IDs.to.keep)

perc.unclass.filt.df <- perc.unclass.df %>% 
  dplyr::filter(seq.ID %in% seq.IDs.to.keep)
dim(perc.unclass.filt.df)
```

```{r}
(mean.perc.unclass.df <- perc.unclass.filt.df %>% 
   group_by(method) %>% 
   summarise(perc.unclass = mean(perc.unclass), .groups="drop") %>% 
   mutate(perc.class.chr = sprintf("%.1f%%", 100-perc.unclass)))
```

* Mean % classified: 95.7% for Kraken2, 88.0% for MetaPhlAn4

```{r}
(perc.classifed.p <- perc.unclass.filt.df %>% 
   ggplot(aes(x=method, y=100-perc.unclass)) +
   stat_summary(fun="mean", geom="col", width=0.5) +
   geom_point(shape=16) +
   geom_path(aes(group=seq.ID), alpha=0.01) +
   geom_text(aes(y=5, label=perc.class.chr), 
             color="white", data=mean.perc.unclass.df) +
   labs(x="", y="% classified",
        title="DO samples prior to stool.ID aggregation") +
   theme_classic(base_size=8))
```

```{r}
pdf("../plots/percent_classified_kraken_mgbc_v_mpa4.pdf", width=1.5, height=1.75)
perc.classifed.p
dev.off()
```

# Compare per-sample relabs

```{r}
mpa.v.kraken.genus.relab.df <- merge(
  
  # mpa
  mpa.genus.df %>% 
    rownames_to_column("genus") %>% 
    pivot_longer(-genus, names_to="seq.ID", values_to="mpa"),
  
  # kraken
  kraken.genus.df %>% 
    rownames_to_column("genus") %>% 
    pivot_longer(-genus, names_to="seq.ID", values_to="kraken"),
  
  by=c("genus", "seq.ID")
)
```

```{r}
n_distinct(mpa.v.kraken.genus.relab.df$genus)
```

* 41 genera in common

```{r}
mpa.v.kraken.mean.relab.diff.df <- mpa.v.kraken.genus.relab.df %>% 
  mutate(kraken.minus.mpa = kraken - mpa) %>% 
  group_by(genus) %>% 
  summarise(mean = mean(kraken.minus.mpa),
            sd = sd(kraken.minus.mpa),
            se = sd(kraken.minus.mpa) / sqrt(n()),
            .groups="drop")
```

```{r}
(mpa.v.kraken.relab.diff.p <- mpa.v.kraken.mean.relab.diff.df %>% 
   ggplot(aes(x=mean, xmin=mean-sd, xmax=mean+sd, y=reorder(genus, mean))) +
   geom_pointrange(size=0.1) +
   geom_vline(xintercept=0, lty=2) +
   labs(x="Kraken minus MetaPhlAn relab (mean ± standard deviation)", y="Genus") +
   theme_classic(base_size=8))
```

```{r}
pdf("../plots/mpa_v_kraken_relab_diffs.pdf", height=4, width=3)
mpa.v.kraken.relab.diff.p
dev.off()
```

# Age coefficients

```{r}
kraken.v.mpa.age.df <- kraken.v.mpa.fixef.df %>% 
  dplyr::filter(fixef == "age.wks.scaled")
```

Adjust p-values.

```{r}
kraken.v.mpa.age.df <- kraken.v.mpa.age.df %>% 
  mutate(padj.kraken = p.adjust(Pr.chisq..con..kraken, method="BH"),
         padj.mpa = p.adjust(Pr.chisq..con..mpa, method="BH")) %>% 
  mutate(is.sig.kraken = padj.kraken < PADJ.THRESH,
         is.sig.mpa = padj.mpa < PADJ.THRESH)
```

Indicate if significant in both, neither, or just one.

```{r}
kraken.v.mpa.age.df <- kraken.v.mpa.age.df %>% 
  mutate(is.sig = case_when(
    is.sig.kraken & is.sig.mpa ~ "Both",
    is.sig.kraken & !is.sig.mpa ~ "Only Kraken",
    !is.sig.kraken & is.sig.mpa ~ "Only MetaPhlAn",
    !is.sig.kraken & !is.sig.mpa ~ "Neither"
  ))
```

```{r}
kraken.v.mpa.age.coef.rho <- cor(kraken.v.mpa.age.df$solution.kraken, kraken.v.mpa.age.df$solution.mpa, method="spearman")
kraken.v.mpa.age.coef.rho.pval <- cor.test(kraken.v.mpa.age.df$solution.kraken, kraken.v.mpa.age.df$solution.mpa, method="spearman")$p.value
```

```{r}
(kraken.v.mpa.age.coef.p <- kraken.v.mpa.age.df %>% 
   ggplot(aes(x=solution.kraken, y=solution.mpa, label=feature, color=is.sig)) +
   geom_point(shape=16) +
   geom_abline(slope=1, lty=2) +
   geom_text_repel(size=0.36*6, max.overlaps=Inf) +
   labs(x="Age coefficient, Kraken", y="Age coefficient, MetaPhlAn",
        color="Significant in?",
        title=sprintf("rho = %.2f, p-value = %2.2e", kraken.v.mpa.age.coef.rho, kraken.v.mpa.age.coef.rho.pval)) +
   scale_color_manual(values=c(
     `Only Kraken`="lightskyblue", `Only MetaPhlAn`="lightskyblue3",
     Both="navy", Neither="gray80")) +
   theme_classic(base_size=8))
```

```{r}
pdf("../plots/kraken_v_metaphlan_age_coefficients.pdf", height=2.5, width=2.5)
kraken.v.mpa.age.coef.p + theme(legend.position="none")
dev.off()
```

# Diet coefficients

```{r}
kraken.v.mpa.DR.df <- kraken.v.mpa.fixef.df %>% 
  dplyr::filter(fixef == "Diet.5mo.as.AL") %>% 
  dplyr::filter(fixef.value != "AL") # the AL row doesn't add anything
```

Adjust p-values. I think I could call `p.adjust` directly even though each p-value is repeated 4x, but to be safe, I'll compute adjusted p-values for just one diet and then merge them back in.

```{r}
kraken.v.mpa.DR.padj.df <- kraken.v.mpa.DR.df %>% 
  dplyr::filter(fixef.value == "40") %>% 
  mutate(padj.kraken = p.adjust(Pr.chisq..con..kraken, method="BH"),
         padj.mpa = p.adjust(Pr.chisq..con..mpa, method="BH")) %>% 
  mutate(is.sig.kraken = padj.kraken < PADJ.THRESH,
         is.sig.mpa = padj.mpa < PADJ.THRESH) %>% 
  dplyr::select(feature, padj.kraken, padj.mpa, is.sig.kraken, is.sig.mpa)

kraken.v.mpa.DR.df <- kraken.v.mpa.DR.df %>% 
  merge(kraken.v.mpa.DR.padj.df, by="feature")

dim(kraken.v.mpa.DR.df)
```

Indicate if significant in both, neither, or just one.

```{r}
kraken.v.mpa.DR.df <- kraken.v.mpa.DR.df %>% 
  mutate(is.sig = case_when(
    is.sig.kraken & is.sig.mpa & (sign(solution.kraken) == sign(solution.mpa)) ~ "Both, same sign",
    is.sig.kraken & is.sig.mpa & (sign(solution.kraken) != sign(solution.mpa)) ~ "Both, different sign",
    is.sig.kraken & !is.sig.mpa ~ "Only Kraken",
    !is.sig.kraken & is.sig.mpa ~ "Only MetaPhlAn",
    !is.sig.kraken & !is.sig.mpa ~ "Neither"
  ))
```

Calculate correlation for each diet separately.

```{r}
kraken.v.mpa.CR20.df <- kraken.v.mpa.DR.df %>% dplyr::filter(fixef.value == "20")
kraken.v.mpa.CR40.df <- kraken.v.mpa.DR.df %>% dplyr::filter(fixef.value == "40")
kraken.v.mpa.1D.df <- kraken.v.mpa.DR.df %>% dplyr::filter(fixef.value == "1D")
kraken.v.mpa.2D.df <- kraken.v.mpa.DR.df %>% dplyr::filter(fixef.value == "2D")
```

```{r}
kraken.v.mpa.CR20.coef.rho <- cor(kraken.v.mpa.CR20.df$solution.kraken, kraken.v.mpa.CR20.df$solution.mpa, method="spearman")
kraken.v.mpa.CR20.coef.rho.pval <- cor.test(kraken.v.mpa.CR20.df$solution.kraken, kraken.v.mpa.CR20.df$solution.mpa, method="spearman")$p.value

kraken.v.mpa.CR40.coef.rho <- cor(kraken.v.mpa.CR40.df$solution.kraken, kraken.v.mpa.CR40.df$solution.mpa, method="spearman")
kraken.v.mpa.CR40.coef.rho.pval <- cor.test(kraken.v.mpa.CR40.df$solution.kraken, kraken.v.mpa.CR40.df$solution.mpa, method="spearman")$p.value

kraken.v.mpa.1D.coef.rho <- cor(kraken.v.mpa.1D.df$solution.kraken, kraken.v.mpa.1D.df$solution.mpa, method="spearman")
kraken.v.mpa.1D.coef.rho.pval <- cor.test(kraken.v.mpa.1D.df$solution.kraken, kraken.v.mpa.1D.df$solution.mpa, method="spearman")$p.value

kraken.v.mpa.2D.coef.rho <- cor(kraken.v.mpa.2D.df$solution.kraken, kraken.v.mpa.2D.df$solution.mpa, method="spearman")
kraken.v.mpa.2D.coef.rho.pval <- cor.test(kraken.v.mpa.2D.df$solution.kraken, kraken.v.mpa.2D.df$solution.mpa, method="spearman")$p.value
```

Pick which features to label.

```{r}
feats.to.label <- c(
  "Lactobacillus", "Uniqueness", "Shannon", "Simpson", "Ligilactobacillus",
  "Limosilactobacillus", "Muribaculum", "Schaederella", "Turicibacter",
  "Acetatifactor", "Dorea", "Paramuribaculum", "Anaerotruncus")
```

```{r}
kraken.v.mpa.CR20.coef.p <- kraken.v.mpa.CR20.df %>% 
  ggplot(aes(x=solution.kraken, y=solution.mpa, label=feature, color=is.sig)) +
  geom_point(shape=16) +
  geom_abline(slope=1, lty=2) +
  geom_text_repel(size=0.36*5, max.overlaps=Inf) +
  labs(x="Kraken coefficient", y="MetaPhlAn coefficient",
       color="Significant in?",
       title=sprintf("20%% CR, rho = %.2f, p-value = %2.2e",
                     kraken.v.mpa.CR20.coef.rho, kraken.v.mpa.CR20.coef.rho.pval)) +
  scale_color_manual(values=c(
    `Only Kraken`="lightskyblue", `Only MetaPhlAn`="lightskyblue3",
    `Both, same sign`="navy", Neither="gray80", `Both, different sign`="firebrick")) +
  theme_classic(base_size=8)

kraken.v.mpa.CR40.coef.p <- kraken.v.mpa.CR40.df %>% 
  ggplot(aes(x=solution.kraken, y=solution.mpa, label=feature, color=is.sig)) +
  geom_point(shape=16) +
  geom_abline(slope=1, lty=2) +
  geom_text_repel(size=0.36*5, max.overlaps=Inf) +
  labs(x="Kraken coefficient", y="MetaPhlAn coefficient",
       color="Significant in?",
       title=sprintf("40%% CR, rho = %.2f, p-value = %2.2e",
                     kraken.v.mpa.CR40.coef.rho, kraken.v.mpa.CR40.coef.rho.pval)) +
  scale_color_manual(values=c(
    `Only Kraken`="lightskyblue", `Only MetaPhlAn`="lightskyblue3",
    `Both, same sign`="navy", Neither="gray80", `Both, different sign`="firebrick")) +
  theme_classic(base_size=8)

kraken.v.mpa.1D.coef.p <- kraken.v.mpa.1D.df %>% 
  ggplot(aes(x=solution.kraken, y=solution.mpa, label=feature, color=is.sig)) +
  geom_point(shape=16) +
  geom_abline(slope=1, lty=2) +
  geom_text_repel(size=0.36*5, max.overlaps=Inf) +
  labs(x="Kraken coefficient", y="MetaPhlAn coefficient",
       color="Significant in?",
       title=sprintf("1D, rho = %.2f, p-value = %2.2e",
                     kraken.v.mpa.1D.coef.rho, kraken.v.mpa.1D.coef.rho.pval)) +
  scale_color_manual(values=c(
    `Only Kraken`="lightskyblue", `Only MetaPhlAn`="lightskyblue3",
    `Both, same sign`="navy", Neither="gray80", `Both, different sign`="firebrick")) +
  theme_classic(base_size=8)

kraken.v.mpa.2D.coef.p <- kraken.v.mpa.2D.df %>% 
  ggplot(aes(x=solution.kraken, y=solution.mpa, label=feature, color=is.sig)) +
  geom_point(shape=16) +
  geom_abline(slope=1, lty=2) +
  geom_text_repel(size=0.36*5, max.overlaps=Inf) +
  labs(x="Kraken coefficient", y="MetaPhlAn coefficient",
       color="Significant in?",
       title=sprintf("2D, rho = %.2f, p-value = %2.2e",
                     kraken.v.mpa.2D.coef.rho, kraken.v.mpa.2D.coef.rho.pval)) +
  scale_color_manual(values=c(
    `Only Kraken`="lightskyblue", `Only MetaPhlAn`="lightskyblue3",
    `Both, same sign`="navy", Neither="gray80", `Both, different sign`="firebrick")) +
  theme_classic(base_size=8)
```

```{r}
(kraken.v.mpa.DR.coef.p <- (kraken.v.mpa.CR20.coef.p + theme(legend.position="none")) +
  (kraken.v.mpa.CR40.coef.p + theme(legend.position="none")) +
  (kraken.v.mpa.1D.coef.p + theme(legend.position="none")) +
  (kraken.v.mpa.2D.coef.p + theme(legend.position="none")) +
  plot_layout(nrow=1))
```

* According to MetaPhlAn, uniqueness is decreased by 3 of 4 diets

```{r}
pdf("../plots/kraken_v_metaphlan_diet_coefficients.pdf", height=2.5, width=10)
kraken.v.mpa.DR.coef.p + theme(legend.position="none")
dev.off()
```

# Heritability

Adjust p-values.

```{r}
kraken.v.mpa.herit.df <- kraken.v.mpa.herit.df %>% 
  mutate(padj.kraken = p.adjust(LRT.pval.kraken, method="BH"),
         padj.mpa = p.adjust(LRT.pval.mpa, method="BH")) %>% 
  mutate(is.sig.kraken = padj.kraken < PADJ.THRESH,
         is.sig.mpa = padj.mpa < PADJ.THRESH)
```

Indicate if significant in both, neither, or just one.

```{r}
kraken.v.mpa.herit.df <- kraken.v.mpa.herit.df %>% 
  mutate(is.sig = case_when(
    is.sig.kraken & is.sig.mpa ~ "Both",
    is.sig.kraken & !is.sig.mpa ~ "Only Kraken",
    !is.sig.kraken & is.sig.mpa ~ "Only MetaPhlAn",
    !is.sig.kraken & !is.sig.mpa ~ "Neither"
  ))
```

```{r}
kraken.v.mpa.herit.cor <- cor(kraken.v.mpa.herit.df$herit.kraken, kraken.v.mpa.herit.df$herit.mpa, method="spearman")
kraken.v.mpa.herit.cor.pval <- cor.test(kraken.v.mpa.herit.df$herit.kraken, kraken.v.mpa.herit.df$herit.mpa, method="spearman")$p.value
```

```{r}
(kraken.v.mpa.herit.p <- kraken.v.mpa.herit.df %>% 
  ggplot(aes(x=herit.kraken, y=herit.mpa, label=feature, color=is.sig)) +
  geom_point(shape=16) +
  geom_abline(slope=1, lty=2) +
  geom_text_repel(size=0.36*6, max.overlaps=Inf) +
  labs(x="Heritability, Kraken", y="Heritability, MetaPhlAn",
       color="Significant in?",
       title=sprintf("rho = %.2f, p-value = %2.2e",
                     kraken.v.mpa.herit.cor, kraken.v.mpa.herit.cor.pval)) +
  scale_color_manual(values=c(
    `Only Kraken`="lightskyblue", `Only MetaPhlAn`="lightskyblue3",
    `Both`="navy", Neither="gray80")) +
  theme_classic(base_size=8))å
```

```{r}
pdf("../plots/kraken_v_metaphlan_heritability.pdf", height=2.5, width=2.5)
kraken.v.mpa.herit.p + theme(legend.position="none")
dev.off()
```