---
title: "Analysis of Kraken results BEFORE aggregating by stool ID"
---

# Load libraries

```{r, warning=F, message=F}
library(tidyverse)
library(phyloseq)
library(speedyseq) # for faster tax_glom

diet.palette <- c(AL="seashell4",
                  `1D`="skyblue",
                  `2D`="royalblue4",
                  `20`="orange",
                  `40`="firebrick")
```

# Import metadata

```{r}
seq.meta.df <- read.table("../data/metadata/sequencing_metadata_after_QC_n3473_230620.txt", sep="\t", header=T)
lib.meta.df <- read.table("../data/metadata/library_metadata_after_QC_n3213_230620.txt", sep="\t", header=T)
stool.meta.df <- read.table("../data/metadata/stool_metadata_after_QC_n3124_230620.txt", sep="\t", header=T)
mouse.meta.df <- read.csv("../data/metadata/AnimalData_Processed_20230712.csv")
```

```{r}
seq.meta.annot.df <- seq.meta.df %>%
  merge(lib.meta.df, by="lib.ID") %>%
  merge(stool.meta.df, by="stool.ID") %>% 
  merge(mouse.meta.df, by.x="mouse.ID", by.y="MouseID", all.x=T)

dim(seq.meta.df)
dim(seq.meta.annot.df)
```

```{r}
seq.meta.annot.df %>% dplyr::count(sample.type)
```

* Of the 3473 sequencing samples, 3343 are DO samples, 71 are negative controls, and 59 are positive controls

```{r}
n_distinct(seq.meta.annot.df$stool.ID)

seq.meta.annot.df %>% 
  dplyr::filter(sample.type == "DO") %>% 
  pull(stool.ID) %>% n_distinct()
```

* 3124 unique stool IDs, 2997 unique DO stool IDs

# Import Kraken data

```{r}
kraken.df <- read.table("../data/kraken_matrix_after_QC_n1303x3473_230620.txt", sep="\t", header=T, row.names=1)
dim(kraken.df)
```

# Import Kraken taxonomy

```{r}
kraken.tax.df <- read.table("../data/kraken_taxonomy_n1303.txt", sep="\t", header=T, quote="", row.names=1)
dim(kraken.tax.df)
```

# Create phyloseq object

```{r}
tmp.kraken.sample.meta.df <- seq.meta.annot.df %>% column_to_rownames("seq.ID")
kraken.sample.meta.df <- tmp.kraken.sample.meta.df[colnames(kraken.df), ]

physeq <- phyloseq(
  kraken.df %>% as.matrix %>% otu_table(taxa_are_rows=T),
  kraken.sample.meta.df %>% sample_data,
  kraken.tax.df %>% as.matrix %>% tax_table
)
```

```{r}
nsamples(physeq)
physeq %>% 
  subset_samples(sample.type=="DO") %>% 
  nsamples
```

* 3473 samples, 3343 DO samples

```{r}
physeq %>% 
  subset_samples(sample.type=="DO") %>%
  sample_data() %>% 
  data.frame() %>% 
  rownames_to_column("seq.ID") %>% 
  summarise(across(c(seq.ID, lib.ID, stool.ID, mouse.ID), n_distinct))
```
* 3343 DO samples correspond to 3086 unique libraries, 2997 stool samples, and 913 mice

# Aggregate to genera

```{r}
physeq.genus <- physeq %>% tax_glom(taxrank="genus")
```

# Compute distances

## With controls

```{r}
# Bray-Curtis on relative abundance
genus.bc.dist.w.ctls <- physeq.genus %>%
  transform_sample_counts(function(x) x/sum(x)) %>%
  phyloseq::distance(method="bray")
```

## Without controls

```{r}
physeq.genus.no.ctls <- physeq.genus %>%
  subset_samples(sample.type == "DO")

meta.df.no.ctls <- physeq.genus.no.ctls %>% 
  sample_data() %>% 
  data.frame

# Bray-Curtis on relative abundance
genus.bc.dist <- physeq.genus.no.ctls %>%
  transform_sample_counts(function(x) x/sum(x)) %>%
  phyloseq::distance(method="bray")
```

# PCoA

## With controls

```{r}
pcoa.genus.bc.w.ctls <- genus.bc.dist.w.ctls %>% 
  cmdscale(eig=T, k=3)

frac.expl.by.pcoa1 <- pcoa.genus.bc.w.ctls$eig[1] / sum(pcoa.genus.bc.w.ctls$eig)
frac.expl.by.pcoa2 <- pcoa.genus.bc.w.ctls$eig[2] / sum(pcoa.genus.bc.w.ctls$eig)

pcoa.genus.bc.w.ctls.df <- merge(
  data.frame(pcoa.genus.bc.w.ctls$points) %>% setNames(c("PCoA1", "PCoA2", "PCoA3")),
  kraken.sample.meta.df,
  by="row.names")
```

```{r}
(pcoa.w.ctls.p <- pcoa.genus.bc.w.ctls.df %>% 
   arrange(sample.type) %>% 
   ggplot(aes(x=PCoA1, y=PCoA2, color=sample.type)) +
   geom_point(shape=16, size=0.5) +
   scale_color_manual(values=c(DO="gray40", NEG="brown4", POS="steelblue")) +
   labs(color="Sample type",
        x=sprintf("PCoA1 (%.0f%%)", 100*frac.expl.by.pcoa1),
        y=sprintf("PCoA2 (%.0f%%)", 100*frac.expl.by.pcoa2)) +
   theme_classic(base_size=8))
```

```{r}
# pdf("../plots/pcoa_w_controls.pdf", height=2, width=2)
# pcoa.w.ctls.p + theme(legend.position="none")
# dev.off()
```

```{r}
pcoa.genus.bc.w.ctls.df %>% 
  dplyr::filter(sample.type == "POS" & (PCoA2 < 0.4)) %>%
  pull(Row.names)
```

* The two outlier poscons are JVOMNOOB and BZIZNTZA

## Without controls

```{r}
pcoa.genus.bc <- genus.bc.dist %>% 
  cmdscale(eig=T, k=3)

frac.expl.by.pcoa1.no.ctls <- pcoa.genus.bc$eig[1] / sum(pcoa.genus.bc$eig)
frac.expl.by.pcoa2.no.ctls <- pcoa.genus.bc$eig[2] / sum(pcoa.genus.bc$eig)

pcoa.genus.bc.df <- merge(
  data.frame(pcoa.genus.bc$points) %>% setNames(c("PCoA1", "PCoA2", "PCoA3")),
  meta.df.no.ctls,
  by="row.names")
```

### Same DNA, different libraries

```{r}
stool.IDs.w.multiple.libraries <- pcoa.genus.bc.df %>% 
  group_by(stool.ID) %>% 
  summarise(lib.IDs=paste(sort(unique(lib.ID)), collapse="|")) %>%
  dplyr::filter(str_detect(lib.IDs, "\\|")) %>% 
  pull(stool.ID)

length(stool.IDs.w.multiple.libraries)
```

* 89 stool samples where a library was made twice from the same DNA

```{r}
(pcoa.same.DNA.p <- pcoa.genus.bc.df %>% 
   ggplot(aes(x=PCoA1, y=PCoA2)) +
   geom_point(shape=16, size=0.5, alpha=0.1) +
   geom_path(aes(color=stool.ID, group=stool.ID), linewidth=0.5, data=(
     pcoa.genus.bc.df %>% dplyr::filter(stool.ID %in% stool.IDs.w.multiple.libraries))) +
   labs(x=sprintf("PCoA1 (%.0f%%)", 100*frac.expl.by.pcoa1.no.ctls),
        y=sprintf("PCoA2 (%.0f%%)", 100*frac.expl.by.pcoa2.no.ctls),
        title=sprintf("Same DNA, different libraries\nn=%i stool samples",
                      length(stool.IDs.w.multiple.libraries))) +
   theme_classic(base_size=8) +
   theme(legend.position="none"))
```

```{r}
# pdf("../plots/pcoa_same_DNA_multiple_libs.pdf", width=2, height=2)
# pcoa.same.DNA.p
# dev.off()
```


### Same library, sequenced twice

```{r}
lib.IDs.seqd.more.than.once <- pcoa.genus.bc.df %>% 
  dplyr::count(lib.ID) %>% 
  dplyr::filter(n > 1) %>% 
  pull(lib.ID)

length(lib.IDs.seqd.more.than.once)
```

* 256 libraries sequenced at least twice (255 libraries sequenced twice, 1 library sequenced thrice)

```{r}
(pcoa.same.lib.p <- pcoa.genus.bc.df %>% 
   ggplot(aes(x=PCoA1, y=PCoA2)) +
   geom_point(shape=16, size=0.5, alpha=0.1) +
   geom_path(aes(color=lib.ID, group=lib.ID), linewidth=0.5, data=(
     pcoa.genus.bc.df %>% dplyr::filter(lib.ID %in% lib.IDs.seqd.more.than.once))) +
   labs(x=sprintf("PCoA1 (%.0f%%)", 100*frac.expl.by.pcoa1.no.ctls),
        y=sprintf("PCoA2 (%.0f%%)", 100*frac.expl.by.pcoa2.no.ctls),
        title=sprintf("Same library, sequenced twice\nn=%i",
                      length(lib.IDs.seqd.more.than.once))) +
   theme_classic(base_size=8) +
   theme(legend.position="none"))
```

* Pretty similar

```{r}
# pdf("../plots/pcoa_same_lib.pdf", width=2, height=2)
# pcoa.same.lib.p
# dev.off()
```
