Will compare MetaPhlAn4 versus Kraken2 + MGBC taxonomic results.

* % unclassified
* Mean difference in relative abundance across all samples

# Load libraries

```{r, warning=F, message=F}
library(tidyverse)
library(phyloseq)
library(speedyseq) # faster tax_glom
library(foreach)
```

# Import metadata

```{r}
seq.meta.df <- read.table(
  "../data/metadata/sequencing_metadata_after_QC_n3473_230620.txt", 
  sep="\t", header=T)
lib.meta.df <- read.table(
  "../data/metadata/library_metadata_after_QC_n3213_230620.txt",
  sep="\t", header=T)
stool.meta.df <- read.table(
  "../data/metadata/stool_metadata_after_QC_n3124_240418.txt", 
  sep="\t", header=T)
mouse.meta.df <- read.csv("../data/metadata/AnimalData_Processed_20230712.csv")
```

```{r}
seq.meta.annot.df <- seq.meta.df %>%
  merge(lib.meta.df, by="lib.ID") %>%
  merge(stool.meta.df, by="stool.ID") %>% 
  merge(mouse.meta.df, by.x="mouse.ID", by.y="MouseID", all.x=T)

stool.meta.annot.df <- stool.meta.df %>%
  merge(mouse.meta.df, by.x="mouse.ID", by.y="MouseID")
```

# Import data

## MetaPhlAn

Need MetaPhlAn output prior to aggregation in order to get % unclassified, but we'll use aggregated data for relative abundance comparison.

```{r}
mpa.df <- read.table(
  "../data/metaphlan.txt",
  sep="\t", header=T)

mpa.genus.df <- read.table(
  "../data/metaphlan_genus_agg_by_stool_ID_n376x2997.txt",
  sep="\t", header=T) %>% 
  column_to_rownames("genus")

dim(mpa.genus.df)
```

* 376 genera x 2997 samples

## Kraken

% unclassified stored separately.

```{r}
kraken.perc.unclass.df <- read.table(
  "../results/percent_kraken_unclassified_n4352_221102.txt",
  sep="\t", header=T)
```

Use phyloseq to create genus-level Kraken relative abundances.

```{r}
# import. data
agg.kraken.df <- read.table(
  "../data/kraken_matrix_agg_by_stool_ID_n1303x2997.txt", 
  sep="\t", header=T, row.names=1)

# import taxonomy
kraken.tax.df <- read.table(
  "../data/kraken_taxonomy_n1303.txt", 
  sep="\t", header=T, quote="", row.names=1)

# create phyloseq object
agg.physeq <- phyloseq(
  agg.kraken.df %>% as.matrix %>% otu_table(taxa_are_rows=T),
  stool.meta.annot.df %>% column_to_rownames("stool.ID") %>% sample_data,
  kraken.tax.df %>% as.matrix %>% tax_table)

# aggregate to genera
agg.physeq.genus <- agg.physeq %>% tax_glom(taxrank="genus")

# convert to relative abundances
agg.physeq.genus.relab <- agg.physeq.genus %>% 
  transform_sample_counts(function(x) 100*x/sum(x))

# convert to dataframe
kraken.genus.df <- agg.physeq.genus.relab %>% 
  otu_table() %>% data.frame()
rownames(kraken.genus.df) <- data.frame(tax_table(agg.physeq.genus.relab))$genus

dim(kraken.genus.df)
```

* 252 genera x 2997 samples

# % unclassified

```{r}
perc.unclass.df <- rbind(
  
  # mpa
  mpa.df %>% 
    dplyr::filter(clade_name == "UNCLASSIFIED") %>% 
    dplyr::select(-clade_name) %>% 
    t() %>% data.frame() %>% setNames("perc.unclass") %>% 
    rownames_to_column("seq.ID") %>% 
    mutate(method = "MetaPhlAn4"),
  
  # kraken
  kraken.perc.unclass.df %>% 
    dplyr::rename(perc.unclass = perc.unclass.kraken) %>% 
    mutate(method = "Kraken2")
)
```

```{r}
perc.unclass.df %>% 
  dplyr::count(method)
```

* 4352 Kraken results but only 3473 MetaPhlAn results -- we'll subset to these 3473
* We should also subset to non-control samples

```{r}
seq.IDs.to.keep <- seq.meta.annot.df %>% 
  dplyr::filter(sample.type == "DO") %>% 
  dplyr::filter(seq.ID %in% colnames(mpa.df)) %>% 
  pull(seq.ID)
length(seq.IDs.to.keep)

perc.unclass.filt.df <- perc.unclass.df %>% 
  dplyr::filter(seq.ID %in% seq.IDs.to.keep)
dim(perc.unclass.filt.df)
```

```{r}
(mean.perc.unclass.df <- perc.unclass.filt.df %>% 
   group_by(method) %>% 
   summarise(perc.unclass = mean(perc.unclass), .groups="drop") %>% 
   mutate(perc.class.chr = sprintf("%.1f%%", 100-perc.unclass)))
```

* Mean % classified: 95.7% for Kraken2, 88.0% for MetaPhlAn4

```{r}
(perc.classifed.p <- perc.unclass.filt.df %>% 
   ggplot(aes(x=method, y=100-perc.unclass)) +
   stat_summary(fun="mean", geom="col", width=0.5) +
   geom_point(shape=16) +
   geom_path(aes(group=seq.ID), alpha=0.01) +
   geom_text(aes(y=5, label=perc.class.chr), 
             color="white", data=mean.perc.unclass.df) +
   labs(x="", y="% classified",
        title="DO samples prior to stool.ID aggregation") +
   theme_classic(base_size=8))
```

```{r}
# pdf("../plots/percent_classified_kraken_mgbc_v_mpa4.pdf", width=1.5, height=1.75)
# perc.classifed.p
# dev.off()
```

# Per-sample relabs

```{r}
mpa.v.kraken.genus.relab.df <- merge(
  
  # mpa
  mpa.genus.df %>% 
    rownames_to_column("genus") %>% 
    pivot_longer(-genus, names_to="seq.ID", values_to="mpa"),
  
  # kraken
  kraken.genus.df %>% 
    rownames_to_column("genus") %>% 
    pivot_longer(-genus, names_to="seq.ID", values_to="kraken"),
  
  by=c("genus", "seq.ID")
)
```

```{r}
n_distinct(mpa.v.kraken.genus.relab.df$genus)
```

* 41 genera in common

```{r}
mpa.v.kraken.mean.relab.diff.df <- mpa.v.kraken.genus.relab.df %>% 
  mutate(kraken.minus.mpa = kraken - mpa) %>% 
  group_by(genus) %>% 
  summarise(mean = mean(kraken.minus.mpa),
            sd = sd(kraken.minus.mpa),
            se = sd(kraken.minus.mpa) / sqrt(n()),
            .groups="drop")
```

```{r}
(mpa.v.kraken.relab.diff.p <- mpa.v.kraken.mean.relab.diff.df %>% 
   ggplot(aes(x=mean, xmin=mean-sd, xmax=mean+sd, y=reorder(genus, mean))) +
   geom_pointrange(size=0.1) +
   geom_vline(xintercept=0, lty=2) +
   labs(x="Kraken minus MetaPhlAn relab (mean ± standard deviation)", y="Genus") +
   theme_classic(base_size=8))
```

```{r}
# pdf("../plots/mpa_v_kraken_relab_diffs.pdf", height=4, width=2)
# mpa.v.kraken.relab.diff.p
# dev.off()
```
