We will run PERMANOVA on the full dataset to determine if age and diet have a significant overall effect. This analysis is in its own script because it's slow.

```{r, warning=F, message=F}
library(tidyverse)
library(phyloseq)
library(vegan) # adonis2
library(parallel) # to speed up PERMANOVA
```

# Import metadata

```{r}
stool.meta.df <- read.table(
  "../data/metadata/stool_metadata_after_QC_no_controls_n2997_240418.txt", 
  sep="\t", header=T)
mouse.meta.df <- read.csv(
  "../data/metadata/AnimalData_Processed_20230712.csv")
stool.meta.annot.df <- stool.meta.df %>%
  merge(mouse.meta.df, by.x="mouse.ID", by.y="MouseID")
```

Pre-randomization timepoints are AL.

```{r}
stool.meta.annot.df <- stool.meta.annot.df %>% 
  mutate(Diet.5mo.as.AL=case_when(
    age.wks < 25 ~ "AL",
    TRUE ~ as.character(Diet)))
```

# Import pathway data

```{r}
pathway.tpm.df <- read.table(
  "../data/pathabundance_tpm_agg_by_stool_ID_n422x2997.txt",
  sep="\t", header=T, quote="", row.names=1)

dim(pathway.tpm.df)
```

## log2(TPM)

```{r}
zero.replacement.per.pathway <- apply(pathway.tpm.df, MARGIN=1, FUN=function(x) {min(x[x != 0]) / 2})
```

```{r}
pathway.tpm.mat.no.zeros <- foreach(this.pathway = rownames(pathway.tpm.df), .combine="rbind") %do% {
  
  this.pathway.vec <- as.numeric(pathway.tpm.df[this.pathway, ])
  this.pathway.vec[this.pathway.vec == 0] <- zero.replacement.per.pathway[[this.pathway]]
  
  return(this.pathway.vec)
  
}

pathway.tpm.df.no.zeros <- data.frame(pathway.tpm.mat.no.zeros)
rownames(pathway.tpm.df.no.zeros) <- rownames(pathway.tpm.df)
colnames(pathway.tpm.df.no.zeros) <- colnames(pathway.tpm.df)
```

```{r}
pathway.log2tpm.df <- log2(pathway.tpm.df.no.zeros)
dim(pathway.log2tpm.df)
```

# Compute distances

We'll start with `agg.physeq.genus` made in `kraken.Rmd`.

Bray-Curtis distance on genus-level relative abundances.

```{r}
agg.genus.bc.dist <- agg.physeq.genus %>%
  transform_sample_counts(function(x) x/sum(x)) %>%
  phyloseq::distance(method="bray")

agg.genus.bc.dist.mat <- as.matrix(agg.genus.bc.dist)
```

Euclidean distance on log2(TPM) pathway abundances.

```{r}
pathway.dist <- dist(t(pathway.log2tpm.df))
pathway.dist.mat <- as.matrix(pathway.dist)
dim(pathway.dist.mat)
```

# PERMANOVA

## Random subset of samples

```{r}
# subset to N random samples so that it runs faster
set.seed(1899)
n.random.stool.IDs <- sample(rownames(agg.genus.bc.dist.mat), size=500)
this.dist <- as.dist(agg.genus.bc.dist.mat[n.random.stool.IDs, n.random.stool.IDs])
this.tmp.meta.df <- stool.meta.annot.df %>% column_to_rownames("stool.ID")
this.meta.df <- this.tmp.meta.df[n.random.stool.IDs, ] %>% rownames_to_column("stool.ID")

# confirm that rownames of distance matrix match the order in the metadata
# identical(this.meta.df$stool.ID, rownames(as.matrix(this.dist)))

permanova.res <- adonis2(
  this.dist ~ age.approx.months + Diet.5mo.as.AL,
  by="margin",
  data=this.meta.df, 
  permutations=999, 
  parallel=4)
```

## All samples

```{r}
# confirm that rownames of distance matrix match the order in the metadata
# identical(stool.meta.annot.df$stool.ID, rownames(agg.genus.bc.dist.mat))

permanova.res <- adonis2(
  agg.genus.bc.dist.mat ~ age.approx.months + Diet.5mo.as.AL,
  by="margin",
  data=stool.meta.annot.df, 
  permutations=999, 
  parallel=4)
```

```{r}
permanova.res
```

## Pathways

```{r}
# confirm that rownames of distance matrix match the order in the metadata
# identical(stool.meta.annot.df$stool.ID, rownames(pathway.dist.mat))

pathway.permanova.res <- adonis2(
  pathway.dist.mat ~ age.approx.months + Diet.5mo.as.AL,
  by="margin",
  data=stool.meta.annot.df, 
  permutations=999, 
  parallel=4)
```

```{r}
pathway.permanova.res
```
