Microbiome-phenotype longitudinal association and mediation analysis.

Cross-sectional association analysis at the very end.

# Load libraries

```{r, warning=F, message=F}
library(tidyverse)
library(patchwork) # plot_layout

PADJ.THRESH <- 0.01

diet.palette <- c(AL="seashell4",
                  `1D`="skyblue",
                  `2D`="royalblue4",
                  `20`="orange",
                  `40`="firebrick")
domain.colors <- c(
  AS="aquamarine3", 
  BW="royalblue", 
  CBC="firebrick", 
  Echo="brown",
  FACS="tomato", 
  Frailty="tan4", 
  Glu="chocolate",
  Grip="darkblue", 
  MetCage="green4", 
  PIXI="blueviolet", 
  Rotarod="mediumblue", 
  Void="pink", 
  Wheel="gold"
)
```

# Import metadata

```{r}
stool.meta.df <- read.table(
  "../data/metadata/stool_metadata_after_QC_no_controls_n2997_240418.txt", 
  sep="\t", header=T)
mouse.meta.df <- read.csv(
  "../data/metadata/AnimalData_Processed_20230712.csv")
stool.meta.annot.df <- stool.meta.df %>%
  merge(mouse.meta.df, by.x="mouse.ID", by.y="MouseID")

# identify pre-randomization samples as AL
stool.meta.annot.df <- stool.meta.annot.df %>%
  mutate(Diet.5mo.as.AL=factor(case_when(
    age.wks < 25 ~ "AL",
    TRUE ~ as.character(Diet)),
    levels=c("AL", "1D", "2D", "20", "40")))
```

# Import original data (to plot examples)

```{r}
genus.df <- read.table(
  "../results/kraken_genus_clr_filt_w_comm_n107x2997_231017.txt", 
  sep="\t", header=T, row.names=1)
pathway.df <- read.table(
  "../results/DO_pathway_log2tpm_filt_w_comm_n273x2997.txt",
  sep="\t", header=T, row.names=1)
pheno.df <- read.table(
  "../results/phenotype_values_closest_to_each_stool_ID_within_100_days_n2915x210_230811.txt", sep="\t", header=T)
```

Combine.

```{r}
full.df <- pheno.df %>%
  merge(stool.meta.annot.df, by="stool.ID") %>%
  merge(data.frame(t(genus.df)) %>% rownames_to_column("stool.ID"), by="stool.ID") %>%
  merge(data.frame(t(pathway.df)) %>% rownames_to_column("stool.ID"), by="stool.ID")

dim(full.df)
```

## Import short and long pathway names

```{r}
pathway.names.df <- read.table(
  "../data/metadata/pathway_long_and_short_names_n703.txt",
  sep="\t", header=T, quote="")
```

Add additional columns for plotting.

```{r}
pathway.names.df <- pathway.names.df %>% 
  mutate(after.colon = str_sub(str_extract(original, "(?=:)(.*)"), 3)) %>% 
  mutate(after.colon.short = ifelse(
    nchar(after.colon) > 30, 
    paste0(str_sub(after.colon, 1, 27), "..."),
    after.colon)) %>% 
  mutate(after.colon.short.w.ID = paste0(
    after.colon.short, " (", short, ")")) %>% 
  relocate(after.colon.short.w.ID)
```

We need `short.clean` and `after.colon.short.w.ID`.

# Import association results

```{r}
tax.pheno.res.df <- read.table(
  "../results/lme4_mb_pheno_assoc_n107_kraken_feats_n209_phenos.txt.gz",
  sep="\t", header=T)

fxn.pheno.res.df <- read.table(
  "../results/lme4_mb_pheno_assoc_n273_pathways_n209_phenos.txt.gz",
  sep="\t", header=T)

dim(tax.pheno.res.df)
dim(fxn.pheno.res.df)
```

## Remove phenos

Ignore BWTest traits (because they were just pulled from the weekly bodyweight measurements) except MetCage_BWTest (which was measured independently).

```{r}
tax.pheno.res.df <- tax.pheno.res.df %>% 
  dplyr::filter(!str_detect(pheno, "BWTest") | (pheno == "MetCage_BWTest"))

fxn.pheno.res.df <- fxn.pheno.res.df %>% 
  dplyr::filter(!str_detect(pheno, "BWTest") | (pheno == "MetCage_BWTest"))

dim(tax.pheno.res.df)
dim(fxn.pheno.res.df)
```

## Adjust p-values

```{r}
tax.pheno.res.df <- tax.pheno.res.df %>% 
  mutate(assoc_padj = p.adjust(mb.pval, method="BH")) %>%
  mutate(assoc_is_sig = assoc_padj < PADJ.THRESH)

fxn.pheno.res.df <- fxn.pheno.res.df %>% 
  mutate(assoc_padj = p.adjust(mb.pval, method="BH")) %>% 
  mutate(assoc_is_sig = assoc_padj < PADJ.THRESH)
```

```{r}
tax.pheno.res.df %>% 
  dplyr::count(assoc_is_sig) %>% 
  mutate(frac = n / sum(n))
```

```{r}
fxn.pheno.res.df %>% 
  dplyr::count(assoc_is_sig) %>% 
  mutate(frac = n / sum(n))
```

* The fraction of significant associations is a little lower for pathways: 0.9% (tax) and 0.4% (fxn) of tests 
* 185 taxonomic associations, 202 functional associations with padj < 0.01

### p-value histograms

```{r}
(tax.mb.pheno.assoc.pval.hist <- tax.pheno.res.df %>% 
   mutate(assoc_is_sig = factor(assoc_is_sig, levels=c(T,F))) %>% 
   ggplot(aes(mb.pval, fill=assoc_is_sig)) +
   geom_histogram(bins=100) +
   labs(x="p-value", y="# associations", 
        title="Genera", fill="padj < 0.01") +
   scale_fill_manual(values=c(`TRUE`="royalblue", `FALSE`="gray")) +
   theme_classic(base_size=8) +
   theme(legend.position="none"))
```

```{r}
# pdf("../plots/mb_pheno_assoc_pval_histogram_tax.pdf", width=2, height=2)
# tax.mb.pheno.assoc.pval.hist
# dev.off()
```

```{r}
(fxn.mb.pheno.assoc.pval.hist <- fxn.pheno.res.df %>% 
   mutate(assoc_is_sig = factor(assoc_is_sig, levels=c(T,F))) %>% 
   ggplot(aes(mb.pval, fill=assoc_is_sig)) +
   geom_histogram(bins=100) +
   labs(x="p-value", y="# associations", 
        title="Pathways", fill="padj < 0.01") +
   scale_fill_manual(values=c(`TRUE`="royalblue", `FALSE`="gray")) +
   theme_classic(base_size=8) +
   theme(legend.position="none"))
```

```{r}
# pdf("../plots/mb_pheno_assoc_pval_histogram_fxn.pdf", width=2, height=2)
# fxn.mb.pheno.assoc.pval.hist
# dev.off()
```

* Both functional and taxonomic features have very nice-looking p-value distributions

# Import mediation results

```{r}
tax.mediation.df <- read.table(
  "../results/mediation_n107_genus_feats_n209_phenos.txt.gz",
  sep="\t", header=T)

fxn.mediation.df <- read.table(
  "../results/mediation_n273_humann_pathway_feats_n209_phenos.txt.gz",
  sep="\t", header=T)

dim(tax.mediation.df)
dim(fxn.mediation.df)
```

# Record feature names

```{r}
all.genus.feats <- rownames(genus.df)
all.pathway.feats <- rownames(pathway.df)
```

Make sure taxonomic uniqueness is always "uniq"

```{r}
tax.pheno.res.df <- tax.pheno.res.df %>% 
  mutate(feature = case_match(feature, "uniq_genus" ~ "uniq", .default=feature))

str_subset(all.genus.feats, "niq")
str_subset(unique(tax.pheno.res.df$feature), "niq")
str_subset(unique(tax.mediation.df$feature), "niq")
```

Make sure functional uniqueness is always "Uniqueness"

```{r}
str_subset(all.pathway.feats, "niq")
str_subset(unique(fxn.pheno.res.df$feature), "niq")
str_subset(unique(fxn.mediation.df$feature), "niq")
```

```{r}
community.features <- c(
  "Chao1", "Shannon", "Simpson", "PCoA1", "PCoA2", "PCoA3", "uniq",
  "Uniqueness")
```

# Merge association and mediation results

```{r}
tax.assoc.mediation.merged.df <- merge(tax.pheno.res.df, tax.mediation.df, by=c("pheno", "feature"))
fxn.assoc.mediation.merged.df <- merge(fxn.pheno.res.df, fxn.mediation.df, by=c("pheno", "feature"))
```

## Adjust metadata

```{r}
tax.assoc.mediation.merged.df <- tax.assoc.mediation.merged.df %>% 
  separate(col=pheno, into=c("domain", "trait"), sep="_", remove=F)

fxn.assoc.mediation.merged.df <- fxn.assoc.mediation.merged.df %>% 
  separate(col=pheno, into=c("domain", "trait"), sep="_", remove=F)
```

Differentiate between individual and community features.

```{r}
tax.assoc.mediation.merged.df <- tax.assoc.mediation.merged.df %>% 
  mutate(feature.type = ifelse(feature %in% community.features, "community", "individual"))

fxn.assoc.mediation.merged.df <- fxn.assoc.mediation.merged.df %>% 
  mutate(feature.type = ifelse(feature %in% community.features, "community", "individual"))
```

Make both of the glucose domains into "Glu" and change the traits to "GlucoseF" and "GlucoseNF".

```{r}
tax.assoc.mediation.merged.df <- tax.assoc.mediation.merged.df %>% 
  mutate(trait = case_when(
    domain == "Glu.F" & trait == "Glucose" ~ "GlucoseF",
    domain == "Glu.NF" & trait == "Glucose" ~ "GlucoseNF",
    TRUE ~ trait)) %>% 
  mutate(domain = ifelse(domain %in% c("Glu.F", "Glu.NF"), "Glu", domain))

fxn.assoc.mediation.merged.df <- fxn.assoc.mediation.merged.df %>% 
  mutate(trait = case_when(
    domain == "Glu.F" & trait == "Glucose" ~ "GlucoseF",
    domain == "Glu.NF" & trait == "Glucose" ~ "GlucoseNF",
    TRUE ~ trait)) %>% 
  mutate(domain = ifelse(domain %in% c("Glu.F", "Glu.NF"), "Glu", domain))
```

## Adjust mediation p-values

Separately within diet.

```{r}
tax.assoc.mediation.merged.df <- tax.assoc.mediation.merged.df %>% 
  group_by(diet) %>% 
  mutate(ACME_padj = p.adjust(ACME_p, method="BH")) %>% 
  ungroup() %>% 
  mutate(ACME_is_sig = ACME_padj < PADJ.THRESH)

fxn.assoc.mediation.merged.df <- fxn.assoc.mediation.merged.df %>% 
  group_by(diet) %>% 
  mutate(ACME_padj = p.adjust(ACME_p, method="BH")) %>% 
  ungroup() %>% 
  mutate(ACME_is_sig = ACME_padj < PADJ.THRESH)
```

```{r}
tax.assoc.mediation.merged.df %>% 
  dplyr::count(diet, ACME_is_sig) %>% 
  group_by(diet) %>% 
  mutate(frac = n/sum(n)) %>% 
  ungroup() %>% 
  dplyr::filter(ACME_is_sig)
```

```{r}
fxn.assoc.mediation.merged.df %>% 
  dplyr::count(diet, ACME_is_sig) %>% 
  group_by(diet) %>% 
  mutate(frac = n/sum(n)) %>% 
  ungroup() %>% 
  dplyr::filter(ACME_is_sig)
```

* For 40% CR, 1.3% of taxonomic and 0.8% of functional associations are hits

# Proportion mediated

```{r}
tax.assoc.mediation.merged.df %>%
  
  # subset to significant mediation pairs
  dplyr::filter(ACME_is_sig) %>%
  
  # subset to pairs where direct and mediation effect are in the same direction
  # https://stats.stackexchange.com/questions/504427/mediation-in-r-impossible-proportion-mediated-values-output-interpretation
  dplyr::filter(sign(ACME_est) == sign(ADE_est)) %>% 
  
  pull(prop.med) %>% 
  summary()
```

```{r}
fxn.assoc.mediation.merged.df %>%
  dplyr::filter(ACME_is_sig) %>%
  dplyr::filter(sign(ACME_est) == sign(ADE_est)) %>% 
  pull(prop.med) %>% 
  summary()
```

* For both taxonomic and functional features, proportion mediated is mean 19%, median 14%

# Overlap

## Taxonomic

```{r}
num.sig.tax.assoc.p <- tax.assoc.mediation.merged.df %>% 
  dplyr::filter(diet == "40") %>% 
  dplyr::count(assoc_is_sig) %>% 
  dplyr::filter(assoc_is_sig) %>% 
  mutate(label="Association") %>% 
  ggplot(aes(y=label, x=n, label=n)) +
  geom_col(fill="skyblue") +
  geom_text(size=6*0.36) +
  scale_x_continuous(position="top", limits=c(0, 300)) + 
  labs(y="", x="") +
  theme_classic(base_size=8)
```

```{r}
num.sig.tax.mediation.p <- tax.assoc.mediation.merged.df %>%
  dplyr::filter(ACME_is_sig) %>% 
  dplyr::count(diet, assoc_is_sig) %>% 
  mutate(label=factor(
    paste0("Mediation", diet),
    levels=c("Mediation1D", "Mediation2D", "Mediation20", "Mediation40"))) %>% 
  ggplot(aes(y=label, x=n, fill=assoc_is_sig, label=n)) +
  geom_col(position="stack") +
  geom_text(position="stack", size=6*0.36) +
  labs(y="", x="# significant results", fill="Also significant association?") +
  scale_fill_manual(values=c(`TRUE`="skyblue", `FALSE`="gray")) +
  xlim(0, 300) +
  theme_classic(base_size=8) +
  theme(legend.position="none")

assoc.v.mediation.tax.feats.barplot <- num.sig.tax.assoc.p + 
  num.sig.tax.mediation.p + plot_layout(ncol=1, heights=c(1,4))
```

```{r}
# pdf("../plots/assoc_v_mediation_tax_feats_barplots.pdf", width=2.5, height=1.5)
# assoc.v.mediation.tax.feats.barplot
# dev.off()
```

## Functional

```{r}
num.sig.fxn.assoc.p <- fxn.assoc.mediation.merged.df %>% 
  dplyr::filter(diet == "40") %>% 
  dplyr::count(assoc_is_sig) %>% 
  dplyr::filter(assoc_is_sig) %>% 
  mutate(label="Association") %>% 
  ggplot(aes(y=label, x=n, label=n)) +
  geom_col(fill="skyblue") +
  geom_text(size=6*0.36) +
  scale_x_continuous(position="top", limits=c(0, 420)) + 
  labs(y="", x="") +
  theme_classic(base_size=8)
```

```{r}
num.sig.fxn.mediation.p <- fxn.assoc.mediation.merged.df %>%
  dplyr::filter(ACME_is_sig) %>% 
  dplyr::count(diet, assoc_is_sig) %>% 
  mutate(label=factor(
    paste0("Mediation", diet),
    levels=c("Mediation1D", "Mediation2D", "Mediation20", "Mediation40"))) %>% 
  ggplot(aes(y=label, x=n, fill=assoc_is_sig, label=n)) +
  geom_col(position="stack") +
  geom_text(position="stack", size=6*0.36) +
  labs(y="", x="# significant results", fill="Also significant association?") +
  scale_fill_manual(values=c(`TRUE`="skyblue", `FALSE`="gray")) +
  xlim(0, 420) +
  theme_classic(base_size=8) +
  theme(legend.position="none")

assoc.v.mediation.fxn.feats.barplot <- num.sig.fxn.assoc.p + 
  num.sig.fxn.mediation.p + plot_layout(ncol=1, heights=c(1,4))
```

```{r}
# pdf("../plots/assoc_v_mediation_fxn_feats_barplots.pdf", width=2.5, height=1.5)
# assoc.v.mediation.fxn.feats.barplot
# dev.off()
```

# Domains with the most associations

```{r}
frac.hits.per.domain.df <- rbind(
  
  # genera
  tax.assoc.mediation.merged.df %>% 
    dplyr::filter(diet == "40") %>% # assoc results are same for all diets
    mutate(assoc_is_sig = factor(assoc_is_sig)) %>% 
    dplyr::count(domain, assoc_is_sig, .drop=F) %>% 
    group_by(domain) %>% mutate(frac = n/sum(n)) %>% ungroup() %>% 
    dplyr::filter(assoc_is_sig == T) %>% 
    mutate(genera.or.pathways = "genera"),
  
  # pathways
  fxn.assoc.mediation.merged.df %>% 
    dplyr::filter(diet == "40") %>% # assoc results are same for all diets
    mutate(assoc_is_sig = factor(assoc_is_sig)) %>% 
    dplyr::count(domain, assoc_is_sig, .drop=F) %>% 
    group_by(domain) %>% mutate(frac = n/sum(n)) %>% ungroup() %>% 
    dplyr::filter(assoc_is_sig == T) %>% 
    mutate(genera.or.pathways = "pathways")
)
```

```{r}
(frac.hits.per.domain.p <- frac.hits.per.domain.df %>% 
   ggplot(aes(x=100*frac, y=reorder(domain, frac), fill=domain)) +
   geom_col() +
   labs(x=sprintf("%% microbiome-phenotype pairs with padj < %.2f", PADJ.THRESH), 
        y="Domain") +
   facet_wrap(~genera.or.pathways, scales="free_x") +
   scale_fill_manual(values=domain.colors) +
   theme_classic(base_size=8) +
   theme(legend.position="none"))
```

```{r}
# pdf("../plots/mb_pheno_assoc_frac_hits_per_domain_barplot.pdf", height=2, width=3)
# frac.hits.per.domain.p
# dev.off()
```

# Heatmaps

## Taxonomic bodyweight associations

```{r}
bw.traits <- c("BW", "BWTest", "TTM", "PercFat")
bw.tax.feats <- c(
  "Alistipes", "COE1", "Lachnospira", "Ligilactobacillus", # positive coefficient
  "Angelakisella", "Bifidobacterium", # negative coefficient
  "Dubosiella", "Paramuribaculum")
```

```{r}
(tax.bw.assoc.heatmap <- tax.assoc.mediation.merged.df %>% 
   dplyr::filter(diet == "40") %>% # assoc results are same for all diets
   dplyr::filter(feature %in% bw.tax.feats) %>% 
   dplyr::filter(trait %in% bw.traits) %>% 
   mutate(trait = factor(case_when(
     trait == "BW" ~ "Body weight",
     trait == "BWTest" ~ "Body weight (Met cage)",
     trait == "TTM" ~ "Total tissue mass (DEXA)",
     trait == "PercFat" ~ "% fat (DEXA)"), levels=c(
       "% fat (DEXA)", "Total tissue mass (DEXA)",
       "Body weight (Met cage)", "Body weight"))) %>%
   mutate(feature = factor(feature, levels=bw.tax.feats)) %>%
   mutate(signif.symbol = case_when(
    assoc_padj < PADJ.THRESH ~ "*",
    TRUE ~ "")) %>% 
   ggplot(aes(x=feature, y=trait, fill=mb.coef, label=signif.symbol)) +
   geom_tile() +
   geom_text() +
   theme_classic(base_size=8) +
   theme(axis.text.x=element_text(angle=45, hjust=1)) +
   labs(x="Taxonomic feature", y="Phenotype", title="Body mass-related phenotypes", 
        fill="Coefficient") +
   scale_fill_gradient2(high="firebrick", low="dodgerblue"))
```

```{r}
# pdf("../plots/mb_pheno_assoc_bw_select_tax_feats_heatmap.pdf", height=2, width=4)
# tax.bw.assoc.heatmap
# dev.off()
```

## Significant by both association and mediation (40% CR)

Taxonomic and functional features together.

```{r}
traits.to.show.in.heatmap <- c(
  "BW", "BWTest", "PercFat", "TTM", "Delta", # body composition
  "EE", "VO2", "VCO2", # energy consumption
  "Frailty", "CoatCondition", "DistendedAbdomen", # frailty
  "Kyphosis", "TailStiffening", "VisionLoss", # more frailty
  "Myeloid.PercViable", "Lymph.PercViable" # immune from FACS
)
length(traits.to.show.in.heatmap)

tax.feats.for.heatmap <- c(
  "Akkermansia", "Alistipes", "Angelakisella", "Bacteroides", "Bifidobacterium", 
  "Clostridium", "COE1", "Desulfovibrio", "Dubosiella",
  "Duncaniella", "Emergencia", "Escherichia", "Lachnospira", "Ligilactobacillus",
  "Muribaculum", "MuribaculaceaeNOV", "Paramuribaculum", "Roseburia", 
  "Turicibacter", "uniq")
length(tax.feats.for.heatmap)

fxn.feats.for.heatmap <- fxn.assoc.mediation.merged.df %>%
  dplyr::filter(diet == "40") %>% 
  dplyr::filter(trait %in% traits.to.show.in.heatmap) %>%
  dplyr::filter(assoc_is_sig & ACME_is_sig) %>%
  pull(feature) %>% unique()
length(fxn.feats.for.heatmap)
```

Merge taxonomic and functional results prior to plotting.

```{r}
plot.df.for.tax.and.fxn.heatmap <- rbind(
  
  tax.assoc.mediation.merged.df %>% 
    dplyr::filter(diet == "40") %>% 
    dplyr::filter(trait %in% traits.to.show.in.heatmap) %>%
    dplyr::filter(feature %in% tax.feats.for.heatmap) %>%
    mutate(trait = factor(trait, levels=rev(traits.to.show.in.heatmap))) %>% 
    dplyr::select(feature, trait, mb.coef, assoc_is_sig, ACME_is_sig) %>% 
    mutate(tax.or.fxn = "genera"),
  
  fxn.assoc.mediation.merged.df %>% 
    dplyr::filter(diet == "40") %>% 
    dplyr::filter(trait %in% traits.to.show.in.heatmap) %>%
    dplyr::filter(feature %in% fxn.feats.for.heatmap) %>%
    merge(pathway.names.df, by.x="feature", by.y="short.clean") %>%
    dplyr::mutate(feature = after.colon.short.w.ID) %>% # overwrite old feature column
    dplyr::select(feature, trait, mb.coef, assoc_is_sig, ACME_is_sig) %>% 
    mutate(tax.or.fxn = "pathways")
  
) %>% 
  
  # specify order of traits
  mutate(trait = factor(trait, levels=rev(traits.to.show.in.heatmap))) %>% 
  
  # create symbol to indicate signif in assoc or both assoc and mediation
  mutate(is.sig.symbol = case_when(
     (assoc_is_sig & !ACME_is_sig) ~ "+", 
     (assoc_is_sig & ACME_is_sig) ~ "++", 
     TRUE ~ ""))
```

```{r}
(tax.and.fxn.heatmap.tall <- plot.df.for.tax.and.fxn.heatmap %>% 
   ggplot(aes(y=feature, x=trait, fill=mb.coef, label=is.sig.symbol)) +
   geom_tile() +
   geom_text(size=6*.36) +
   facet_wrap(~tax.or.fxn, scales="free_y", ncol=1) +
   theme_classic(base_size=8) +
   theme(axis.text.x=element_text(angle=45, hjust=1),
         legend.position="none") +
   labs(y="Feature", x="Phenotype", 
        fill="Coefficient") +
   scale_fill_gradient2(high="firebrick", low="dodgerblue"))
```

```{r}
# pdf("../plots/signif_assoc_and_med40_heatmap_tall_tax_and_fxn.pdf", height=6, width=4.5)
# tax.and.fxn.heatmap.tall
# dev.off()
```

# Examples

## Paramuribaculum versus % fat

```{r}
(paramuribaculum.v.percfat.split.by.diet.p <- full.df %>% 
   ggplot(aes(x=Paramuribaculum, y=PIXI_PercFat)) +
   geom_point(aes(color=Diet.5mo.as.AL), shape=16, size=0.5) +
   geom_smooth(method="lm", linewidth=0.5) +
   facet_wrap(~Diet.5mo.as.AL, nrow=1) +
   scale_color_manual(values=diet.palette) +
   labs(x="Paramuribaculum (CLR)", y="% fat (DEXA)") +
   theme_classic(base_size=8) +
   theme(legend.position="none"))
```

```{r}
# pdf("../plots/paramuribaculum_v_percfat_split_by_diet.pdf", height=1, width=4)
# paramuribaculum.v.percfat.split.by.diet.p
# dev.off()
```

## Akkermansia versus energy expenditure

```{r}
(akkermansia.v.MetCageEE.p <- full.df %>% 
   ggplot(aes(x=Akkermansia, y=MetCage_EE)) +
   geom_point(aes(color=Diet.5mo.as.AL), shape=16, size=0.5) +
   geom_smooth(method="lm", linewidth=0.5) +
   facet_wrap(~Diet.5mo.as.AL, nrow=1) +
   scale_color_manual(values=diet.palette) +
   theme_classic(base_size=8) +
   labs(x="CLR", y="Energy expenditure (metabolic cage)", title="Akkermansia") +
   theme(legend.position="none"))
```

```{r}
# pdf("../plots/akkermansia_v_EE.pdf", height=1.25, width=3.5)
# akkermansia.v.MetCageEE.p
# dev.off()
```

## Methionine pathway versus VCO2

```{r}
(PWY7977.v.MetCageVCO2.p <- full.df %>% 
   ggplot(aes(x=PWY7977, y=MetCage_VCO2)) +
   geom_point(aes(color=Diet.5mo.as.AL), shape=16, size=0.5) +
   geom_smooth(method="lm", linewidth=0.5) +
   facet_wrap(~Diet.5mo.as.AL, nrow=1) +
   scale_color_manual(values=diet.palette) +
   theme_classic(base_size=8) +
   labs(x="log2(TPM)", y="VCO2 (metabolic cage)", title="L-methionine biosynthesis IV (PWY-7977)") +
   theme(legend.position="none"))
```

```{r}
# pdf("../plots/PWY7977_v_MetCageVCO2.pdf", height=1.25, width=3.5)
# PWY7977.v.MetCageVCO2.p
# dev.off()
```

# Cross-sectional microbiome-phenotype associations

## Import linear model results

Model at 5 months: y_pheno ~ y_mb
Model at other ages: y_pheno ~ y_mb + DR

```{r}
tax.pheno.res.xs.df <- read.table(
  "../results/lm_cross_sectional_mb_pheno_assoc_diet_covar_n107_kraken_feats.txt",
  sep="\t", header=T)
```

```{r}
fxn.pheno.res.xs.df <- read.table(
  "../results/lm_cross_sectional_mb_pheno_assoc_diet_covar_n273_pathways.txt",
  sep="\t", header=T)
```

## Exclude phenos

Ignore BWTest traits except MetCage_BWTest.

```{r}
tax.pheno.res.xs.df <- tax.pheno.res.xs.df %>% 
  dplyr::filter(!str_detect(pheno, "BWTest") | (pheno == "MetCage_BWTest"))

fxn.pheno.res.xs.df <- fxn.pheno.res.xs.df %>% 
  dplyr::filter(!str_detect(pheno, "BWTest") | (pheno == "MetCage_BWTest"))

dim(tax.pheno.res.xs.df)
dim(fxn.pheno.res.xs.df)
```

Separate pheno into domain and trait.

```{r}
tax.pheno.res.xs.df <- tax.pheno.res.xs.df %>% 
  mutate(pheno = case_match(pheno, "SurvDays" ~ "SurvDays_SurvDays", .default=pheno)) %>% 
  separate(col=pheno, into=c("domain", "trait"), sep="_", remove=F)

fxn.pheno.res.xs.df <- fxn.pheno.res.xs.df %>% 
  mutate(pheno = case_match(pheno, "SurvDays" ~ "SurvDays_SurvDays", .default=pheno)) %>% 
  separate(col=pheno, into=c("domain", "trait"), sep="_", remove=F)
```

Make both of the glucose domains into "Glu" and change the traits to "GlucoseF" and "GlucoseNF".

```{r}
tax.pheno.res.xs.df <- tax.pheno.res.xs.df %>% 
  mutate(trait = case_when(
    domain == "Glu.F" & trait == "Glucose" ~ "GlucoseF",
    domain == "Glu.NF" & trait == "Glucose" ~ "GlucoseNF",
    TRUE ~ trait)) %>% 
  mutate(domain = ifelse(domain %in% c("Glu.F", "Glu.NF"), "Glu", domain))

fxn.pheno.res.xs.df <- fxn.pheno.res.xs.df %>% 
  mutate(trait = case_when(
    domain == "Glu.F" & trait == "Glucose" ~ "GlucoseF",
    domain == "Glu.NF" & trait == "Glucose" ~ "GlucoseNF",
    TRUE ~ trait)) %>% 
  mutate(domain = ifelse(domain %in% c("Glu.F", "Glu.NF"), "Glu", domain))
```

## Adjust p-values

Separately within age.

```{r}
tax.pheno.res.xs.df <- tax.pheno.res.xs.df %>% 
  group_by(age) %>% 
  mutate(padj = p.adjust(pval, method="BH")) %>% 
  mutate(is.sig = padj < PADJ.THRESH) %>% 
  ungroup()
  
fxn.pheno.res.xs.df <- fxn.pheno.res.xs.df %>% 
  group_by(age) %>% 
  mutate(padj = p.adjust(pval, method="BH")) %>% 
  mutate(is.sig = padj < PADJ.THRESH) %>% 
  ungroup()
```

## Domains with the most associations

```{r}
(frac.xs.hits.per.domain.barplots <- rbind(
  
  # tax
  tax.pheno.res.xs.df %>% 
    mutate(is.sig = factor(is.sig)) %>% 
    dplyr::count(domain, is.sig, .drop=F) %>% 
    group_by(domain) %>% mutate(frac = n/sum(n)) %>% ungroup() %>% 
    dplyr::filter(is.sig == T) %>% 
    mutate(genera.or.pathways = "genera"),
  
  # fxn
  fxn.pheno.res.xs.df %>% 
    mutate(is.sig = factor(is.sig)) %>% 
    dplyr::count(domain, is.sig, .drop=F) %>% 
    group_by(domain) %>% mutate(frac = n/sum(n)) %>% ungroup() %>% 
    dplyr::filter(is.sig == T) %>% 
    mutate(genera.or.pathways = "pathways")
) %>% 
  
  ggplot(aes(x=100*frac, y=reorder(domain, frac), fill=domain)) +
  geom_col() +
  labs(x=sprintf("%% microbiome-phenotype pairs with padj < %.2f", PADJ.THRESH), 
       y="Domain") +
  scale_fill_manual(values=domain.colors) +
  theme_classic(base_size=8) +
  facet_wrap(~genera.or.pathways, scales="free_x") +
  theme(legend.position="none"))
```

```{r}
# pdf("../plots/mb_pheno_assoc_frac_xs_hits_per_domain_barplot.pdf", height=2, width=2)
# frac.xs.hits.per.domain.barplots
# dev.off()
```
