---
title: "Prediction"
---

```{r, warning=F, message=F}
library(tidyverse)
library(randomForest)
library(caret)
library(broom) # tidy
library(ggpubr) # stat_compare_means

age.color <- "seagreen"
diet.color <- "mediumorchid"
diet.palette <- c(AL="seashell4",
                  `1D`="skyblue",
                  `2D`="royalblue4",
                  `20`="orange",
                  `40`="firebrick")
```

# Import metadata

```{r}
stool.meta.df <- read.table(
  "../data/metadata/stool_metadata_after_QC_no_controls_n2997_240418.txt", 
  sep="\t", header=T)
mouse.meta.df <- read.csv("../data/metadata/AnimalData_Processed_20230712.csv") %>% 
  mutate(Cage=paste0("c", HID)) # create character version of cage

stool.meta.annot.df <- stool.meta.df %>%
  merge(mouse.meta.df, by.x="mouse.ID", by.y="MouseID")
```

Pre-randomization timepoints are AL.

```{r}
stool.meta.annot.df <- stool.meta.annot.df %>% 
  mutate(Diet.5mo.as.AL=case_when(
    age.wks < 25 ~ "AL",
    TRUE ~ as.character(Diet)))
```

# Import taxonomic data

Will use relative abundances for top 100 most abundant genera.

```{r}
genus.data.df.feats.in.rows <- read.table(
  "../data/kraken_genus_relab_n100x2997.txt",
  sep="\t", header=T, row.names=1)

tax.feats <- rownames(genus.data.df.feats.in.rows)
length(tax.feats)
```

Add annotations.

```{r}
genus.annot.df <- genus.data.df.feats.in.rows %>% 
  t() %>% data.frame() %>% rownames_to_column("stool.ID") %>% 
  merge(stool.meta.annot.df, by="stool.ID")
dim(genus.annot.df)
```

# Import pathway data

Will use log2(TPM) abundances for the 252 pathways that passed prevalence filtration.

```{r}
pathway.log2tpm.df.feats.in.rows <- read.table(
  "../results/DO_pathway_log2tpm_filt_w_comm_n273x2997.txt",
  sep="\t", header=T, quote="")
```

Exclude uniqueness and transpose to get samples in the rows.

```{r}
pathway.log2tpm.df <- pathway.log2tpm.df.feats.in.rows %>%
  dplyr::filter(feature != "Uniqueness") %>% 
  column_to_rownames("feature") %>% 
  t() %>% data.frame()

fxn.feats <- colnames(pathway.log2tpm.df)
length(fxn.feats)
```

* 272 pathways (after prevalence filtration)


Add annotations.

```{r}
pathway.log2tpm.annot.df <- pathway.log2tpm.df %>% 
  rownames_to_column("stool.ID") %>% 
  merge(stool.meta.annot.df, by="stool.ID")
dim(pathway.log2tpm.annot.df)
```

# Predict age of AL mice

## Genera

```{r}
genus.AL.annot.df <- genus.annot.df %>% 
  dplyr::filter(Diet == "AL")
```

We will train on 70%, test on 30%, and stratify by cage. Separate cages into training and testing.

```{r}
# extract vector of cages for stratification
AL.cage.vec <- genus.AL.annot.df$Cage

# separate into training and testing
AL.cages.unique <- unique(AL.cage.vec)
n.AL.training.cages <- floor(0.7*length(AL.cages.unique))

set.seed(200)
AL.training.cages <- sample(AL.cages.unique, size=n.AL.training.cages, replace=F)
AL.testing.cages <- setdiff(AL.cages.unique, AL.training.cages)

# training
genus.AL.training.data.df.w.age <- genus.AL.annot.df %>% 
  
  # subset to cages selected for training
  dplyr::filter(Cage %in% AL.training.cages) %>% 
  
  # keep just age and microbiome features
  dplyr::select(age.wks, all_of(tax.feats))

# testing
genus.AL.testing.data.df.w.age <- genus.AL.annot.df %>% 
  
  # subset to cages selected for testing
  dplyr::filter(Cage %in% AL.testing.cages) %>% 
  
  # keep just age and microbiome features
    dplyr::select(age.wks, all_of(tax.feats))

nrow(genus.AL.training.data.df.w.age)
nrow(genus.AL.testing.data.df.w.age)
```

* 374 samples used for training, 199 for testing

Train RF with default parameters.

```{r}
rf.age.AL.genus.training <- randomForest(
  age.wks ~ .,
  data=genus.AL.training.data.df.w.age,
  importance=T)
```

```{r}
pred.v.obs.df.age.AL.genus.testing <- data.frame(
  pred = predict(rf.age.AL.genus.training, newdata=genus.AL.testing.data.df.w.age),
  obs = genus.AL.testing.data.df.w.age$age.wks) %>% 
  mutate(sample.type = "testing")

# these are OOB predictions
pred.v.obs.df.age.AL.genus.training <- data.frame(
  pred = rf.age.AL.genus.training$predicted,
  obs = genus.AL.training.data.df.w.age$age.wks) %>% 
  mutate(sample.type = "training")

pred.v.obs.df.age.AL.genus <- rbind(
  pred.v.obs.df.age.AL.genus.testing,
  pred.v.obs.df.age.AL.genus.training) %>% 
  mutate(pred.minus.obs = pred - obs)
```

```{r}
(age.prediction.AL.p <- pred.v.obs.df.age.AL.genus %>% 
   arrange(desc(sample.type)) %>% {
     ggplot(., aes(x=obs*0.23, y=pred*0.23, color=sample.type)) +
       geom_jitter(width=0.25, shape=16) +
       geom_smooth(method="lm", linewidth=0.5) +
       geom_abline(slope=1, lty=2) +
       labs(x="Actual age (months)", y="Predicted age (months)",
            title="Age prediction") +
       scale_color_manual(values=c(training="gray80", testing=age.color)) +
       theme_classic(base_size=8) +
       theme(legend.position="none")
   })
```

```{r}
# pdf("../plots/age_prediction_100_genera.pdf", width=2, height=2)
# age.prediction.AL.p
# dev.off()
```

## Pathways

```{r}
pathway.AL.log2tpm.annot.df <- pathway.log2tpm.annot.df %>% 
  dplyr::filter(Diet == "AL")
```

We can use the stratified cages from the taxonomic prediction.

```{r}
# training
pathway.AL.training.data.df.w.age <- pathway.AL.log2tpm.annot.df %>% 
  
  # subset to cages selected for training
  dplyr::filter(Cage %in% AL.training.cages) %>% 
  
  # keep just age and microbiome features
  dplyr::select(age.wks, all_of(fxn.feats))

# testing
pathway.AL.testing.data.df.w.age <- pathway.AL.log2tpm.annot.df %>% 
  
  # subset to cages selected for testing
  dplyr::filter(Cage %in% AL.testing.cages) %>% 
  
  # keep just age and microbiome features
  dplyr::select(age.wks, all_of(fxn.feats))

dim(pathway.AL.training.data.df.w.age)
dim(pathway.AL.testing.data.df.w.age)
```

Train RF with default parameters.

```{r}
rf.age.AL.pathway.training <- randomForest(
  age.wks ~ .,
  data=pathway.AL.training.data.df.w.age,
  importance=T)
```

```{r}
pred.v.obs.df.age.AL.pathway.testing <- data.frame(
  pred = predict(rf.age.AL.pathway.training, newdata=pathway.AL.testing.data.df.w.age),
  obs = pathway.AL.testing.data.df.w.age$age.wks) %>% 
  mutate(sample.type = "testing")

pred.v.obs.df.age.AL.pathway.training <- data.frame(
  pred = rf.age.AL.pathway.training$predicted,
  obs = pathway.AL.training.data.df.w.age$age.wks) %>% 
  mutate(sample.type = "training")

pred.v.obs.df.age.AL.pathway <- rbind(
  pred.v.obs.df.age.AL.pathway.testing,
  pred.v.obs.df.age.AL.pathway.training) %>% 
  mutate(pred.minus.obs = pred - obs)
```

```{r}
(age.prediction.AL.pathway.p <- pred.v.obs.df.age.AL.pathway %>% 
   arrange(desc(sample.type)) %>%
   ggplot(aes(x=obs*0.23, y=pred*0.23, color=sample.type)) +
   geom_jitter(width=0.25, shape=16) +
   geom_smooth(method="lm", linewidth=0.5) +
   geom_abline(slope=1, lty=2) +
   scale_color_manual(values=c(training="gray80", testing=age.color)) +
   labs(x="Actual age (months)", y="Predicted age (months)",
        title="Functional prediction") +
   theme_classic(base_size=8) +
   theme(legend.position="none"))
```

```{r}
# pdf("../plots/age_prediction_272_pathways_log2tpm.pdf", width=2, height=2)
# age.prediction.AL.pathway.p
# dev.off()
```

# Train on AL, predict age for DR

## Genera

We can train on all AL samples now, rather than on the 70% training set as above.

```{r}
genus.AL.data.df.w.age <- genus.AL.annot.df %>% 
  
  # keep just age and microbiome features
  dplyr::select(age.wks, all_of(tax.feats))
```

```{r}
rf.age.AL.genus <- randomForest(
  age.wks ~ .,
  data=genus.AL.data.df.w.age,
  importance=T)
```

```{r}
# subset to DR samples
genus.DR.annot.df <- genus.annot.df %>% 
  dplyr::filter(Diet != "AL")

# keep just microbiome features
genus.DR.data.df <- genus.DR.annot.df %>% dplyr::select(all_of(tax.feats))
dim(genus.DR.data.df)
```

Predictions on DR.

```{r}
age.preds.on.DR.trained.on.AL <- predict(
  object = rf.age.AL.genus, 
  newdata = genus.DR.data.df)

DR.pred.df <- data.frame(
  pred = age.preds.on.DR.trained.on.AL,
  obs = genus.DR.annot.df$age.wks,
  Diet = genus.DR.annot.df$Diet
)
```

Predictions on AL -- use OOB preds.

```{r}
AL.pred.df <- data.frame(
  pred = rf.age.AL.genus$predicted, 
  obs = genus.AL.annot.df$age.wks,
  Diet = "AL")
```

Compare DR predictions to AL predictions.

```{r}
pred.v.obs.df.DR.and.AL <- rbind(AL.pred.df, DR.pred.df) %>% 
  mutate(is.AL = (Diet == "AL")) %>% 
  mutate(age.approx.wks = case_when(
    obs %in% c(20, 21, 22, 23) ~ 21,
    obs %in% c(35, 38) ~ 35,
    obs %in% c(43, 44) ~ 44,
    obs %in% c(68, 69) ~ 69,
    obs %in% c(95, 96, 97) ~ 96,
    obs %in% c(120, 121, 122) ~ 121,
    obs %in% c(148, 149) ~ 148,
    obs %in% c(172, 173, 174, 175) ~ 174,
    obs %in% c(200, 201, 203) ~ 200)) %>% 
  mutate(age.approx.months = round(age.approx.wks/4.34)) %>% 
  mutate(pred.minus.obs = pred - obs)
```

Compute t-test between AL and DR predictions at each age.

```{r}
DR.v.AL.pred.t.test.per.age.pval.df <- pred.v.obs.df.DR.and.AL %>% 
  dplyr::filter(age.approx.months %in% c(5, 10, 16, 22, 28)) %>%
  group_by(age.approx.months) %>%
  do(tidy(t.test(pred ~ is.AL, data=.))) %>%
  ungroup() %>% 
  dplyr::select(age.approx.months, p.value) %>% 
  mutate(t.test.pval.symbol = case_when(
    p.value < 0.0001 ~ "****",
    p.value < 0.001 ~ "***",
    p.value < 0.01 ~ "**",
    p.value < 0.05 ~ "*",
    TRUE ~ "ns"))
```

```{r}
(AL.v.DR.age.preds.p <- pred.v.obs.df.DR.and.AL %>% 
   arrange(is.AL) %>% 
   dplyr::filter(age.approx.months %in% c(5, 10, 16, 22, 28)) %>% 
   ggplot(aes(x=obs/4.34, y=pred/4.34, color=is.AL)) +
   geom_abline(slope=1, lty=2) +
   geom_vline(xintercept=6, lty=2) +
   geom_text(aes(x=age.approx.months, y=30, label=t.test.pval.symbol, color=NULL), 
             size=0.36*6, data=DR.v.AL.pred.t.test.per.age.pval.df) +
   geom_jitter(width=0.5, shape=16, size=0.5) +
   stat_summary(aes(x=age.approx.months, group=is.AL), fun="mean", geom="crossbar", width=3) +
   labs(x="Actual age (months)", y="Predicted age (months)", 
        color="Dietary group", title="Taxonomic features") +
   scale_color_manual(values = c(`TRUE`="seashell4", `FALSE`=diet.color)) +
   theme_classic(base_size=8) +
   theme(legend.position="none"))
```

* We consistently predict *higher* age for DR samples
* Why do we predict higher age at 5 months? Because OOB accuracy will be a little better than testing on never before seen samples. Therefore, the AL samples will be closer to y=x.

```{r}
# pdf("../plots/age_prediction_DR_trained_on_AL.pdf", width=2, height=2)
# AL.v.DR.age.preds.p
# dev.off()
```

## Pathways

Train on AL samples now.

```{r}
pathway.AL.data.df.w.age <- pathway.AL.log2tpm.annot.df %>% 
  dplyr::select(age.wks, all_of(fxn.feats)) # keep just age and microbiome features
```

```{r}
rf.age.AL.pathway <- randomForest(
  age.wks ~ .,
  data=pathway.AL.data.df.w.age,
  importance=T)
```

```{r}
# subset to DR samples
pathway.DR.annot.df <- pathway.log2tpm.annot.df %>% 
  dplyr::filter(Diet != "AL")

# keep just microbiome features
pathway.DR.data.df <- pathway.DR.annot.df %>% 
  dplyr::select(all_of(fxn.feats))
dim(pathway.DR.data.df)
```

Predictions on DR.

```{r}
age.preds.on.DR.trained.on.AL.pathways <- predict(
  object = rf.age.AL.pathway, 
  newdata = pathway.DR.data.df)

pathway.DR.pred.df <- data.frame(
  pred = age.preds.on.DR.trained.on.AL.pathways,
  obs = pathway.DR.annot.df$age.wks,
  Diet = pathway.DR.annot.df$Diet
)
```

Predictions on AL -- use OOB preds.

```{r}
pathway.AL.pred.df <- data.frame(
  pred = rf.age.AL.pathway$predicted, 
  obs = pathway.AL.log2tpm.annot.df$age.wks,
  Diet = "AL")
```

Compare DR predictions to AL predictions.

```{r}
pred.v.obs.df.DR.and.AL.pathway <- rbind(pathway.AL.pred.df, pathway.DR.pred.df) %>% 
  mutate(is.AL = (Diet == "AL")) %>% 
  mutate(age.approx.wks = case_when(
    obs %in% c(20, 21, 22, 23) ~ 21,
    obs %in% c(35, 38) ~ 35,
    obs %in% c(43, 44) ~ 44,
    obs %in% c(68, 69) ~ 69,
    obs %in% c(95, 96, 97) ~ 96,
    obs %in% c(120, 121, 122) ~ 121,
    obs %in% c(148, 149) ~ 148,
    obs %in% c(172, 173, 174, 175) ~ 174,
    obs %in% c(200, 201, 203) ~ 200)) %>% 
  mutate(age.approx.months = round(age.approx.wks/4.34)) %>% 
  mutate(pred.minus.obs = pred - obs)
```

Compute t-test between AL and DR predictions at each age.

```{r}
pathway.DR.v.AL.pred.t.test.per.age.pval.df <- pred.v.obs.df.DR.and.AL.pathway %>% 
  dplyr::filter(age.approx.months %in% c(5, 10, 16, 22, 28)) %>%
  group_by(age.approx.months) %>%
  do(tidy(t.test(pred ~ is.AL, data=.))) %>%
  ungroup() %>% 
  dplyr::select(age.approx.months, p.value) %>% 
  mutate(t.test.pval.symbol = case_when(
    p.value < 0.0001 ~ "****",
    p.value < 0.001 ~ "***",
    p.value < 0.01 ~ "**",
    p.value < 0.05 ~ "*",
    TRUE ~ "ns"))
```

```{r}
(AL.v.DR.age.preds.pathway.p <- pred.v.obs.df.DR.and.AL.pathway %>% 
   arrange(is.AL) %>% 
   dplyr::filter(age.approx.months %in% c(5, 10, 16, 22, 28)) %>% 
   ggplot(aes(x=obs/4.34, y=pred/4.34, color=is.AL)) +
   geom_abline(slope=1, lty=2) +
   geom_vline(xintercept=6, lty=2) +
   geom_text(aes(x=age.approx.months, y=28, label=t.test.pval.symbol, color=NULL), 
             size=0.36*6, data=pathway.DR.v.AL.pred.t.test.per.age.pval.df) +
   geom_jitter(width=0.5, shape=16, size=0.5) +
   stat_summary(aes(x=age.approx.months, group=is.AL), fun="mean", geom="crossbar", width=3) +
   # geom_smooth(se=F) +
   labs(x="Actual age (months)", y="Predicted age (months)", 
        color="Dietary group", title="Functional features") +
   scale_color_manual(values = c(`TRUE`="seashell4", `FALSE`=diet.color)) +
   theme_classic(base_size=8) +
   theme(legend.position="none"))
```

```{r}
# pdf("../plots/age_prediction_DR_trained_on_AL_pathway.pdf", width=2, height=2)
# AL.v.DR.age.preds.pathway.p
# dev.off()
```

# Train on 40% CR, predict age for rest

## Genera

```{r}
genus.40.annot.df <- genus.annot.df %>% 
  dplyr::filter(Diet == "40")

genus.not40.annot.df <- genus.annot.df %>% 
  dplyr::filter(Diet != "40")
```

```{r}
genus.40.data.df.w.age <- genus.40.annot.df %>% 
  dplyr::select(age.wks, all_of(tax.feats)) # keep just age and microbiome features

genus.not40.data.df.w.age <- genus.not40.annot.df %>% 
  dplyr::select(age.wks, all_of(tax.feats)) # keep just age and microbiome features

dim(genus.40.data.df.w.age)
dim(genus.not40.data.df.w.age)
```

```{r}
rf.age.40.genus <- randomForest(
  age.wks ~ .,
  data=genus.40.data.df.w.age,
  importance=T)
```

```{r}
age.preds.trained.on.40 <- predict(
  object = rf.age.40.genus, 
  newdata = genus.not40.data.df.w.age)

genus.not40.annot.df.w.preds.trained.on.40 <- genus.not40.annot.df %>% 
  mutate(pred = age.preds.trained.on.40) %>% 
  mutate(actual.minus.pred = age.wks - pred)
```

```{r}
(train.40.test.rest.genus.p <- genus.not40.annot.df.w.preds.trained.on.40 %>% 
   dplyr::filter(age.approx.months %in% c(5, 10, 16, 22, 28)) %>%
   mutate(Diet = factor(Diet, levels = c("AL", "1D", "2D", "20"))) %>% 
   ggplot(aes(x=Diet, y=pred*0.23, fill=Diet)) +
   geom_boxplot(outlier.shape=16, outlier.size=0.5, linewidth=0.5) +
   geom_hline(aes(yintercept = age.approx.months), lty=2, linewidth=0.5) +
   facet_wrap(~age.approx.months, nrow=1) +
   stat_compare_means(method="t.test", label="p.signif", ref.group="AL") +
   labs(x="", y="Predicted age (months)", title="Trained on 40%, dotted line shows actual age") +
   scale_fill_manual(values=diet.palette) +
   theme_classic(base_size=8) +
   theme(legend.position="none"))
```

```{r}
# pdf("../plots/age_prediction_genera_trained_on_40_boxplots.pdf", height=1.5, width=4)
# train.40.test.rest.genus.p
# dev.off()
```

## Pathways

```{r}
pathway.40.log2tpm.annot.df <- pathway.log2tpm.annot.df %>% 
  dplyr::filter(Diet == "40")

pathway.not40.log2tpm.annot.df <- pathway.log2tpm.annot.df %>% 
  dplyr::filter(Diet != "40")
```

```{r}
pathway.40.data.df.w.age <- pathway.40.log2tpm.annot.df %>% 
  dplyr::select(age.wks, all_of(fxn.feats)) # keep just age and microbiome features

pathway.not40.data.df.w.age <- pathway.not40.log2tpm.annot.df %>% 
  dplyr::select(age.wks, all_of(fxn.feats)) # keep just age and microbiome features

dim(pathway.40.data.df.w.age)
dim(pathway.not40.data.df.w.age)
```

```{r}
rf.age.40.pathway <- randomForest(
  age.wks ~ .,
  data=pathway.40.data.df.w.age,
  importance=T)
```

```{r}
age.preds.trained.on.40.pathway <- predict(
  object = rf.age.40.pathway, 
  newdata = pathway.not40.data.df.w.age)

pathway.annot.df.w.preds.trained.on.40 <- pathway.not40.log2tpm.annot.df %>% 
  mutate(pred = age.preds.trained.on.40.pathway) %>% 
  mutate(actual.minus.pred = age.wks - pred)
```

```{r}
(train.40.test.rest.pathway.p <- pathway.annot.df.w.preds.trained.on.40 %>% 
   dplyr::filter(age.approx.months %in% c(5, 10, 16, 22, 28)) %>%
   mutate(Diet = factor(Diet, levels = c("AL", "1D", "2D", "20"))) %>% 
   ggplot(aes(x=Diet, y=pred*0.23, fill=Diet)) +
   geom_boxplot(outlier.shape=16, outlier.size=0.5, linewidth=0.5) +
   geom_hline(aes(yintercept = age.approx.months), lty=2, linewidth=0.5) +
   facet_wrap(~age.approx.months, nrow=1) +
   stat_compare_means(method="t.test", label="p.signif", ref.group="AL") +
   labs(x="", y="Predicted age (months)", title="Trained on 40%, dotted line shows actual age") +
   scale_fill_manual(values=diet.palette) +
   theme_classic(base_size=8) +
   theme(legend.position="none"))
```

```{r}
# pdf("../plots/age_prediction_pathways_trained_on_40_boxplots.pdf", height=1.5, width=4)
# train.40.test.rest.pathway.p
# dev.off()
```

# Predict diet

## Genera

Use cross-validation to get a prediction per cross-validation fold. Stratify by cage.

```{r}
run_rf_diet_one_age_strat_by_cage <- function(this.age.approx.month) {
  
  cat(this.age.approx.month, "\n")
  
  this.age.annot.df <- genus.annot.df %>% 
    dplyr::filter(age.approx.months == this.age.approx.month)
  
  colnames(this.age.annot.df)
  
  # extract vector of cages for stratification & remove cage
  this.cage.vec <- this.age.annot.df$Cage
  this.age.data.df <- this.age.annot.df %>%
    dplyr::select(Diet, all_of(tax.feats))
  
  # use groupKFold to make sure that cages aren't split between training and testing
  these.folds <- groupKFold(this.cage.vec, k=10)
  this.train.control <- trainControl(
    index=these.folds, method="cv",
    savePredictions="final")
  
  # perform cross-validation to get predictions on validation sets
  this.diet.rf.strat.by.cage <- train(
    Diet ~ .,
    data = this.age.data.df, 
    method = 'rf', 
    
    # no need to do parameter search, use rf default for classification
    tuneGrid = data.frame(.mtry = floor(sqrt(length(tax.feats)))),
    trControl = this.train.control)
  
  # return predictions on validation sets
  return(this.diet.rf.strat.by.cage$pred %>% 
           mutate(age.months = this.age.approx.month))
}
```

```{r}
all.rf.diet.pred.res <- sapply(
  c(5, 10, 16, 22, 28),
  run_rf_diet_one_age_strat_by_cage
)
```

```{r}
all.rf.diet.pred.df <- all.rf.diet.pred.res %>% 
  t() %>% data.frame() %>% 
  unnest(everything()) %>% 
  mutate(is.correct = (pred == obs))
```

Calculate accuracy on each fold.

```{r}
all.rf.diet.accuracy.df <- all.rf.diet.pred.df %>% 
  group_by(age.months, Resample) %>% 
  summarise(accuracy = sum(is.correct) / n(), .groups="drop")
```

Add t-test p-value (null hypothesis is that accuracy is *greater* than 20%).

```{r}
all.rf.diet.t.test.df <- all.rf.diet.accuracy.df %>% 
  dplyr::filter(age.months %in% c(5, 10, 16, 22, 28)) %>% 
  mutate(accuracy.minus.20 = accuracy - 0.2) %>% 
  group_by(age.months) %>% 
  summarise(t.test.pval = t.test(accuracy.minus.20, alternative="greater")$p.value, .groups="drop") %>% 
  mutate(t.test.pval.symbol = case_when(
    t.test.pval < 0.0001 ~ "****",
    t.test.pval < 0.001 ~ "***",
    t.test.pval < 0.01 ~ "**",
    t.test.pval < 0.05 ~ "*",
    TRUE ~ "ns"))
```

```{r}
(diet.pred.v.age.p <- all.rf.diet.accuracy.df %>% 
   dplyr::filter(age.months %in% c(5, 10, 16, 22, 28)) %>% 
   mutate(is.post.randomiz = (age.months > 5)) %>% 
   ggplot(aes(x=factor(age.months), y=100*accuracy, fill=is.post.randomiz)) +
   stat_summary(fun="mean", geom="col") +
   geom_point(size=0.5, shape=16) +
   geom_text(aes(y=85, label=t.test.pval.symbol, fill=NULL),
             data=all.rf.diet.t.test.df) +
   geom_hline(yintercept=20, lty=2) +
   labs(x="Age (months)", y="Accuracy", 
        title="Prediction of dietary group, genera") +
   scale_fill_manual(values=c(`FALSE`="gray80", `TRUE`=diet.color)) +
   ylim(0, 90) +
   theme_classic(base_size=8) +
   theme(legend.position="none"))
```

```{r}
# pdf("../plots/diet_prediction_accuracy_v_age_genera.pdf", width=2, height=2)
# diet.pred.v.age.p
# dev.off()
```

### Accuracy by diet

```{r}
no.5mo.rf.diet.pred.df <- all.rf.diet.pred.df %>% 
  dplyr::filter(age.months != 5)
```

```{r}
(diet.pred.v.diet.p <- no.5mo.rf.diet.pred.df %>% 
   group_by(obs) %>% 
   summarise(accuracy = sum(is.correct) / n(), .groups="drop") %>% 
   mutate(diet = factor(obs, levels = c("AL", "1D", "2D", "20", "40"))) %>% 
   ggplot(aes(x=diet, y=100*accuracy, fill=diet)) +
   geom_col() +
   scale_fill_manual(values=diet.palette) +
   ylim(0, 80) +
   theme_classic(base_size=8) +
   theme(legend.position="none") +
   labs(x="Dietary group", y="Accuracy", 
        title="Post-randomization, genera"))
```

```{r}
# pdf("../plots/diet_prediction_accuracy_v_diet_post_randomiz.pdf", width=2, height=2)
# diet.pred.v.diet.p
# dev.off()
```

## Pathways

```{r}
run_rf_diet_one_age_pathway_strat_by_cage <- function(this.age.approx.month) {
  
  cat(this.age.approx.month, "\n")
  
  # subset to this age
  this.age.pathway.annot.df <- pathway.log2tpm.annot.df %>% 
    dplyr::filter(age.approx.months == this.age.approx.month)
  
  # extract vector of cages for stratification
  this.cage.vec <- this.age.pathway.annot.df$Cage
  
  # keep pathways and Diet
  this.age.pathway.data.df <- this.age.pathway.annot.df %>% 
    dplyr::select(Diet, all_of(fxn.feats))
  
  # use groupKFold to make sure that cages aren't split between training and testing
  these.pathway.folds <- groupKFold(this.cage.vec, k=10)
  this.pathway.train.control <- trainControl(
    index=these.pathway.folds, method="cv",
    savePredictions="final")
  
  # perform cross-validation to get predictions on validation sets
  this.diet.pathway.rf.strat.by.cage <- train(
    Diet ~ .,
    data = this.age.pathway.data.df, 
    method = 'rf', 
    
    # no need to do parameter search, use rf default for classification
    tuneGrid = data.frame(.mtry = floor(sqrt(length(fxn.feats)))),
    trControl = this.pathway.train.control)
  
  # return predictions on validation sets
  return(this.diet.pathway.rf.strat.by.cage$pred %>% 
           mutate(age.months = this.age.approx.month))
}
```

```{r}
all.rf.diet.pathway.pred.res <- sapply(
  c(5, 10, 16, 22, 28),
  run_rf_diet_one_age_pathway_strat_by_cage
)
```

```{r}
all.rf.diet.pathway.pred.df <- all.rf.diet.pathway.pred.res %>% 
  t() %>% data.frame() %>% 
  unnest(everything()) %>% 
  mutate(is.correct = (pred == obs))
```

Calculate accuracy on each fold.

```{r}
all.rf.diet.pathway.accuracy.df <- all.rf.diet.pathway.pred.df %>% 
  group_by(age.months, Resample) %>% 
  summarise(accuracy = sum(is.correct) / n(), .groups="drop")
```

Add t-test p-value.

```{r}
all.rf.diet.pathway.t.test.df <- all.rf.diet.pathway.accuracy.df %>% 
  dplyr::filter(age.months %in% c(5, 10, 16, 22, 28)) %>% 
  mutate(accuracy.minus.20 = accuracy - 0.2) %>% 
  group_by(age.months) %>% 
  summarise(t.test.pval = t.test(accuracy.minus.20, alternative="greater")$p.value, .groups="drop") %>% 
  mutate(t.test.pval.symbol = case_when(
    t.test.pval < 0.0001 ~ "****",
    t.test.pval < 0.001 ~ "***",
    t.test.pval < 0.01 ~ "**",
    t.test.pval < 0.05 ~ "*",
    TRUE ~ "ns"))
```

```{r}
(diet.pred.v.age.pathway.p <- all.rf.diet.pathway.accuracy.df %>% 
   dplyr::filter(age.months %in% c(5, 10, 16, 22, 28)) %>% 
   mutate(is.post.randomiz = (age.months > 5)) %>% 
   ggplot(aes(x=factor(age.months), y=100*accuracy, fill=is.post.randomiz)) +
   stat_summary(fun="mean", geom="col") +
   geom_point(size=0.5, shape=16) +
   geom_text(aes(y=85, label=t.test.pval.symbol, fill=NULL),
             data=all.rf.diet.pathway.t.test.df) +
   geom_hline(yintercept=20, lty=2) +
   labs(x="Age (months)", y="Accuracy", 
        title="Prediction of dietary group, pathways") +
   scale_fill_manual(values=c(`FALSE`="gray80", `TRUE`=diet.color)) +
   ylim(0, 90) +
   theme_classic(base_size=8) +
   theme(legend.position="none"))
```

```{r}
# pdf("../plots/diet_prediction_accuracy_v_age_pathway.pdf", width=2, height=2)
# diet.pred.v.age.pathway.p
# dev.off()
```

### Accuracy by diet

```{r}
no.5mo.rf.diet.pathway.pred.df <- all.rf.diet.pathway.pred.df %>% 
  dplyr::filter(age.months != 5)
```

```{r}
(diet.pathway.pred.v.diet.p <- no.5mo.rf.diet.pathway.pred.df %>% 
   group_by(obs) %>% 
   summarise(accuracy = sum(is.correct) / n(), .groups="drop") %>% 
   mutate(diet = factor(obs, levels = c("AL", "1D", "2D", "20", "40"))) %>% 
   ggplot(aes(x=diet, y=100*accuracy, fill=diet)) +
   geom_col() +
   scale_fill_manual(values=diet.palette) +
   ylim(0, 80) +
   theme_classic(base_size=8) +
   theme(legend.position="none") +
   labs(x="", y="Accuracy", title="Post-randomization, pathways"))
```

```{r}
# pdf("../plots/diet_prediction_accuracy_v_diet_post_randomiz_pathway.pdf", width=2, height=2)
# diet.pathway.pred.v.diet.p
# dev.off()
```

## Is accuracy better for genera than pathways?

```{r}
t.test(
  all.rf.diet.accuracy.df %>% dplyr::filter(age.months > 5) %>% pull(accuracy),
  all.rf.diet.pathway.accuracy.df %>% dplyr::filter(age.months > 5) %>% pull(accuracy))
```
